Evans Day Abstract Analysis
========================================================
`r Sys.Date()`
---------------------------------------------

Analyses to try:
  1. Former smokers only
  2. COPD vs COPD+Cancer vs Healthy as compared to Cancer instead of COPD


```{r GlobalVariables, include=FALSE}
cacheOption = FALSE

```


```{r defineFunctions, include=FALSE}
foldChange <- function(eset, groups, genes){
  apply(exprs(eset)[genes, groups==0], 1, mean) - 
    apply(exprs(eset)[genes, groups==1], 1, mean)
}

medianFilter <- function(eset){
  md <- median(exprs(eset))
  passFilter <- logical(featureNumber(eset))
  for(i in 1:featureNumber(eset)){
    passFilter[i] <- sum(exprs(eset)[i, ] > md) > 0
  }
  
  eset <- eset[passFilter, ]
  return(eset)
  
}

```


```{r LoadData, echo=FALSE, cache=cacheOption, include=FALSE}
# set the working directory to the directory of this script
setwd("/restricted/projectnb/pulmarray/LinGA_protected/Allegro/COPD_Cancer/experiments/2014-09-03_Evans_Day")

# load the data using a script that removes patients with:
#  Cancer = NA
#  SMK    = 3
source("/protected/projects/pulmarray/Allegro/COPD_Cancer/scripts/AllegroSetup.R")

# clean the data (i.e. remove DKs and NAs)
eset <- removeFactorLevel(eset, "FinalCaDXc", "DK")
eset <- removeFactorLevel(eset, "AllegroCOPDc", "DK")
eset <- cleanNAForAnalysis(eset, "AGEcalc")
eset <- cleanNAForAnalysis(eset, "RIN")
eset <- removeBioReps(eset)
eset <- calcIndicator(eset, "FinalCaDXc", "AllegroCOPDc")

holdEset <- eset

library(znorm)

```

### GOAL 1: Analysis comparing COPD to COPD+Cancer group
```{r Analysis1COPDvsCOPDCancerFormer, cache=cacheOption, fig.width=8,fig.height=7, fig.align='center'}

# Note: use the median filter only after final eset for analysis is formed

# CLEAN the data further
# remove the disease-free patients
eset <- holdEset
eset <- removeFactorLevel(eset, "indicator", "1")
# remove the Cancer-only patients
eset <- removeFactorLevel(eset, "indicator", "2")
esetF <- medianFilter(eset)

# Analysis 1: COPD vs COPD+Cancer in all smokers; correcting for smoking, age, and RIN
analysis1 <- lmFitWrapper(esetF, c("FinalCaDXc", "SMKc", "AGEcalc", "RIN"), 1, p.value=0.10, 
                          name="Analysis 1: COPD vs COPD+Cancer in all smokers")
hMap1 <- generate_heatmap2(analysis1, tp="FinalCaDXc")

# Analysis 1a
analysis1a <- lmFitWrapper(analysis1$eset, c("FinalCaDXc", "SMKc", "AGEcalc", "RIN"), 1, p.value=0.15,
                           name="Analysis 1a: COPD vs COPD+Cancer in all smokers")
hMap1a <- generate_heatmap2(analysis1a, tp="FinalCaDXc")

# Analysis 2: COPD vc COPD+Cancer in formers only; correcting for Age and RIN
eset <- removeFactorLevel(eset, "SMKc", "1")
esetF <- medianFilter(eset)
analysis2 <- lmFitWrapper(esetF, c("FinalCaDXc", "AGEcalc", "RIN"), 1, p.value=0.15,
                          name="Analysis 2: COPD vc COPD+Cancer in formers only")
hMap2 <- generate_heatmap2(analysis2, tp="FinalCaDXc")
```

The problem above seems to be related a block of patients in the COPD-only,
former smoker group that differ significantly from the rest of the former smokers. Could they possibly be current smokers? Try the above analysis with just current smokers

```{r Analysis3COPDvsCOPDCancerCurrent, cache=cacheOption, fig.width=8,fig.height=7, fig.align='center'}

# Analysis 3: COPD vc COPD+Cancer in currents only
# Note: use the median filter only after final eset for analysis is formed
# Try analysis 2 in current smokers only (see note above)
eset <- holdEset
eset <- removeFactorLevel(eset, "indicator", "1")
eset <- removeFactorLevel(eset, "indicator", "2")
eset <- removeFactorLevel(eset, "SMKc", "2")
esetF <- medianFilter(eset)
analysis3 <- lmFitWrapper(esetF, c("FinalCaDXc", "AGEcalc", "RIN"), 1, p.value=0.25,
                          name="Analysis 3: COPD vc COPD+Cancer in currents only")
# Analysis 3a
print("Try not correcting for multiple comparisons...")
analysis3a <- lmFitWrapper(esetF, c("FinalCaDXc", "AGEcalc", "RIN"), 1, adjust.method="none", p.value=0.01,
                           name="Analysis 3a: COPD vc COPD+Cancer in currents only (no adjustment)")
hMap3 <- generate_heatmap2(analysis3a, tp="FinalCaDXc")
```


### GOAL 2: Analysis of COPD+Cancer vs. everyone in formers only
```{r Analysis4AllOrNone, fig.width=8, fig.height=7, fig.align='center', cache=cacheOption}
# Analysis 4: All-or-none in all smokers
# Generate an all-or-none binary variable and filter down to desired patient subset
eset <- holdEset
eset$AllOrNone <- eset$indicator
eset$AllOrNone[eset$indicator != 4] <- 1
eset <- removeFactorLevel(eset, "AllOrNone", "2")
eset <- removeFactorLevel(eset, "AllOrNone", "3")
esetF <- medianFilter(eset)
analysis4 <- lmFitWrapper(esetF, c("AllOrNone", "SMKc", "AGEcalc", "RIN"), 1, p.value=0.05,
                          name="Analysis4: AllOrNone in all smokers")
hMap4 <- generate_heatmap2(analysis4, tp="indicator")

# Analysis 5: All-or-none in former smokers only
eset <- removeFactorLevel(eset, "SMKc", "1")
esetF <- medianFilter(eset)
analysis5 <- lmFitWrapper(esetF, c("AllOrNone", "AGEcalc", "RIN"), 1, p.value=0.01,
                          name="Analysis5: All-or-none in formers only")
hMap5 <- generate_heatmap2(analysis5, tp="indicator")
```

There appears to be a lot of heterogeneity related to the signal above even though there are many significant genes. Try this in the PFT-defined COPD cohort only

```{r COPDCancerVAllPFT, fig.width=8, fig.height=7, fig.align='center', cache=cacheOption}
# Analysis 6: All-or-none, PFT-defined COPD
# Filter down to patients with PFT-defined COPD; redefine 'indicator' variable
eset <- holdEset
eset <- cleanNAForAnalysis(eset, "COPD2_R7")
eset <- removeFactorLevel(eset, "COPD2_R7", "DK")
eset <- calcIndicator(eset, "FinalCaDXc", "COPD2_R7")
eset$AllOrNone <- eset$indicator
eset$AllOrNone[eset$indicator != 4] <- 1
eset <- removeFactorLevel(eset, "AllOrNone", "2")
eset <- removeFactorLevel(eset, "AllOrNone", "3")
esetF <- medianFilter(eset)
analysis6 <- lmFitWrapper(esetF, c("AllOrNone", "SMKc", "AGEcalc", "RIN"), 1, p.value=0.25,
                          name="Analysis6: All-or-none, PFT-defined COPD R<0.7")
hMap6 <- generate_heatmap2(analysis6, tp="indicator")

# Analysis 7: All-or-none, PFT-defined COPD in formers only
# still very heterogenous signal based on smoking
# remove the current smokers and run the above (model6) only in formers
eset <- removeFactorLevel(eset, "SMKc", "1")
esetF <- medianFilter(eset)
analysis7 <- lmFitWrapper(esetF, c("AllOrNone", "AGEcalc", "RIN"), 1, p.value=0.25,
                          name="Analysis7: All-or-none, PFT-defined COPD, formers only")
hMap7 <- generate_heatmap2(analysis7, tp="indicator")

# Analysis 8: Model 7 + Pack Years
# There appears to be a gradient among the disease-free and cancer-only groups; try including PY 
eset <- cleanNAForAnalysis(eset, "PYc")
esetF <- medianFilter(eset)
analysis8 <- lmFitWrapper(esetF, c("AllOrNone", "AGEcalc", "RIN", "PYc"), 1, p.value=0.25,
                          name="Analysis8: All-or-none, PFT-defined COPD, formers only, PYs")
hMap8 <- generate_heatmap2(analysis8, tp="indicator")

# Analysis 9: Model 8 + Gender
# the above appears more homogenous within each group but gender now appears to be playing a role
eset <- removeFactorLevel(eset, "GENDERc", "DK")
esetF <- medianFilter(eset)
analysis9 <- lmFitWrapper(esetF, c("AllOrNone", "AGEcalc", "RIN", "PYc", "GENDERc"), 1, p.value=0.15,
                          name="Analysis9: All-or-none, PFT-defined COPD, formers only, PYs + Gender")
hMap9 <- generate_heatma2p(analysis9, tp="indicator")

# The heatmap from analysis #9 appears to be the most clean so far defining genes that
# are more homogenous within the COPD+Cancer and COPD groups (former smokers only)
# Define the up and down gene clusters
clusters9 <- return_cluster(analysis9$inds, analysis9$eset, n.clusters=2)

# Save the up and down gene clusters out separately
save_entrez(analysis9$inds[clusters9==1], rownames(analysis9$eset), filename="AllOrNoneUP.AGERINPYGENDER.txt")
save_entrez(analysis9$inds[clusters9==2], rownames(analysis9$eset), filename="AllOrNoneDOWN.AGERINPYGENDER.txt")

# Create heatmaps of each gene cluster separately
generate_heatmap(analysis9$inds[clusters9==1], analysis9$eset, tp="indicator", mn="Model9Up")
generate_heatmap(analysis9$inds[clusters9==2], analysis9$eset, tp="indicator", mn="Model9Down")

# Use t-tests to compare the 50 up-genes in all 4 groups
# Compare up-genes in COPD vs COPD+Cancer
testUPCOPDCancervCOPD <- t.test(exprs(analysis9$eset)[analysis9$inds[clusters9==1], analysis9$eset$indicator==3], exprs(analysis9$eset)[analysis9$inds[clusters9==1], analysis9$eset$indicator==4])
testUPCOPDCancervCOPD

# Compare up-genes in Cancer vs COPD+Cancer
testUPCOPDCancervCancer <- t.test(exprs(analysis9$eset)[analysis9$inds[clusters9==1], analysis9$eset$indicator==2], exprs(analysis9$eset)[analysis9$inds[clusters9==1], analysis9$eset$indicator==4])
testUPCOPDCancervCancer

# Compare up-genes in DiseaseFree vs COPD+Cancer
testUPCOPDCancervDiseaseFree <- t.test(exprs(analysis9$eset)[analysis9$inds[clusters9==1], analysis9$eset$indicator==1], exprs(analysis9$eset)[analysis9$inds[clusters9==1], analysis9$eset$indicator==4])
testUPCOPDCancervDiseaseFree

# Use t-tests to compare the 38 down-genes in all 4 groups
# Compare down-genes in COPD vs COPD+Cancer
testDOWNCOPDCancervCOPD <- t.test(exprs(analysis9$eset)[analysis9$inds[clusters9==2], analysis9$eset$indicator==3], exprs(analysis9$eset)[analysis9$inds[clusters9==2], analysis9$eset$indicator==4])
testDOWNCOPDCancervCOPD

# Compare down-genes in Cancer vs COPD+Cancer
testDOWNCOPDCancervCancer <- t.test(exprs(analysis9$eset)[analysis9$inds[clusters9==2], analysis9$eset$indicator==2], exprs(analysis9$eset)[analysis9$inds[clusters9==2], analysis9$eset$indicator==4])
testDOWNCOPDCancervCancer

# Compare down-genes in DiseaseFree vs COPD+Cancer
testDOWNCOPDCancervDiseaseFree <- t.test(exprs(analysis9$eset)[analysis9$inds[clusters9==2], analysis9$eset$indicator==1], exprs(analysis9$eset)[analysis9$inds[clusters9==2], analysis9$eset$indicator==4])
testDOWNCOPDCancervDiseaseFree
```


## Call smoker status by molecular profile of Genome Biology Genes
```{r MolecularSmokingStatus, fig.width=8, fig.height=7, fig.align='center', cache=cacheOption}
# Analysis 10: code smoking status based on genome biology up-in-current-smokers gene panel
eset <- holdEset

# load smoking related genes; of interest are genes up in current smokers
smokerGenes <- read.table("../../geneSubsets/pnas_genome_biology_smokers_genesets.gmx", fill=TRUE, head=TRUE)
down_in_smokers_PNAS <- smokerGenes$Down_In_Smokers_PNAS
up_in_smokers_PNAS <- smokerGenes$Up_In_Smokers_PNAS
down_in_current_smokers_genome_biology <- smokerGenes$Down_In_Current_Smokers_Genome_Biology
up_in_current_smokers_genome_biology <- smokerGenes$Up_In_Current_Smokers_Genome_Biology
down_in_smokers_PNAS <- down_in_smokers_PNAS[down_in_smokers_PNAS!="" & down_in_smokers_PNAS!="na"]
up_in_smokers_PNAS <- up_in_smokers_PNAS[up_in_smokers_PNAS!="" & up_in_smokers_PNAS!="na"]
down_in_current_smokers_genome_biology <- down_in_current_smokers_genome_biology[down_in_current_smokers_genome_biology!="" & down_in_current_smokers_genome_biology!="na"]
up_in_current_smokers_genome_biology <- up_in_current_smokers_genome_biology[up_in_current_smokers_genome_biology!="" & up_in_current_smokers_genome_biology!="na"]

analysis10 <- lmFitWrapper(eset, c("SMKc"), varOfInterest=1, p.value=0.05,
                           name="Analysis10: Define Molecular Smoking Status based on Genome Biology")

# Find the genes-up-current-smokers in our list of genes
upCurrentGBinds <- match(up_in_current_smokers_genome_biology, analysis10$fit$genes$Symbol)
upCurrentGBinds <- upCurrentGBinds[!is.na(upCurrentGBinds)]

# Make a heatmap of the 50 up-in-current-smokers genes
hMap10 <- generate_heatmap(upCurrentGBinds, eset, tp="indicator",
                           mn="Analysis10: 50 up genes in current smokers Genome Biology")

# Use PCA to reduce the dimensionality of the 50 up genes
pcaSMK <- prcomp(znorm(x=exprs(eset)[upCurrentGBinds, ], margin=1), center=FALSE, scale.=FALSE)
# Plot the PC1 and PC2 to see how the clinically defined smoking status matches with the PCs of the 50 GB genes
plot(pcaSMK$rotation[, 1], pcaSMK$rotation[, 2], col=eset$SMKc, 
     main="PC1, PC2 of 23 genes up in current smokers (Gen. Biol.)",
     xlab="PC1", ylab="PC2")
legend("topright", c("Former", "Current"), fill=eset$SMKc,title="Clinical SMK Status")

# Now use mclust to define the binary 1/2 smoking status based on the Genome Biology 50 genes
library(mclust)

mClustMod1 <- Mclust(pcaSMK$rotation[, 1])
plot(mClustMod1, "BIC", cex.axis=1)
plot(mClustMod1, "classification", cex.axis=1)
plot(mClustMod1, "uncertainty", cex.axis=1)
plot(mClustMod1, "density", cex.axis=1)

# Determine how well the molecular classifications using PCA and mclust match with the clinical smoking status
cor(mClustMod1$classification, as.numeric(eset$SMKc))

qs4 <- quantile(mClustMod1$uncertainty)
qs10 <- quantile(mClustMod1$uncertainty, probs=seq(0,1,0.1))

recodeSMK <- factor(eset$SMKc, levels=c(levels(eset$SMKc), "uncertain"))
recodeSMK[mClustMod1$uncertainty>qs10[10]] <- "uncertain"

recodeSMK_colors <- c("1"="red", "2"="blue", "uncertain"="darkgrey");
plot(pcaSMK$rotation[, 1], pcaSMK$rotation[, 2], col=recodeSMK_colors[recodeSMK], 
     main="PC1, PC2 of 23 genes up in current smokers (Gen. Biol.)",
     xlab="PC1", ylab="PC2")
legend("topright", c("Former", "Current", "Uncertain"), fill=recodeSMK_colors)

recodeSMK2 <- mClustMod1$classification
recodeSMK2[mClustMod1$uncertainty>qs10[10]] <- "uncertain"
recodeSMK2 <- factor(recodeSMK2)
eset$SMK2 <- recodeSMK2

plot(pcaSMK$rotation[, 1], pcaSMK$rotation[, 2], col=recodeSMK_colors[recodeSMK2], 
     main="PC1, PC2 of 23 genes up in current smokers (Gen. Biol.)",
     xlab="PC1", ylab="PC2")
legend("topright", c("Former", "Current", "Uncertain"), fill=recodeSMK_colors)
```
## questions  
1. remove those patients seemingly misclassified or just re-classify?
2. what percentage of unclear patients should I remove?



## Run analysis with new smoking coding and compare to COPDvCOPD+Cancer from model 2
```{r newSmokingCalls}

eset <- holdEset
eset <- removeFactorLevel(eset, "indicator", "1")
eset <- removeFactorLevel(eset, "indicator", "2")
eset <- removeFactorLevel(eset, "SMKc", "1")
esetF <- medianFilter(eset)
model11 <- model.matrix(~1 + FinalCaDXc + AGEcalc + RIN, data=esetF)
fit11 <- lmFit(esetF, model11)
fit11 <- eBayes(fit11)
results11 <- decideTests(fit11, adjust.method="fdr", p.value=0.15)
summary(results11)

inds11 <- which(results11[,2] != 0)
# This will produce the same heatmap as before
generate_heatmap(inds11, esetF, tp="FinalCaDXc")
# save these genes out as gene symbols
save_entrez(inds11, rownms=fit11$genes$Symbol, filename="tempgeneSymbols.model11.COPDvCOPDCancer.txt")
# Now show same map but with new smoking status
esetTemp <- esetF
esetTemp$SMKc <- esetTemp$SMK2
generate_heatmap(inds11, esetTemp, tp="FinalCaDXc")

clabelsT <- cbind("FinalCaDXc" = cancer_colors[esetTemp$FinalCaDXc],
                 "Smoking Status" = recodeSMK_colors[esetTemp$SMKc],
                 "Gender" = gender_colors[esetTemp$GENDERc])
 
heatmap3(esetTemp[inds11,], col = bluered,
         hclustfun=function(d) hclust(d, method="average"),
         col.clustering = esetTemp$FinalCaDXc, ColSideColors = clabelsT,
         main = "Molecular Smoking Codes, COPD v COPD+Cancer")

# Try running same model now but with updated smoking status and removing 'uncertain'
eset <- holdEset
eset$SMK2 <- recodeSMK2
eset <- removeFactorLevel(eset, "indicator", "1")
eset <- removeFactorLevel(eset, "indicator", "2")
eset <- removeFactorLevel(eset, "SMK2", "1")
eset <- removeFactorLevel(eset, "SMK2", "uncertain")
esetF <- medianFilter(eset)

model12 <- model.matrix(~1 + FinalCaDXc + AGEcalc, data=esetF)
fit12 <- lmFit(esetF, model12)
fit12 <- eBayes(fit12)
results12 <- decideTests(fit12, adjust.method="fdr", p.value=0.2)
summary(results12)

inds12 <- which(results12[, 2] != 0)
geneSymbols12 <- fit12$genes$Symbol[inds12]
esetF$SMKc <- esetF$SMK2
generate_heatmap(inds12, esetF, tp="FinalCaDXc")

# The results from above (model2) that indicate at least two distinct clusters of patients
# within each of the COPD and COPD+Cancer groups seems to be replicated here
# Need to check if the clusters are similar or the same

# try using pca with mclust to define clusters within each disease group

pca12 <- prcomp(znorm(x=exprs(esetF)[inds12, ], margin=1), center=FALSE, scale.=FALSE)

clusters12 <- factor(return_cluster(inds12, esetF, n.clusters=4,
                                    type=COLUMNS, colClus=esetF$FinalCaDXc))

clabelsT <- cbind("FinalCaDXc" = cancer_colors[esetF$FinalCaDXc],
                 "Smoking Status" = recodeSMK_colors[esetF$SMK2],
                 "Gender" = gender_colors[esetF$GENDERc],
                  "Cluster" = copdca_colors[clusters12])

hMapTemp <- heatmap3(eset[inds12,], col = bluered,
                     hclustfun=function(d) hclust(d, method="average"),
                     col.clustering = esetF$FinalCaDXc, ColSideColors = clabelsT,
                     main = "Molecular Smoking Codes, COPD and COPD+Cancer Unsupervised",
                     keep.dendro=TRUE)

clusters12 <- cutree(as.hclust(hMapTemp$Colv), h=2)
hMapTemp2 <- heatmap3(eset[inds12,], col = bluered,
                     hclustfun=function(d) hclust(d, method="average"),
                     col.clustering = esetF$FinalCaDXc, ColSideColors = clabelsT,
                     main = "Molecular Smoking Codes, COPD and COPD+Cancer Unsupervised")

eset <- removeFactorLevel(eset, "SMKnew", "uncertain")
eset <- removeFactorLevel(eset, "COPD2_R7", "DK")
eset <- cleanNAForAnalysis(eset, "COPD2_R7")
eset <- calcIndicator(eset, "FinalCaDXc", "COPD2_R7")

model11 <- model.matrix(~1 + FinalCaDXc*AGEcalc + RIN, data=eset)
fit11 <- lmFit(eset, model11)
fit11 <- eBayes(fit11)
results11 <- decideTests(fit11, adjust.method="fdr", p.value=0.15)

```




