
New List for Sunday night:
1. Steiling COPD signal enrichment and/or differential analysis - heatmap
2. Nature medicine cancer enrichment and/or differential analysis - heatmap
3. COPD+Cancer interaction in clean and super clean eset
4. Enrichr results from interaction - look for UPR and EMT


New List for Monday night:
1. Choose eset - choices - clean eset, super clean eset, formers or all smokers
2. Previous signature enrichment
3. Genes by interaction overlap and by t-tests/contrasts
4. Figure of overlap and each set on its own
5. ATF4 and/or emt targets
6. Splity by cluster for enrichment - GSEA, enrichr
7. row labeled heatmaps
8. atf4 leading edge(s) heatmaps in clinical and cell data
9. EMT enrichment by chi-squared
10. EMT leading edge heatmap in clinical (and cell?)
  
```{r defineFunctions, include=FALSE}
foldChange <- function(eset, groups, genes){
  apply(exprs(eset)[genes, groups==0], 1, mean) - 
    apply(exprs(eset)[genes, groups==1], 1, mean)
}

medianFilter <- function(eset){
  md <- median(exprs(eset))
  passFilter <- logical(featureNumber(eset))
  for(i in 1:featureNumber(eset)){
    passFilter[i] <- sum(exprs(eset)[i, ] > md) > 0
  }
  
  eset <- eset[passFilter, ]
  return(eset)
  
}

saveGeneList <- function(analysis, filename=paste(Sys.Date(), "Temp", analysis$name, ".txt", sep="")){
  
  write.table(analysis$geneSymbols, file=filename, quote=FALSE,row.names=FALSE, col.names=FALSE)

}

saveRankedList <- function(fit, varColInd, filename = "tempRankedList"){
  filename = paste(paste(Sys.Date(), colnames(fit$t)[varColInd], filename, sep="_"), "rnk", sep=".")
  print(filename)
  temp <- data.frame(geneSymbols = fit$genes$Symbol, ts = fit$t[, varColInd])
  write.table(temp, file=filename, row.names=FALSE, quote=FALSE, col.names=TRUE, sep="\t", eol="\n")
  
  
}

saveGeneSet <- function(fit, inds, filename = "tempGeneSet", geneSetName="GeneSet", description=""){
  filename = paste(paste(Sys.Date(), filename, sep="_"), "gmx", sep=".")
  print(filename)
  temp <- data.frame(geneSymbols=c(geneSetName, description, fit$genes$Symbol[inds]))
  write.table(temp, file=filename, row.names=FALSE, quote=FALSE, col.names=FALSE)
  
}


```

```{r setup, eval=TRUE, echo=TRUE, include=FALSE, results='hide'}

setwd("/protected/projects/pulmarray/Allegro/COPD_Cancer/experiments/2015-05-03_CBM_ATS_2015/")
source("/protected/projects/pulmarray/Allegro/COPD_Cancer/scripts/AllegroSetup.R")
source("plotGSEA.R")
# fix the one patient with wonky data
# eventually this should just be saved in the RDS file
holdEset <- eset
holdEset$FEV1Pc[holdEset$FEV1Pc==89.2] <- 0.892

emt_genes <- read.csv("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/emt_gene_symbols_byers_2013.csv", header=FALSE)
temp <- as.factor(unique(as.character(emt_genes$V1)))
emt_genes <- data.frame("EMT_Genes" = temp)

GOLD_colors = c("0" = "gray100", "1" = "gray75", "2" = "gray50", "3" = "gray25", "4" = "black")

# load ATF4 BEASB2 gene perturbation data for analysis
atf4_data <- readRDS("/protected/projects/pulmarray/Allegro/ATF4/ATF4data_2015MAR17_ExpressionSet.rds")

lam_sign <- read.delim("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/katie98genesEntrez.txt")
lam_sign[, 3] <- paste(lam_sign[, 2], "_at", sep="")

atf4_targets <- read.delim("ATF4_Targets_Enrichr_TRANSFACJASPAR_270Genes.txt", sep=",", as.is=TRUE, header=FALSE)
atf4_targets <- as.character(atf4_targets)

a4LEup <- read.delim("ATF4_targets_cluster1_270Genes_LeadingEdge_Up.csv", header=FALSE)
a4LEup <- as.character(a4LEup$V1)

a4LEdn <- read.delim("ATF4_targets_cluster2_270Genes_LeadingEdge_Down.csv", header=FALSE)
a4LEdn <- as.character(a4LEdn$V1)
library(znorm)
library(diptest)



```

```{r cleanup}

# Remaining question - should we keep or remove the COPD GOLD Status 1 patients? i.e. mild disease
# What about patients using or their charts indicating they have used corticosteroids?
esetClean <- holdEset
esetClean <- cleanNAForAnalysis(esetClean, "COPD2_R7")
esetClean <- cleanNAForAnalysis(esetClean, "RATIOc")
esetClean <- cleanNAForAnalysis(esetClean, "PYc")
esetClean <- cleanNAForAnalysis(esetClean, "RIN")
esetClean <- removeFactorLevel(esetClean, "COPD2_R7", "DK")
esetClean <- removeFactorLevel(esetClean, "FinalCaDXc", "DK")
esetClean <- removeFactorLevel(esetClean, "COPD2_R7", "DK")
esetClean <- removeFactorLevel(esetClean, "GENDERc", "DK")

esetClean <- calcIndicator(esetClean, "FinalCaDXc", "COPD2_R7")
esetClean$smkindic <- as.numeric(as.character(esetClean$indicator))
esetClean$smkindic[esetClean$SMKc==2] <- esetClean$smkindic[esetClean$SMKc==2] + 4
esetClean$smkindic <- as.factor(esetClean$smkindic)

# Define the GOLD Stage appropriate for all patients
esetClean$GOLD <- factor(x=rep(0, sampleNumber(esetClean)), levels=c("0","1","2","3","4"))
# GOLD Stage 1 Ratio < 0.7, FEV1 >= 8
esetClean$GOLD[esetClean$RATIOc < 0.7 & esetClean$FEV1Pc >= 0.8] <- 1
# GOLD Stage 2 Ratio < 0.7, 50 <= FEV1 < 80
esetClean$GOLD[esetClean$RATIOc < 0.7 & esetClean$FEV1Pc < 0.8] <- 2
# GOLD Stage 3 Ratio < 0.7, 30 <= FEV1 <= 50
esetClean$GOLD[esetClean$RATIOc < 0.7 & esetClean$FEV1Pc < 0.5] <- 3
# GOLD Stage 4 Ratio < 0.7, FEV1 < 30%
esetClean$GOLD[esetClean$RATIOc < 0.7 & esetClean$FEV1Pc < 0.3] <- 4

esetCleanFormers <- removeFactorLevel(esetClean, "SMKc", "1")
# Generate a super clean dataset devoid of all patients with other pulmonary diseases
esetSuperClean <- esetClean
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroPNAc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroTBc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroSarcoidc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroIPFc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroIPFc", "DK")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroAtelectasisc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroAtelectasisc", "DK")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroAsthmac", "DK")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroAsthmac", "1")

# clean up ATF4 data
varLabels(atf4_data) <- "SampleID"
atf4_data$SampleID[grep("ATF4", atf4_data$SampleID)] <- "ATF4"
atf4_data$SampleID[grep("Ctrl", atf4_data$SampleID)] <- "Control"
atf4_data$SampleID <- as.factor(atf4_data$SampleID)

atf4_colors <- c("TRUE" = "red", "FALSE" = "white")


```


```{r modelCreation, eval=TRUE, echo=TRUE, results='hide'}

# Katie uses age, sex, smoking, and pack years
# She also attempts to account for corticosteroids by removing them, repeating
# the analysis and checking for enrichment in the overlapping genes


fdrVals <- c(0.05, 0.1, 0.2)
intrxA <- list()
intrxB <- list()
intrxC <- list()
intrxGenes <- list()

#eset <- medianFilter(esetSuperClean)
eset <- medianFilter(esetCleanFormers)
#eset <- medianFilter(esetClean)
for(i in 1:length(fdrVals)){
  
  intrxA[[i]] <- lmFitWrapper(eset, c("COPD2_R7", "FinalCaDXc", "AGEcalc",
                                     "GENDERc", "PYc", "RIN",
                                    "COPD2_R7*FinalCaDXc"),
                             name="COPD Signal",
                             adjust.method="fdr", p.value=fdrVals[i],
                             varOfInterest=7)
                             
	intrxB[[i]] <- lmFitWrapper(eset, c("FEV1Pc", "FinalCaDXc", "AGEcalc",
                                     "GENDERc", "PYc", "RIN",
                                     "FEV1Pc*FinalCaDXc"),
                             name="COPD Signal",
                             adjust.method="fdr", p.value=fdrVals[i],
                             varOfInterest=7)

	intrxC[[i]] <- lmFitWrapper(eset, c("RATIOc", "FinalCaDXc", "AGEcalc",
                                     "GENDERc", "PYc", "RIN",
                                     "RATIOc*FinalCaDXc"),
                             name="COPD Signal",
                             adjust.method="fdr", p.value=fdrVals[i],
                             varOfInterest=7)

	overlappingGenes <- intrxA[[i]]$geneSymbols[intrxA[[i]]$geneSymbols %in% intrxB[[i]]$geneSymbols]
	intrxGenes[[i]] <- overlappingGenes[overlappingGenes %in% intrxC[[i]]$geneSymbols]

	length(intrxA[[i]]$geneSymbols)
	length(intrxB[[i]]$geneSymbols)
	length(intrxC[[i]]$geneSymbols)
	length(intrxGenes[[i]])
	
}


# generate ATF4 model(s)
atf4Model <- lmFitWrapper(atf4_data, c("SampleID"), name="ATF4 Perturbation Data", adjust.method="fdr",
                          p.value=0.05, varOfInterest=1)

# define a model for comparing COPD and COPD+Cancer from which a gene set and ranked list can be extracted. Vars of interest will be primarily cancer
esetCC <- removeFactorLevel(esetCleanFormers, "COPD2_R7", "0")
esetCC <- medianFilter(esetCC)
ccModel <- lmFitWrapper(esetCC, c("FinalCaDXc", "AGEcalc", "GENDERc", "PYc", "RIN"),
                        name="Investigating cancer signal in COPD patients",
                        adjust.method="fdr", p.value=0.1, varOfInterest=1)

esetCC2 <- removeFactorLevel(esetCleanFormers, "FinalCaDXc", "0")
ccModel2 <- lmFitWrapper(esetCC2, c("COPD2_R7", "AGEcalc", "GENDERc", "PYc", "RIN"),
                        name="Investigating copd signal in Cancer patients",
                        adjust.method="fdr", p.value=0.1, varOfInterest=1)

if(FALSE){
  saveRankedList(ccModel$fit, 2, filename="Cancer_Signal_Formers_clean")
  saveRankedList(ccModel2$fit, 2, filename="COPD_Signal_Formers_clean")
  write.table(intrxGenes[[1]], quote=FALSE, row.names=FALSE, col.names=FALSE, file="308_genes_formers_esetClean_fdr05.txt")
}


```


```{r SteilingCOPD}
#eset <- esetCleanFormers
#forGenes <- lmFitWrapper(eset, c("indicator"), name="For Genes",
#                         adjust.method="fdr",
#                         p.value=0.05, varOfInterest=1)

#cpinds <- match(as.character(lam_sign$X), featureNames(eset))
#cpinds <- cpinds[!is.na(cpinds)]

#eset <- eset[cpinds, ]
#cpMod <- lmFitWrapper(eset, c("COPD2_R7", "GENDERc", "PYc", "SMKc", "AGEcalc"), 
#                      name="Steiling COPD Genes Test", 
#                      adjust.method="fdr", p.value=0.2, varOfInterest=1)

#clabels <- cbind(copd_colors[eset$COPD2_R7])
#heatmap3(exprs(eset), col = bluered,
#         hclustfun=function(d) hclust(d, method="ward.D"),
#         col.clustering = eset$COPD2_R7, ColSideColors = clabels,
#         main = "Steiling Gene Set",
#         labCol="")

# heatmap doesn't look great - let's just use GSEA
# Generate COPD ranked list
#eset <- esetSuperClean
#copdSignal <- lmFitWrapper(eset,
#                           c("COPD2_R7", "GENDERc", "PYc", "SMKc", "AGEcalc"),
#                           name="COPD Signal",
#                           adjust.method="fdr",
#                           p.value=0.05, varOfInterest=1)

#saveRankedList(copdSignal$fit, 2, filename="COPDSignal")

pdf("lam_GSEA_plots.pdf")
plot.gsea("COPD_GSEA_fig_resources/2015-05-04_COPD2_R71_COPD_Signal_Formers_clean.rnk", c("COPD_GSEA_fig_resources/lam_44_genes_down_GSEA.txt", "COPD_GSEA_fig_resources//lam_54_genes_up_GSEA.txt"), lab.rank="COPD t-stat in Allegro Cohort", lab.gene.set=c("Lam 44 Genes Down", "Lam 54 Genes Up"))

dev.off()

```


```{r natmedCancer}

eset <- esetSuperClean
cancerSignal <- lmFitWrapper(eset,
                           c("FinalCaDXc", "GENDERc", "PYc", "SMKc", "AGEcalc"),
                           name="Cancer Signal",
                           adjust.method="fdr",
                           p.value=0.05, varOfInterest=1)

saveRankedList(cancerSignal$fit, 2, filename="CancerSignal")
```



```{r saveGeneSetsGSEA}
# output the list of overlapping genes for A-C esetSuperClean interaction mods
inds <- match(intrxGenes[[1]], intrxA[[1]]$fit$genes$Symbol)
saveGeneSet(intrxA[[1]]$fit, inds, filename="overlappingGenesABC_fdr05", geneSetName="TripleOverlap05", description="Set of 270 genes overlapping the interaction terms for models A-C using esetSuperClean")

# save a ranked list for the interaction terms
saveRankedList(intrxA[[1]]$fit, varColInd=8, filename="esetCleanFormers_intrxA_copdcancer")
saveRankedList(intrxB[[1]]$fit, varColInd=8, filename="eseCleanFormer_intrxB_copdcancer")
saveRankedList(intrxC[[1]]$fit, varColInd=8, filename="esetCleanFormer_intrxC_copdcancer")



```


```{r interactionFigures}

eset <- intrxA[[1]]$eset
eset$indicator <- relevel(relevel(eset$indicator, ref="3"), "1")
genes <- intrxGenes[[1]]
inds <- match(genes, intrxA[[1]]$fit$genes$Symbol)

clabels <- cbind(copdca_colors[eset$indicator])

pdf("2015-05-03_TripleOverlapping_FDR05_270Genes_SuperClean_FormerSmokers.pdf")

heatmap3(exprs(eset)[inds, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = eset$indicator, ColSideColors = clabels,
         main = "270 Genes Significant for Overlapping Interaction Terms\nFDR < 0.05\nFormer Smokers Only",
         labCol="")

dev.off()

# Generate one with ATF4 indicators

rlabels <- cbind(atf4_colors[as.character(genes %in% atf4_targets)])

pdf("2015-05-03_TripleOverlapping_FDR05_270Genes_SuperClean_FormerSmokers_ATF4.pdf")
heatmap3(exprs(eset)[inds, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = eset$indicator, ColSideColors = clabels,
         main = "270 Genes Significant for Overlapping Interaction Terms\nFDR < 0.05\nFormer Smokers Only",
         labCol="", RowSideColors=rlabels)

dev.off()

afCs <- return_cluster(inds, eset, n.clusters=2, mthd="ward.D")
rlabels <- smoking_colors[afCs]
heatmap3(exprs(eset)[inds, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = eset$indicator, ColSideColors = clabels,
         main = "270 Genes Significant for Overlapping Interaction Terms\nFDR < 0.05\nFormer Smokers Only",
         labCol="", RowSideColors=rlabels)

# save clusters 1 and 2
if(FALSE){
  c1 <- inds[afCs==1]
  saveGeneSet(intrxA[[1]]$fit, c1, filename="overlappingGenesABC_fdr05_cluster1", geneSetName="TripleOverlap05_C1", description="Cluster 1 from set of 270 genes overlapping the interaction terms")
  
  c2 <- inds[afCs==2]
    saveGeneSet(intrxA[[1]]$fit, c2, filename="overlappingGenesABC_fdr05_cluster2", geneSetName="TripleOverlap05_C2", description="Cluster 2 from set of 270 genes overlapping the interaction terms")

  }


#a4LEdn, a4LEup
inds <- match(a4LEdn, intrxA[[1]]$fit$genes$Symbol)
genes <- a4LEdn[!is.na(inds)]
inds <- inds[!is.na(inds)]
rlabels <- cbind(atf4_colors[as.character(genes %in% atf4_targets)])

heatmap3(exprs(eset)[inds, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = eset$indicator, ColSideColors = clabels,
         main = "270 Genes Significant for Overlapping Interaction Terms\nFDR < 0.05\nFormer Smokers Only",
         labCol="", RowSideColors=rlabels)

esetTemp <- removeFactorLevel(esetSuperClean, "SMKc", "1")
forGenes <- lmFitWrapper(esetTemp, c("FinalCaDXc"), 1, "fdr", 0.1, "forGenes")
indsUp <- match(a4LEup, forGenes$fit$genes$Symbol)
indsDn <- match(a4LEdn, forGenes$fit$genes$Symbol)
indsDn <- indsDn[!is.na(indsDn)]

clabels <- cbind(copdca_colors[esetTemp$indicator])
heatmap3(exprs(esetTemp)[indsUp, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = eset$indicator, ColSideColors = clabels,
         main = "ATF4Up 270 Genes Significant for Overlapping Interaction Terms\nFDR < 0.05\nFormer Smokers Only",
         labCol="")
```


To get a positive t statistic - 
  If COPD.Cancer - COPD > 0, for a positive t statistic, (Cancer - Normal) can be negative or 0. If it is positive, it must be less so than (C.C - CP)
  If COPD.Cancer - COPD = 0, for a positive t statistic, (Ca - N) must be negative (i.e. normal must be greater than cancer)
If COPD.Cancer - COPD < 0, for a positive t statistic, (Ca - N) must be more negative than (C.C - CP) (i.e. COPD >> COPD.Cancer)
```{r interactionDirectionality}
eset <- esetSuperClean
eset <- removeFactorLevel(eset, "SMKc", "1")

# setup the variables for contrast building
eset$Cancer <- as.character(eset$FinalCaDXc)
eset$Cancer[eset$FinalCaDXc==1] <- "Cancer"
eset$Cancer[eset$FinalCaDXc==0] <- "noCancer"
eset$COPD <- as.character(eset$COPD2_R7)
eset$COPD[eset$COPD2_R7==1] <- "COPD"
eset$COPD[eset$COPD2_R7==0] <- "noCOPD"
eset$CC <- as.factor(paste(eset$Cancer, eset$COPD, sep="."))


design <- model.matrix(~0 + eset$CC)
colnames(design) <- levels(eset$CC)

fit <- lmFit(eset, design)

# of interest: (COPD+Cancer - COPD) - (Cancer - COPD) = COPD+Cancer - Cancer
# realistically we want a few groups of the genes significant for these terms
# 1. the genes differing between COPD.Cancer and Cancer
# 2. the (1) genes and those also changing between COPD.Cancer and COPD
cont.matrix <- makeContrasts(
  #Difference=(Cancer.COPD-noCancer.COPD)-(Cancer.noCOPD-noCancer.COPD),
  #COPDvCancer=Cancer.noCOPD-noCancer.COPD,
  COPDCavCOPD=Cancer.COPD-noCancer.COPD,
  COPDCavCancer=Cancer.COPD-Cancer.noCOPD,
  COPDCavNormal=Cancer.COPD-noCancer.noCOPD,
  COPDvNormal=noCancer.COPD-noCancer.noCOPD,
  CancervNormal=Cancer.noCOPD-noCancer.noCOPD,
  CancervCOPD=Cancer.noCOPD-noCancer.COPD,
  levels=design)


fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)

fdr <- 0.05

inds <- match(intrxGenes[[1]], fit2$genes$Symbol)
fitTemp <- fit2[inds,]
results <- decideTests(fitTemp, adjust.method="none", p.value=fdr)
# view the results at fdr < 0.05
vennDiagram(results[,c(1,5)])
vennDiagram(results[ ,c(1:3, 6)])
vennDiagram(results[, 1:3])

# For positive interaction statistic
# which are greater in COPD.Cancer v COPD but not different or there is a negative change between cancer and normal
ind1 <- which(results[, 1] > 0 & results[, 5] <= 0)
ind2 <- which(results[, 1] == 0 & results[, 5] < 0)

# For negative interaction statistic


# For investigating EMT genes
resultsAll <- decideTests(fit2, adjust.method="fdr", p.value=0.2)
summary(resultsAll)

copdca.contr <- which(resultsAll[, 1] != 0 & resultsAll[, 2] != 0 & resultsAll[, 3] != 0)

if(TRUE){
  saveRankedList(fit2, 1, filename="COPDCavCOPD_contr_esetSuperCleanFormer")
  saveRankedList(fit2, 2, filename="COPDCavCancer_contr_esetSuperCleanFormer")
  saveRankedList(fit2, 3, filename="COPDCavNormal_contr_esetSuperCleanFormer")
}
esetContr <- eset
fitContr <- fit2

# determine if there are more emt genes in the copdca.contr than expected by chance using chisquared test

inEMT <- as.factor(forGenes$fit$genes$Symbol %in% emt_genes$EMT_Genes)
inCCC <- as.factor(forGenes$fit$genes$Symbol %in% fit2$genes$Symbol[copdca.contr])
chisq.test(inEMT, inCCC)

generate_heatmap(copdca.contr, eset, mthd="ward.D", tp="indicator")

```

```{r emt}
#eset <- intrxA[[1]]$eset
eset <- esetContr
eset$indicator <- relevel(relevel(eset$indicator, ref="3"), "1")
emt_inds <- match(emt_genes$EMT_Genes, fitContr$genes$Symbol)
emt_inds <- emt_inds[!is.na(emt_inds)]

eset <- removeFactorLevel(eset, "SMKc", "1")
clabels <- cbind(copdca_colors[eset$indicator], smoking_colors[eset$SMKc])
heatmap3(exprs(eset)[emt_inds, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = eset$indicator, ColSideColors = clabels,
         main = "EMT Genes",
         labCol="")



eset2 <- removeFactorLevel(eset, "FinalCaDXc", "0")
clabels <- cbind(copdca_colors[eset2$indicator], GOLD_colors[eset2$GOLD])
heatmap3(exprs(eset2)[emt_inds, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = eset2$indicator, ColSideColors = clabels,
         main = "EMT Genes",
         labCol="")

```


```{r emtLeadingEdge}
emtLE <- read.table("EMT_intoRatioIntrx_leadingEdge.txt")
eset <- removeFactorLevel(esetSuperClean, "SMKc", "1")
eset$indicator <- relevel(relevel(eset$indicator, ref="3"), "1")

forGenes <- lmFitWrapper(eset, "FinalCaDXc", 1, "fdr", 0.05, "forGenes")
inds <- match(emtLE$V1, forGenes$fit$genes$Symbol)

clabels <- cbind(copdca_colors[eset$indicator])
pdf("emt_leading_edge_genes_inSILCC_data.pdf")
heatmap3(exprs(eset)[inds, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = eset$indicator, ColSideColors = clabels,
         main = "EMT Genes Leading eDge",
         labCol="")
dev.off()

forGenesA <- lmFitWrapper(atf4_data, "SampleID", 1, "fdr", 0.05, "forGenesA")
inds <- match(emtLE$V1, forGenesA$fit$genes$Symbol)


tempColors <- c("Control" = "black", "ATF4" = "red")
clabels <- cbind(tempColors[atf4_data$SampleID])
pdf("emt_leading_edge_genes_intATF4_data.pdf")
heatmap3(exprs(atf4_data)[inds, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = atf4_data$SampleID, ColSideColors = clabels,
         main = "EMT Genes Leading Edge in ATF4 Data",
         labCol="")

dev.off()


```