The Changing COPD Signal in COPD Patients with Cancer
========================================================
`r Sys.Date()`
---------------------------------------------


```{r GlobalVariables, include=FALSE}
cacheOption = FALSE

```


```{r defineFunctions, include=FALSE}
foldChange <- function(eset, groups, genes){
  apply(exprs(eset)[genes, groups==0], 1, mean) - 
    apply(exprs(eset)[genes, groups==1], 1, mean)
}

medianFilter <- function(eset){
  md <- median(exprs(eset))
  passFilter <- logical(featureNumber(eset))
  for(i in 1:featureNumber(eset)){
    passFilter[i] <- sum(exprs(eset)[i, ] > md) > 0
  }
  
  eset <- eset[passFilter, ]
  return(eset)
  
}

```


```{r LoadData, echo=FALSE, cache=cacheOption, include=FALSE}
# set the working directory to the directory of this script
setwd("/restricted/projectnb/pulmarray/LinGA_protected/Allegro/COPD_Cancer/experiments/2014-10-24")

# load the data using a script that removes patients with:
#  Cancer = NA
#  SMK    = 3
source("/protected/projects/pulmarray/Allegro/COPD_Cancer/scripts/AllegroSetup.R")

# clean the data (i.e. remove DKs and NAs)
eset <- removeFactorLevel(eset, "FinalCaDXc", "DK")
eset <- removeFactorLevel(eset, "AllegroCOPDc", "DK")
eset <- cleanNAForAnalysis(eset, "AGEcalc")
eset <- cleanNAForAnalysis(eset, "RIN")
eset <- removeBioReps(eset)
eset <- calcIndicator(eset, "FinalCaDXc", "AllegroCOPDc")

# fix the one patient sample who has a FEV1% predicted of 89.2
eset$FEV1Pc[eset$FEV1Pc==89.2] <- 0.892

eset$BATCH <- factor(eset$BATCH)
holdEset <- eset

library(znorm)
library(mclust)

```

## Thoughts for the day

I need a more directed approach than the semi-amorphous approach I have been taking
Specifically identify the goal of the analysis. This has not really changed. The goal
of the analyis is to find genes related to COPD pathogenesis that are playing a role in 
carcinogenesis of those patients with COPD. The biggest problem I've encountered is the
unexplained heterogeneity of both patient groups. I think that may be OK. The other
problem I've had is that the analysis method I used to get to my Evan's Day results
was not particularly statistically sound - I grouped patients into two groups,
1) COPD+Cancer and 2) disease-free, cancer-only, and COPD-only. On the plus side I did find
genes that seemed to differ between cancer and COPD+Cancer that were still cancer related.
They did not drastically differ between COPD and COPD+Cancer, which again, I think may be OK.

One problem - the COPD-only group is very small
One solution - use bootstrapping to upsample the group

1.) One possible method: Find cancer related genes acting differently in COPD and/or COPD+Cancer
This is not quite right because we're looking for why COPD has increased incidence of cancer

2.) A better possible method: Find the COPD genes (i.e. use Katie's paper as a model)

3.) Another possible idea is to use molecular signals for cancer, COPD, smoking in lieu of clinical
(i.e. use bronchogen, Katie's 98 genes, 100 smoking genes)

4.) Another idea - also use miRNA that Anna is working with to examine the miRNA-mRNA network 
interactions associated with COPD+Cancer

5.) Does the COPD of youth look different from teh COPD of old age
i.e. do genes change with PFTs differently at young age compared with old age
the question here would obviously be related to the field. Does the COPD field look different
in youth compared with elder age. Are low PFTs related to different mechanisms when younger?

6.) Find genes related to the transcription factors from Evans Day Enrichr results and examine
for interesting genes

```{r SOX2relatedGenes}
# Based on paper: "SOX2 Gene Regulates the Transcriptional Netework of Oncogenes and Affects Tumorigenesis of Human Lung Cancer Cells"
# Chen et al., 2012

# Generate model that was used to create Evans Day plots
eset <- holdEset
eset <- cleanNAForAnalysis(eset, "COPD2_R7")
eset <- removeFactorLevel(eset, "COPD2_R7", "DK")
eset <- calcIndicator(eset, "FinalCaDXc", "COPD2_R7")
eset$AllOrNone <- eset$indicator
eset$AllOrNone[eset$indicator != 4] <- 1
eset <- removeFactorLevel(eset, "AllOrNone", "2")
eset <- removeFactorLevel(eset, "AllOrNone", "3")
eset <- removeFactorLevel(eset, "SMKc", "1")
eset <- cleanNAForAnalysis(eset, "PYc")
eset <- removeFactorLevel(eset, "GENDERc", "DK")
esetF <- medianFilter(eset)

analysisED <- lmFitWrapper(esetF, c("AllOrNone", "AGEcalc", "RIN", "PYc", "GENDERc"), 1,
                          p.value=0.15,
                          name="Analysis9: All-or-none, PFT-defined COPD, formers only, PYs + Gender")

analysisED2 <- lmFitWrapper(eset, c("AllOrNone", "AGEcalc", "RIN", "PYc", "GENDERc"), 1,
                          p.value=0.15,
                          name="Analysis9: All-or-none, PFT-defined COPD, formers only, PYs + Gender")

kegg_pathways_in_cancer <- read.table("geneSet_pathwaysInCancer(KEGG).txt", header=TRUE)
kegg_pathways_in_cancer <- as.character(kegg_pathways_in_cancer[, 1])

picKEGGind <- which(analysisED2$fit$genes$Symbol %in% kegg_pathways_in_cancer)
eset <- holdEset[picKEGGind, ]
eset <- cleanNAForAnalysis(eset, "COPD2_R7")
eset <- removeFactorLevel(eset, "COPD2_R7", "DK")
eset <- calcIndicator(eset, "FinalCaDXc", "COPD2_R7")
eset <- removeFactorLevel(eset, "GENDERc", "DK")

esetN <- eset
exprs(esetN) <- znorm(exprs(esetN), margin=1)
ancova1 <- GlobalAncova(xx=exprs(esetN), group=esetN$FinalCaDXc, covars=eset$COPD2_R7)

```

7.) Send Adam a gmt gene list signature to test using OpenSesame

8.) Use GSEA to test my gene list/ranked list against known signals

