In this experiment I am going to check for separation between various groups based on the EMT gene signature from Byers et al., 2013. This signature includes both mesenchymal and epithelial cell markers, so combined should be able to distinguish between the cell types (of note: this signature was developed in tumor data)
  
```{r defineFunctions, include=FALSE}
foldChange <- function(eset, groups, genes){
  apply(exprs(eset)[genes, groups==0], 1, mean) - 
    apply(exprs(eset)[genes, groups==1], 1, mean)
}

medianFilter <- function(eset){
  md <- median(exprs(eset))
  passFilter <- logical(featureNumber(eset))
  for(i in 1:featureNumber(eset)){
    passFilter[i] <- sum(exprs(eset)[i, ] > md) > 0
  }
  
  eset <- eset[passFilter, ]
  return(eset)
  
}

saveGeneList <- function(analysis, filename=paste(Sys.Date(), "Temp", analysis$name, ".txt", sep="")){
  
  write.table(analysis$geneSymbols, file=filename, quote=FALSE,row.names=FALSE, col.names=FALSE)

}

saveRankedList <- function(fit, varColInd, filename = "tempRankedList"){
  filename = paste(paste(Sys.Date(), colnames(fit$t)[varColInd], filename, sep="_"), "rnk", sep=".")
  print(filename)
  temp <- data.frame(geneSymbols = fit$genes$Symbol, ts = fit$t[, varColInd])
  write.table(temp, file=filename, row.names=FALSE, quote=FALSE, col.names=TRUE, sep="\t", eol="\n")
  
  
}

saveGeneSet <- function(fit, inds, filename = "tempGeneSet", geneSetName="GeneSet", description=""){
  filename = paste(paste(Sys.Date(), filename, sep="_"), "gmx", sep=".")
  print(filename)
  temp <- data.frame(geneSymbols=c(geneSetName, description, fit$genes$Symbol[inds]))
  write.table(temp, file=filename, row.names=FALSE, quote=FALSE, col.names=FALSE)
  
}


```

### Setup
```{r setup, eval=TRUE, echo=TRUE, include=FALSE, results='hide'}

setwd("/protected/projects/pulmarray/Allegro/COPD_Cancer/experiments/Expmt_0039_COPD_Cancer_Smoking_GSVA/")
source("/protected/projects/pulmarray/Allegro/COPD_Cancer/scripts/AllegroSetup.R")
#source("../2015-05-03_CBM_ATS_2015/plotGSEA.R")
# fix the one patient with wonky data
# eventually this should just be saved in the RDS file
holdEset <- eset
holdEset$FEV1Pc[holdEset$FEV1Pc==89.2] <- 0.892

emt_genes <- read.csv("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/emt_gene_symbols_byers_2013.csv", header=FALSE)
temp <- as.factor(unique(as.character(emt_genes$V1)))
emt_genes <- data.frame("EMT_Genes" = temp)

GOLD_colors = c("0" = "gray100", "1" = "gray75", "2" = "gray50", "3" = "gray25", "4" = "black")
greyblack <- colorRampPalette(c("black", "white", "grey"))(256)


# load ATF4 BEASB2 gene perturbation data for analysis
atf4_data <- readRDS("/protected/projects/pulmarray/Allegro/ATF4/ATF4data_2015MAR17_ExpressionSet.rds")

lam_sign <- read.delim("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/katie98genesEntrez.txt")
lam_sign[, 3] <- paste(lam_sign[, 2], "_at", sep="")

lamGMT <- read.csv("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/100525_lam_98overlapgenes_covars.gmt", sep="\t", as.is=TRUE, header=FALSE)
lamUP <- as.character(lamGMT[1, 3:56])
lamDN <- as.character(lamGMT[2, 3:46])

joshLM127dn <- read.csv("/protected/projects/pulmarray/HoggCOPD/A1ATGrifols/experiments/Expmt_0001_LM_Parenchyma/2015-06-08_Joshdn127.gmx", skip=1, header=TRUE)

joshLM127up <- read.csv("/protected/projects/pulmarray/HoggCOPD/A1ATGrifols/experiments/Expmt_0001_LM_Parenchyma/2015-06-08_Joshup127.gmx", skip=1, header=TRUE)

# from parenchymal REAL data
cleLMup <- read.csv("/protected/projects/pulmarray/HoggCOPD/A1ATGrifols/experiments/Expmt_0003_LM_Parenchyma_CLE/2015-06-24_parenchyma_cle_geneset_cluster2_fdr_0.05.gmx")

cleLMdn <- read.csv("/protected/projects/pulmarray/HoggCOPD/A1ATGrifols/experiments/Expmt_0003_LM_Parenchyma_CLE/2015-06-24_parenchyma_cle_geneset_cluster1_fdr_0.05.gmx")

natMed <- read.csv("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/Nature_Medicine_gmx_files/top_100_genes_symbols_quality_adjusted.gmx", sep="\t", as.is=TRUE)
natMedUp <- natMed[2:65, 1]
natMedDn <- natMed[2:37, 2]

tgfB <- read.csv("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/msigdb_C2_CGP_genesets/BIOCARTA_TGFB_PATHWAY.gmx", skip=1)
tgfB <- as.character(tgfB$X..TGF.beta.signaling.pathway)



#atf4_targets <- read.delim("ATF4_Targets_Enrichr_TRANSFACJASPAR_270Genes.txt", sep=",", as.is=TRUE, header=FALSE)
#atf4_targets <- as.character(atf4_targets)

#a4LEup <- read.delim("ATF4_targets_cluster1_270Genes_LeadingEdge_Up.csv", header=FALSE)
#a4LEup <- as.character(a4LEup$V1)

#a4LEdn <- read.delim("ATF4_targets_cluster2_270Genes_LeadingEdge_Down.csv", header=FALSE)
#a4LEdn <- as.character(a4LEdn$V1)
library(znorm)
library(diptest)
library(GSVA)
library(annotate)
library(org.Hs.eg.db)
library(corrplot)
library(ggplot2)
library(beeswarm)


```

### Data Cleaning
```{r cleanup}

# Remaining question - should we keep or remove the COPD GOLD Status 1 patients? i.e. mild disease
# What about patients using or their charts indicating they have used corticosteroids?
esetClean <- holdEset
esetClean <- cleanNAForAnalysis(esetClean, "COPD2_R7")
esetClean <- cleanNAForAnalysis(esetClean, "RATIOc")
esetClean <- cleanNAForAnalysis(esetClean, "PYc")
esetClean <- cleanNAForAnalysis(esetClean, "RIN")
esetClean <- removeFactorLevel(esetClean, "COPD2_R7", "DK")
esetClean <- removeFactorLevel(esetClean, "FinalCaDXc", "DK")
esetClean <- removeFactorLevel(esetClean, "COPD2_R7", "DK")
esetClean <- removeFactorLevel(esetClean, "GENDERc", "DK")

esetClean <- calcIndicator(esetClean, "FinalCaDXc", "COPD2_R7")
esetClean$smkindic <- as.numeric(as.character(esetClean$indicator))
esetClean$smkindic[esetClean$SMKc==2] <- esetClean$smkindic[esetClean$SMKc==2] + 4
esetClean$smkindic <- as.factor(esetClean$smkindic)

# Define the GOLD Stage appropriate for all patients
esetClean$GOLD <- factor(x=rep(0, sampleNumber(esetClean)), levels=c("0","1","2","3","4"))
# GOLD Stage 1 Ratio < 0.7, FEV1 >= 8
esetClean$GOLD[esetClean$RATIOc < 0.7 & esetClean$FEV1Pc >= 0.8] <- 1
# GOLD Stage 2 Ratio < 0.7, 50 <= FEV1 < 80
esetClean$GOLD[esetClean$RATIOc < 0.7 & esetClean$FEV1Pc < 0.8] <- 2
# GOLD Stage 3 Ratio < 0.7, 30 <= FEV1 <= 50
esetClean$GOLD[esetClean$RATIOc < 0.7 & esetClean$FEV1Pc < 0.5] <- 3
# GOLD Stage 4 Ratio < 0.7, FEV1 < 30%
esetClean$GOLD[esetClean$RATIOc < 0.7 & esetClean$FEV1Pc < 0.3] <- 4

esetCleanFormers <- removeFactorLevel(esetClean, "SMKc", "1")
esetCleanCurrents <- removeFactorLevel(esetClean, "SMKc", "2")
# Generate a super clean dataset devoid of all patients with other pulmonary diseases
esetSuperClean <- esetClean
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroPNAc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroTBc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroSarcoidc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroIPFc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroIPFc", "DK")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroAtelectasisc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroAtelectasisc", "DK")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroAsthmac", "DK")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroAsthmac", "1")

esetSuperCleanICS <- removeFactorLevel(esetSuperClean, "AllegroCorticosteroidsc", "1")
esetSuperCleanICS <- removeFactorLevel(esetSuperCleanICS, "AllegroCorticosteroidsc", "DK")

# clean up ATF4 data
varLabels(atf4_data) <- "SampleID"
atf4_data$SampleID[grep("ATF4", atf4_data$SampleID)] <- "ATF4"
atf4_data$SampleID[grep("Ctrl", atf4_data$SampleID)] <- "Control"
atf4_data$SampleID <- as.factor(atf4_data$SampleID)

atf4_colors <- c("ATF4" = "red", "Control" = "black")


```

### ATF4 Models
```{r ATF4models}
# generate ATF4 model(s)
atf4Model <- lmFitWrapper(atf4_data, c("SampleID"),
                          name="ATF4 Perturbation Data", adjust.method="none",
                          p.value=0.05, varOfInterest=1)

clabels <- cbind(atf4_colors[as.character(atf4_data$SampleID)])
heatmap3(exprs(atf4_data)[atf4Model$inds, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = atf4_data$SampleID, ColSideColors = clabels,
         main = "ATF4 Genes",
         labCol="")

hmapClusts <- heatmap3(exprs(atf4_data)[atf4Model$inds,], col = bluered,
                       keep.dendro=TRUE,
                       hclustfun=function(d) hclust(d, method="ward.D"),
                       col.clustering = atf4_data$SampleID, ColSideColors=clabels,
                       main="ATF4 Genes", labCol="")

clusters <- cutree(as.hclust(hmapClusts$Rowv), k=2)
atf4Cluster1_dn <- atf4Model$geneSymbols[clusters==1]
atf4Cluster2_up <- atf4Model$geneSymbols[clusters==2]

atf4dn <- names(clusters)[clusters==1]
atf4up <- names(clusters)[clusters==2]
atf4 <- names(clusters)

rlabels <- cbind(smoking_colors[clusters])
heatmap3(exprs(atf4_data)[atf4Model$inds,], col = bluered,
         keep.dendro=TRUE,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = atf4_data$SampleID, ColSideColors=clabels,
         main="ATF4 Genes", labCol="", RowSideColors=rlabels)



```


### Multiplot function definition 
```{r multiplot}
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)
  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)
  numPlots = length(plots)
  
  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                     ncol = cols, nrow = ceiling(numPlots/cols))
    }
  
  if (numPlots==1) {
    print(plots[[1]])
    } else {
      # Set up the page
      grid.newpage()
      pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
      
      # Make each plot, in the correct location
      for (i in 1:numPlots) {
        # Get the i,j matrix positions of the regions that contain this subplot
        matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
        print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                        layout.pos.col = matchidx$col))
        }
      }
  }

```

### EMT separation between current and former healthy patients
```{r emtNeitherSeparation}
eset <- esetClean
# remove the cancer patients
eset <- removeFactorLevel(eset, "FinalCaDXc", "1")
# remove the COPD patients
eset <- removeFactorLevel(eset, "COPD2_R7", "1")

# set the gene symbols as the row names
gnsbs <- getSYMBOL(return_entrez(featureNames(eset)), data='org.Hs.eg')
#eset2 <- eset[!is.na(temp), ]
#featureNames(eset2) <- getSYMBOL(return_entrez(featureNames(eset2)), data='org.Hs.eg')

# fix the emt gene signature so it matches the names pulled into gnsbs
# this was done all online via ncbi - below are the necessary 5 changes
emtGns <- as.character(emt_genes$EMT_Genes)
emtGns[match("MTAC2D1",emtGns)] <- "TC2N"
emtGns[match("C3orf21",emtGns)] <- "XXYLT1"
emtGns[match("LRRC54",emtGns)] <- "TSKU"
emtGns[match("RBM35A",emtGns)] <- "ESRP1"
emtGns[match("TACSTD1",emtGns)] <- "EPCAM"

inds <- match(emtGns, gnsbs)
# PCA plot generation code from Adam's QC Pipeline
emtPCAsmoking2 <- prcomp(na.omit(t(scale(t(exprs(eset)[inds,])))), center=FALSE, scale=FALSE);
percent.variance <- summary(emtPCAsmoking2)$importance["Proportion of Variance", ] * 100;
# Create a PDF of a plot of PC2 vs PC1
#pdf(output.filename);
smk_color = c("1" = "black", "2" = "grey")
pcCols <- smk_color[as.character(eset$SMKc)]
# Note: 10% extra space is alloted in order to give more room for the sample names
plot(emtPCAsmoking2$rotation[,c("PC1","PC2")], pch=21, cex=1, col=pcCols, bg=pcCols,
     xlim=1.1 * range(emtPCAsmoking2$rotation[,"PC1"]),
     ylim=1.1 * range(emtPCAsmoking2$rotation[,"PC2"]),
     xlab=sprintf("PC1 (%2.0f%%)", percent.variance["PC1"]),
     ylab=sprintf("PC2 (%2.0f%%)",percent.variance["PC2"]),
     main="Principal Component Analysis (PCA)\nacross EMT genes in healthy samples"
  )
legend("topright", c("Current", "Former"), col=c("black", "grey"), pch=21, pt.bg=c("black", "grey"), title="Smoking Status", inset=0.01)

# generate a heatmap of the EMT genes, unsupervised, hierarchical clustering
clabels <- cbind(smk_color[as.character(eset$SMKc)])
heatmap3(exprs(eset)[inds, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D2"),
         col.clustering = "unsupervised", ColSideColors = clabels,
         main = "EMT Genes in Non-Diseased Patients",
         labCol="", labRow=emtGns)

# check for differential expression among the EMT genes by Smoking status
smkModel <- lmFitWrapper(eset[inds,], c("SMKc"), 1, "bonferroni", 0.05, "EMT Genes Different by Smoking Status")
heatmap3(exprs(eset)[smkModel$inds, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = "semisupervised", ColSideColors = clabels,
         main = "DE EMT Genes in Non-Diseased Patients",
         labCol="", labRow=smkModel$geneSymbols)


dd <- cbind(t(exprs(eset)[smkModel$inds,]), pData(eset))
names(dd)[1:length(smkModel$inds)] <- smkModel$geneSymbols
p1 <- ggplot(dd, aes(SMKc, GALNT5)) + geom_boxplot(aes(fill=SMKc)) + geom_jitter()
p2 <- ggplot(dd, aes(SMKc, CARD6)) +  geom_boxplot(aes(fill=SMKc)) + geom_jitter()
p3 <- ggplot(dd, aes(SMKc, TMEM45B)) +  geom_boxplot(aes(fill=SMKc)) + geom_jitter()
p4 <- ggplot(dd, aes(SMKc, SERINC2)) +  geom_boxplot(aes(fill=SMKc)) + geom_jitter()
p5 <- ggplot(dd, aes(SMKc, MAPK13)) +  geom_boxplot(aes(fill=SMKc)) + geom_jitter()
p6 <- ggplot(dd, aes(SMKc, GALNT3)) +  geom_boxplot(aes(fill=SMKc)) + geom_jitter()
p7 <- ggplot(dd, aes(SMKc, SPINT2)) +  geom_boxplot(aes(fill=SMKc)) + geom_jitter()
p8 <- ggplot(dd, aes(SMKc, TACSTD2)) +  geom_boxplot(aes(fill=SMKc)) + geom_jitter()
p9 <- ggplot(dd, aes(SMKc, PRSS8)) +  geom_boxplot(aes(fill=SMKc)) + geom_jitter()

multiplot(p1, p2, p3, p4, p5, p6, p7, p8, p9, cols=3)

#SPINT2 acts as a tumor suppressor - here it is decreased significantly in smokers
# see Epigenetic inactivation of SPINT2 is associated with tumor suppressive function in esophageal squamous cell carcinoma Yue et al., 2014
```

### EMT separation between former smokers with and without COPD
```{r emtCOPDSeparation}
# remove the biological replicates
eset <- esetCleanFormers
# remove the cancer patients
eset <- removeFactorLevel(eset, "FinalCaDXc", "1")

# use gnsbs, emtGns, and inds as above

# PCA plot generation code from Adam's QC Pipeline
emtPCACOPD <- prcomp(na.omit(t(scale(t(exprs(eset)[inds,])))), center=FALSE, scale=FALSE);
percent.variance <- summary(emtPCACOPD)$importance["Proportion of Variance", ] * 100;
# Create a plot of PC2 vs PC1
pcCols <- copd_colors[as.character(eset$COPD2_R7)]
# Note: 10% extra space is alloted in order to give more room for the sample names
plot(emtPCACOPD$rotation[,c("PC1","PC2")], pch=21, cex=1, col=pcCols, bg=pcCols,
     xlim=1.1 * range(emtPCAsmoking2$rotation[,"PC1"]),
     ylim=1.1 * range(emtPCAsmoking2$rotation[,"PC2"]),
     xlab=sprintf("PC1 (%2.0f%%)", percent.variance["PC1"]),
     ylab=sprintf("PC2 (%2.0f%%)",percent.variance["PC2"]),
     main="Principal Component Analysis (PCA)\nacross EMT genes in former smokers with or without COPD"
  )
legend("topright", c("no COPD", "COPD"), col=c("bisque", "blue"), pch=21, pt.bg=c("bisque", "blue"), title="COPD Status", inset=0.01)

# generate a heatmap of the EMT genes, unsupervised, hierarchical clustering
clabels <- cbind(copd_colors[as.character(eset$COPD2_R7)])
heatmap3(exprs(eset)[inds, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D2"),
         col.clustering = "unsupervised", ColSideColors = clabels,
         main = "EMT Genes in Former Smokers with or without COPD",
         labCol="", labRow=emtGns)

# check for differential expression among the EMT genes by Smoking status
copdModel <- lmFitWrapper(eset[inds,], c("COPD2_R7"), 1, "bonferroni", 0.05, "EMT Genes Different by COPD Status among Former Smokers")
heatmap3(exprs(eset)[copdModel$inds, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = "semisupervised", ColSideColors = clabels,
         main = "DE EMT Genes in former smokers with or without COPD",
         labCol="", labRow=copdModel$geneSymbols)


dd <- cbind(t(exprs(eset)[copdModel$inds,]), pData(eset))
names(dd)[1:length(smkModel$inds)] <- copdModel$geneSymbols
p1 <- ggplot(dd, aes(COPD2_R7, ANTXR2)) + geom_boxplot(aes(fill=COPD2_R7)) + geom_jitter()
p2 <- ggplot(dd, aes(COPD2_R7, HNMT)) + geom_boxplot(aes(fill=COPD2_R7)) + geom_jitter()
p3 <- ggplot(dd, aes(COPD2_R7, PRSS22)) + geom_boxplot(aes(fill=COPD2_R7)) + geom_jitter()
p4 <- ggplot(dd, aes(COPD2_R7, SH3YL1)) + geom_boxplot(aes(fill=COPD2_R7)) + geom_jitter()
p5 <- ggplot(dd, aes(COPD2_R7, SERINC2)) + geom_boxplot(aes(fill=COPD2_R7)) + geom_jitter()
p6 <- ggplot(dd, aes(COPD2_R7, GRHL1)) + geom_boxplot(aes(fill=COPD2_R7)) + geom_jitter()
p7 <- ggplot(dd, aes(COPD2_R7, C1orf172)) + geom_boxplot(aes(fill=COPD2_R7)) + geom_jitter()
p8 <- ggplot(dd, aes(COPD2_R7, MPZL2)) + geom_boxplot(aes(fill=COPD2_R7)) + geom_jitter()
p9 <- ggplot(dd, aes(COPD2_R7, INADL)) + geom_boxplot(aes(fill=COPD2_R7)) + geom_jitter()

multiplot(p1, p2, p3)
multiplot(p4, p5, p6)
multiplot(p7, p8, p9)

# upregulated in COPD
# 64063=PRSS22, protease, serine, 22: participates in ECM remodeling and overexpression enhanced tumor cell migration and invasion (Chen et al. 2014 Thyroid hormone enhanced human hepatoma cell motility involved PRSS22 activation via ERK signaling); additionally mild association between expression and metastases was observed in the same paper

# 347735=SERINC2/TDE2 is highly expressed in NSCLC Player et al., 2003 (Identification of TDE2 gene and its expression in NSCLC)

# 126695=KDF1/C1orf172: blocks proliferation and promotes differentiation in epidermal progenitor cells (Lee, Kong, and Weatherbee 2013 Forward genetics identifies...)

# downregulated in COPD
# ANTXR2 anthrax toxin receptor 2: binds collagen IV and laminin, suggesting it may be involved in ECM adhesion. aka capillary morphogenesis gene 2 (CMG2), overexpression enhanced adhesion of endothelial cells to ECM but was negatively associated with cell migration (more gene, less migration; more gene more adhesion - suggestive of role in EMT; Ye et al., Therapeutic potential of CMG2 in tumour-related angiogenesis)... overexpression also inhibited angiogenesis
# Ye et al. 2014 CMG2 inhibits growth of breast cancer cells and is inversely correlated with the disease progression and prognosis

# HNMT histamine N-methyltransferase: nothing sticks out as relevant here

# SH3YL1: Dock4 forms a complex with SH3YL1 and regulates cancer cell migration (opposite of expected result here)

# GRHL1 grainyhead-like 1 acts like a TF during development: in combination with Nkx2.1 forms a positive transcriptional loop to coordinate morphogenesis and differentiation of lung epithelial cells (Varma et al., 2012). varma et al., 2014 "Grhl2...maintains alveolar epithelial cell integrity by directly regulating components of adherens and tight junctions" (NOTE - these refer to Grhl2 not Grhl1); Grhl1 acts as a tumor suppressor in neuroblastoma and is negatively regulated by MYCN and HDAC3 (Fabian et al., 2014); cadherin family genes are direct targets of the grainyhead-like TFs (Wilanowski et al., 2008 - Perturbed desmosomal cadherin expression in Grhl1 null mice)

# MPZL2 myelin protein zero like 2: mediates cell-cell adhesion

# INADL: InaD-like, aka PATJ, localizes to tight junctions and to the apical membrane of epithelial cells; PATJ regulates directional migration of mammalian epithelial cells (Shin, Wang, Margolis 2007)
```

### EMT separation between former smokers with COPD +/- Cancer
```{r emtCOPDCancerSeparation}
eset <- esetCleanFormers
# remove the cancer patients
eset <- removeFactorLevel(eset, "COPD2_R7", "0")

# use gnsbs, emtGns, and inds as above

# PCA plot generation code from Adam's QC Pipeline
emtPCACOPD <- prcomp(na.omit(t(scale(t(exprs(eset)[inds,])))), center=FALSE, scale=FALSE);
percent.variance <- summary(emtPCACOPD)$importance["Proportion of Variance", ] * 100;
# Create a plot of PC2 vs PC1
pcCols <- copdca_colors[as.character(eset$indicator)]
# Note: 10% extra space is alloted in order to give more room for the sample names
plot(emtPCACOPD$rotation[,c("PC1","PC2")], pch=21, cex=1, col=pcCols, bg=pcCols,
     xlim=1.1 * range(emtPCAsmoking2$rotation[,"PC1"]),
     ylim=1.1 * range(emtPCAsmoking2$rotation[,"PC2"]),
     xlab=sprintf("PC1 (%2.0f%%)", percent.variance["PC1"]),
     ylab=sprintf("PC2 (%2.0f%%)",percent.variance["PC2"]),
     main="Principal Component Analysis (PCA)\nacross EMT genes in former smokers with COPD +/- Cancer"
  )
legend("topright", c("COPD", "COPD+Cancer"), col=c("green", "blue"), pch=21, pt.bg=c("green", "blue"), title="COPD Status", inset=0.01)

# generate a heatmap of the EMT genes, unsupervised, hierarchical clustering
clabels <- cbind(copdca_colors[as.character(eset$indicator)])
heatmap3(exprs(eset)[inds, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D2"),
         col.clustering = "unsupervised", ColSideColors = clabels,
         main = "EMT Genes in Former Smokers with COPD +/- Cancer",
         labCol="", labRow=emtGns)

# check for differential expression among the EMT genes by Smoking status
copdModel <- lmFitWrapper(eset[inds,], c("FinalCaDXc"), 1, "fdr", 0.05, "EMT Genes Different by Cancer Status among Former Smokers with COPD")
heatmap3(exprs(eset)[copdModel$inds, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = "semisupervised", ColSideColors = clabels,
         main = "DE EMT Genes in former smokers with or without COPD",
         labCol="", labRow=copdModel$geneSymbols)


dd <- cbind(t(exprs(eset)[copdModel$inds,]), pData(eset))
names(dd)[1:length(copdModel$inds)] <- copdModel$geneSymbols
p1 <- ggplot(dd, aes(COPD2_R7, MAL2)) + geom_boxplot(aes(fill=FinalCaDXc)) + geom_jitter()
p2 <- ggplot(dd, aes(COPD2_R7, MPZL2)) + geom_boxplot(aes(fill=FinalCaDXc)) + geom_jitter()
p3 <- ggplot(dd, aes(COPD2_R7, EPB41L5)) + geom_boxplot(aes(fill=FinalCaDXc)) + geom_jitter()

multiplot(p1, p2, p3)

# 3 genes up-regulated in COPD+Cancer v COPD
# MAL2
# MPZL2
# EPB41L5 - erythrocyte membrane protein band 4.1 like 5; overexpression in polarized MDCK cells affects tightness of cell junctions and results in disorganization of the tight junction markers ZO-1 and PATJ (above)
```







### GSVA Results Generation
```{r GSVAresults}
# generate GSVA for lmCLE signatures, atf4, and emt
eset <- esetCleanFormers

genesets <- list(lmCLEup=as.character(cleLMup$GeneSet),
                 lmCLEdn=as.character(cleLMdn$GeneSet),
                 atf4dn=atf4Cluster1_dn,
                 atf4up=atf4Cluster2_up,
                 emt=as.character(emt_genes$EMT_Genes),
                 lamUP=lamUP,
                 lamDN=lamDN,
                 lm127dn=as.character(joshLM127dn$Genes.down.with.LM),
                 lm127up=as.character(joshLM127up$Genes.up.with.LM),
                 natMedDn=natMedDn,
                 natMedUp=natMedUp,
                 tgfBeta=tgfB)

temp <- getSYMBOL(return_entrez(featureNames(eset)), data='org.Hs.eg')
eset2 <- eset[!is.na(temp), ]
featureNames(eset2) <- getSYMBOL(return_entrez(featureNames(eset2)), data='org.Hs.eg')

gsvaResults <- gsva(eset2, genesets)
#gsvaResults <- gsva(eset2, genesets, no.bootstraps=10, method="zscore")

lm_up=1; lm_dn=2; a_up=4; a_dn=3; emt=5;
lam_UP=6; lam_DN=7; l127d=8; l127u=9; canDn=10; canUp=11; tbeta=12

noCopd.noCancer = gsvaResults$es.obs$indicator==1
noCopd.Cancer = gsvaResults$es.obs$indicator==2
copd.noCancer = gsvaResults$es.obs$indicator==3
copd.Cancer = gsvaResults$es.obs$indicator==4

noCancer = gsvaResults$es.obs$FinalCaDXc==0
cancer = gsvaResults$es.obs$FinalCaDXc==1

copd = gsvaResults$es.obs$COPD2_R7==1
noCopd = gsvaResults$es.obs$COPD2_R7==0


```

### T-Tests and Boxplots for COPD signatures
```{r testsAndPlotsCOPD}
# Test COPD signatures in non-cancer patients
t.test(exprs(gsvaResults$es.obs)[lamUP, noCopd.noCancer], exprs(gsvaResults$es.obs)[lamUP, copd.noCancer])

t.test(exprs(gsvaResults$es.obs)[lamDN, noCopd.noCancer], exprs(gsvaResults$es.obs)[lamDN, copd.noCancer])

# Test COPD signatures in cancer patients
t.test(exprs(gsvaResults$es.obs)[lamUP, noCopd.Cancer], exprs(gsvaResults$es.obs)[lamUP, copd.Cancer])

t.test(exprs(gsvaResults$es.obs)[lamDN, noCopd.Cancer], exprs(gsvaResults$es.obs)[lamDN, copd.Cancer])

# plot the scores down by COPD status across all patients
plot(exprs(gsvaResults$es.obs)[lamDN,] ~ gsvaResults$es.obs$indicator,
     col=copdca_colors,
     names=c("Neither", "Cancer", "COPD", "COPD+Cancer"),
     xlab="Group",
     ylab="Metagene for Genes down with COPD",
     main="Metagenes for COPD-Down genes across groups",
     notch=TRUE)

# plot the scores up by COPD status across all patients
plot(exprs(gsvaResults$es.obs)[lamUP,] ~ gsvaResults$es.obs$indicator,
     col=copdca_colors,
     names=c("Neither", "Cancer", "COPD", "COPD+Cancer"),
     xlab="Group", 
     ylab="Metagene for Genes up with COPD",
     main="Metagenes for COPD-Up genes across groups",
     notch=TRUE)



```

### T-Tests and Boxplots for Cancer signatures
```{r testsAndPlotsCOPD}
# Test cancer signatures in non-COPD patients
t.test(exprs(gsvaResults$es.obs)[canDn, noCopd.noCancer], exprs(gsvaResults$es.obs)[canDn, noCopd.Cancer])

t.test(exprs(gsvaResults$es.obs)[canUp, noCopd.noCancer], exprs(gsvaResults$es.obs)[canUp, noCopd.Cancer])

# Test COPD signatures in COPD patients
t.test(exprs(gsvaResults$es.obs)[canDn, copd.noCancer], exprs(gsvaResults$es.obs)[canDn, copd.Cancer])

t.test(exprs(gsvaResults$es.obs)[canUp, copd.noCancer], exprs(gsvaResults$es.obs)[canUp, copd.Cancer])

```

### Test for COPD Signature Differences within COPD patients by Cancer
```{r CancerChangeCOPDSignature}
t.test(exprs(gsvaResults$es.obs)[lamUP, copd.noCancer], exprs(gsvaResults$es.obs)[lamUP, copd.Cancer])

t.test(exprs(gsvaResults$es.obs)[lamDN, copd.noCancer], exprs(gsvaResults$es.obs)[lamDN, copd.Cancer])
```

### EMT Signature
```{r emtSignatureTesting}
target <- emt
# Test the EMT signature in Control v COPD
t.test(exprs(gsvaResults$es.obs)[target, noCopd.noCancer], exprs(gsvaResults$es.obs)[target, copd.noCancer])

# Test the EMT signature in Control v Cancer
t.test(exprs(gsvaResults$es.obs)[target, noCopd.noCancer], exprs(gsvaResults$es.obs)[target, noCopd.Cancer])

# Test the EMT signature in Control v COPD+Cancer
t.test(exprs(gsvaResults$es.obs)[target, noCopd.noCancer], exprs(gsvaResults$es.obs)[target, copd.Cancer])

# Test the EMT signature in Cancer v COPD+Cancer
t.test(exprs(gsvaResults$es.obs)[target, noCopd.Cancer], exprs(gsvaResults$es.obs)[target, copd.Cancer])

# Test the EMT signature in COPD v COPD+Cancer
t.test(exprs(gsvaResults$es.obs)[target, copd.noCancer], exprs(gsvaResults$es.obs)[target, copd.Cancer])

# Test the EMT signature in COPD v Cancer
t.test(exprs(gsvaResults$es.obs)[target, copd.noCancer], exprs(gsvaResults$es.obs)[target, noCopd.Cancer])

# Test the EMT signature in all-comers, COPD v no COPD
t.test(exprs(gsvaResults$es.obs)[target, copd], exprs(gsvaResults$es.obs)[target, noCopd])

# Test the EMT signature in all-comers, Cancer v no Cancer
t.test(exprs(gsvaResults$es.obs)[target, cancer], exprs(gsvaResults$es.obs)[target, noCancer])


# plot the scores for EMT up by COPD status across all patients
plot(exprs(gsvaResults$es.obs)[target,] ~ gsvaResults$es.obs$indicator,
     col=copdca_colors,
     names=c("Neither", "Cancer", "COPD", "COPD+Cancer"),
     xlab="Group", 
     ylab="Metagene for EMT Genes",
     main="Metagenes for EMT genes across groups",
     notch=TRUE)

qplot(gsvaResults$es.obs$indicator, exprs(gsvaResults$es.obs)[target,],
     col=copdca_colors[gsvaResults$es.obs$indicator],
     names=c("Neither", "Cancer", "COPD", "COPD+Cancer"),
     xlab="Group", 
     ylab="Metagene for EMT Genes",
     main="Metagene for EMT genes across groups")

beeswarm(exprs(gsvaResults$es.obs)[target,] ~ gsvaResults$es.obs$indicator,
     col=copdca_colors,
     xlab="Group", 
     ylab="Metagene for EMT Genes",
     main="Metagenes for EMT genes across groups",
     labels=c("Neither", "Cancer", "COPD", "COPD+Cancer"))


```

### Plots of GSVA ATF4, EMT scores by disease status
```{r gsvaBoxPlots}
tt <- data.frame(gsva=c(exprs(gsvaResults$es.obs)[3, ], exprs(gsvaResults$es.obs)[4, ], exprs(gsvaResults$es.obs)[5, ]), geneset=c(rep(names(genesets)[3], times=396), rep(names(genesets)[4], times=396), rep(names(genesets)[5], times=396)), indicator=rep(gsvaResults$es.obs$indicator, times=3))

p <- ggplot(tt, aes(factor(geneset), gsva))

p + geom_boxplot(aes(fill=factor(indicator)))

cor.test(exprs(gsvaResults$es.obs)[a_up, ], exprs(gsvaResults$es.obs)[emt, ])
plot(exprs(gsvaResults$es.obs)[a_up, ], exprs(gsvaResults$es.obs)[emt, ],
     col=copdca_colors[gsvaResults$es.obs$indicator],
     main="ATF4 Up Genes vs EMT", xlab="ATF4 Up Genes", ylab="EMT Genes")
p <- ggplot(as.data.frame(t(exprs(gsvaResults$es.obs))), aes(atf4up, emt))
p + geom_point(aes(colour=copdca_colors[gsvaResults$es.obs$indicator]))

```


### GSVA Correlation Matrix in all Patients
```{r gsvaCorMatrixall}

target <- gsvaResults$es.obs
numGeneSets <- featureNumber(target)
numComps <- ((numGeneSets*numGeneSets - numGeneSets) / 2)
gsvaCors <- gsvaCorPs <- gsvaCorPsBonf <- matrix(ncol=numGeneSets, nrow=numGeneSets)
rownames(gsvaCors) <- colnames(gsvaCors) <- rownames(gsvaCorPs) <- colnames(gsvaCorPs) <- featureNames(target)
 
for(i in 1:numGeneSets){
  for(j in 1:numGeneSets){
    gsvaCors[i,j] <- cor(t(exprs(target[i, ])), t(exprs(target[j,])))
    gsvaCorPs[i,j] <- cor.test(t(exprs(target[i, ])), t(exprs(target[j,])))$p.value
    gsvaCorPsBonf[i,j] <- gsvaCorPs[i,j]*numComps
  } 
}

#corrplot(gsvaCors, method="circle", cl.ratio=0.2, tl.cex=0.8, order='hclust')
#corrplot(gsvaCors, tl.cex=0.8, order='hclust', tl.pos='l', tl.srt=45)

#use the p-values
corrplot.mixed(gsvaCors, tl.cex=0.8, p.mat=gsvaCorPs, insig="blank", tl.pos='l', tl.srt=30, sig.level= 0.00005, main="All Patients, p < 0.00005")
#order='hclust'

# use bonferroni corrected the pvalues
corrplot.mixed(gsvaCors, tl.cex=0.8, p.mat=gsvaCorPsBonf, insig="blank", tl.pos='l', tl.srt=30, sig.level= 0.000001, main="All Patients, Corrected p < 0.000001")

pdf(paste(Sys.Date(), "gsvaCorMatrix_allPatients.pdf", sep="_"))
corrplot.mixed(gsvaCors, tl.cex=0.8, p.mat=gsvaCorPs, insig="blank", tl.pos='l', tl.srt=30, sig.level= 0.0001, main="AllPatients, p < 0.0001")
dev.off()

dd <- as.data.frame(t(exprs(gsvaResults$es.obs)))
dd <- cbind(dd, pData(gsvaResults$es.obs))
p <- ggplot(dd, aes(atf4up, emt)) + geom_boxplot()

p + facet_grid(COPD2_R7 ~ FinalCaDXc)

#geom_point(aes(colour=copdca_colors[gsvaResults$es.obs$indicator]))
```

### GSVA Correlation Matrix in Patients without Cancer
```{r gsvaCorMatrixnoCancer}
target <- gsvaResults$es.obs[, noCancer]
numGeneSets <- featureNumber(target)
gsvaCors <- gsvaCorPs <- matrix(ncol=numGeneSets, nrow=numGeneSets)
rownames(gsvaCors) <- colnames(gsvaCors) <- rownames(gsvaCorPs) <- colnames(gsvaCorPs) <- featureNames(target)
 
for(i in 1:numGeneSets){
  for(j in 1:numGeneSets){
    gsvaCors[i,j] <- cor(t(exprs(target[i, ])), t(exprs(target[j,])))
    gsvaCorPs[i,j] <- cor.test(t(exprs(target[i, ])), t(exprs(target[j,])))$p.value
  } 
}

#use the p-values to plot the correlation matrix (only those that are significant)
corrplot.mixed(gsvaCors, tl.cex=0.8, p.mat=gsvaCorPs, insig="blank", tl.pos='l', tl.srt=30, sig.level= 0.0001, main="Patients without cancer, p < 0.0001")

pdf(paste(Sys.Date(), "gsvaCorMatrix_noCancer.pdf", sep="_"))
corrplot.mixed(gsvaCors, tl.cex=0.8, p.mat=gsvaCorPs, insig="blank", tl.pos='l', tl.srt=30, sig.level= 0.0001, main="Patients without cancer, p < 0.0001")
dev.off()

```

### GSVA Correlation Matrix in Patients with Cancer
```{r gsvaCorMatrixCancer}
target <- gsvaResults$es.obs[, cancer]
numGeneSets <- featureNumber(target)
gsvaCors <- gsvaCorPs <- matrix(ncol=numGeneSets, nrow=numGeneSets)
rownames(gsvaCors) <- colnames(gsvaCors) <- rownames(gsvaCorPs) <- colnames(gsvaCorPs) <- featureNames(target)
 
for(i in 1:numGeneSets){
  for(j in 1:numGeneSets){
    gsvaCors[i,j] <- cor(t(exprs(target[i, ])), t(exprs(target[j,])))
    gsvaCorPs[i,j] <- cor.test(t(exprs(target[i, ])), t(exprs(target[j,])))$p.value
  } 
}

#use the p-values to plot the correlation matrix (only those that are significant)
corrplot.mixed(gsvaCors, tl.cex=0.8, p.mat=gsvaCorPs, insig="blank", tl.pos='l', tl.srt=30, sig.level= 0.0001, main="Patients with cancer, p < 0.0001")

pdf(paste(Sys.Date(), "gsvaCorMatrix_cancer.pdf", sep="_"))
corrplot.mixed(gsvaCors, tl.cex=0.8, p.mat=gsvaCorPs, insig="blank", tl.pos='l', tl.srt=30, sig.level= 0.0001, main="Patients with cancer, p < 0.0001")
dev.off()

```

### GSVA Correlation Matrix in Patients with COPD and without Cancer
```{r gsvaCorMatrixnCOPDnoCancer}
target <- gsvaResults$es.obs[, copd.noCancer]
numGeneSets <- featureNumber(target)
gsvaCors <- gsvaCorPs <- matrix(ncol=numGeneSets, nrow=numGeneSets)
rownames(gsvaCors) <- colnames(gsvaCors) <- rownames(gsvaCorPs) <- colnames(gsvaCorPs) <- featureNames(target)
 
for(i in 1:numGeneSets){
  for(j in 1:numGeneSets){
    gsvaCors[i,j] <- cor(t(exprs(target[i, ])), t(exprs(target[j,])))
    gsvaCorPs[i,j] <- cor.test(t(exprs(target[i, ])), t(exprs(target[j,])))$p.value
  } 
}

#use the p-values to plot the correlation matrix (only those that are significant)
corrplot.mixed(gsvaCors, tl.cex=0.8, p.mat=gsvaCorPs, insig="blank", tl.pos='l', tl.srt=30, sig.level= 0.0001, main="Patients with COPD and without cancer, p < 0.0001")

pdf(paste(Sys.Date(), "gsvaCorMatrix_copdnoCancer.pdf", sep="_"))
corrplot.mixed(gsvaCors, tl.cex=0.8, p.mat=gsvaCorPs, insig="blank", tl.pos='l', tl.srt=30, sig.level= 0.0001, main="Patients with COPD and without cancer, p < 0.0001")
dev.off()

```

### GSVA Correlation Matrix in Patients with COPD and with Cancer
```{r gsvaCorMatrixCOPDCancer}
target <- gsvaResults$es.obs[, copd.Cancer]
numGeneSets <- featureNumber(target)
gsvaCors <- gsvaCorPs <- matrix(ncol=numGeneSets, nrow=numGeneSets)
rownames(gsvaCors) <- colnames(gsvaCors) <- rownames(gsvaCorPs) <- colnames(gsvaCorPs) <- featureNames(target)
 
for(i in 1:numGeneSets){
  for(j in 1:numGeneSets){
    gsvaCors[i,j] <- cor(t(exprs(target[i, ])), t(exprs(target[j,])))
    gsvaCorPs[i,j] <- cor.test(t(exprs(target[i, ])), t(exprs(target[j,])))$p.value
  } 
}

#use the p-values to plot the correlation matrix (only those that are significant)
corrplot.mixed(gsvaCors, tl.cex=0.8, p.mat=gsvaCorPs, insig="blank", tl.pos='l', tl.srt=30, sig.level= 0.0001, main="Patients with COPD and with cancer, p < 0.0001")

pdf(paste(Sys.Date(), "gsvaCorMatrix_copdCancer.pdf", sep="_"))
corrplot.mixed(gsvaCors, tl.cex=0.8, p.mat=gsvaCorPs, insig="blank", tl.pos='l', tl.srt=30, sig.level= 0.0001, main="Patients with COPD and with cancer, p < 0.0001")
dev.off()

```

### GSVA Correlation Matrix in Patients without COPD
```{r gsvaCorMatrixnoCOPD}
target <- gsvaResults$es.obs[, noCopd]
numGeneSets <- featureNumber(target)
gsvaCors <- gsvaCorPs <- matrix(ncol=numGeneSets, nrow=numGeneSets)
rownames(gsvaCors) <- colnames(gsvaCors) <- rownames(gsvaCorPs) <- colnames(gsvaCorPs) <- featureNames(target)
 
for(i in 1:numGeneSets){
  for(j in 1:numGeneSets){
    gsvaCors[i,j] <- cor(t(exprs(target[i, ])), t(exprs(target[j,])))
    gsvaCorPs[i,j] <- cor.test(t(exprs(target[i, ])), t(exprs(target[j,])))$p.value
  } 
}

#use the p-values to plot the correlation matrix (only those that are significant)
corrplot.mixed(gsvaCors, tl.cex=0.8, p.mat=gsvaCorPs, insig="blank", tl.pos='l', tl.srt=30, sig.level= 0.0001, main="Patients without COPD, p < 0.0001")

pdf(paste(Sys.Date(), "gsvaCorMatrix_noCopd.pdf", sep="_"))
corrplot.mixed(gsvaCors, tl.cex=0.8, p.mat=gsvaCorPs, insig="blank", tl.pos='l', tl.srt=30, sig.level= 0.0001, main="Patients without COPD, p < 0.0001")
dev.off()

```

### GSVA Correlation Matrix in Patients with COPD
```{r gsvaCorMatrixCOPD}
target <- gsvaResults$es.obs[, copd]
numGeneSets <- featureNumber(target)
gsvaCors <- gsvaCorPs <- matrix(ncol=numGeneSets, nrow=numGeneSets)
rownames(gsvaCors) <- colnames(gsvaCors) <- rownames(gsvaCorPs) <- colnames(gsvaCorPs) <- featureNames(target)
 
for(i in 1:numGeneSets){
  for(j in 1:numGeneSets){
    gsvaCors[i,j] <- cor(t(exprs(target[i, ])), t(exprs(target[j,])))
    gsvaCorPs[i,j] <- cor.test(t(exprs(target[i, ])), t(exprs(target[j,])))$p.value
  } 
}

#use the p-values to plot the correlation matrix (only those that are significant)
corrplot.mixed(gsvaCors, tl.cex=0.8, p.mat=gsvaCorPs, insig="blank", tl.pos='l', tl.srt=30, sig.level= 0.0001, main="Patients with COPD, p < 0.0001")

pdf(paste(Sys.Date(), "gsvaCorMatrix_copd.pdf", sep="_"))
corrplot.mixed(gsvaCors, tl.cex=0.8, p.mat=gsvaCorPs, insig="blank", tl.pos='l', tl.srt=30, sig.level= 0.0001, main="Patients with COPD, p < 0.0001")
dev.off()

```


### Does ATF4 up or down differ by COPD Status?
```{r atf4byCOPD}
# Test the ATF4 scores in all patients by COPD status, regardless of cancer status
t.test(exprs(gsvaResults$es.obs)[a_up, noCopd],exprs(gsvaResults$es.obs)[a_up, copd])
t.test(exprs(gsvaResults$es.obs)[a_dn, noCopd],exprs(gsvaResults$es.obs)[a_dn, copd])

# Test the ATF4 scores by COPD status, only in no cancer
t.test(exprs(gsvaResults$es.obs)[a_up, noCopd.noCancer],exprs(gsvaResults$es.obs)[a_up, copd.noCancer])
t.test(exprs(gsvaResults$es.obs)[a_dn, noCopd.noCancer],exprs(gsvaResults$es.obs)[a_dn, copd.noCancer])

# Test the ATF4 scores by COPD status, only in cancer
t.test(exprs(gsvaResults$es.obs)[a_up, noCopd.Cancer],exprs(gsvaResults$es.obs)[a_up, copd.Cancer])
t.test(exprs(gsvaResults$es.obs)[a_dn, noCopd.Cancer],exprs(gsvaResults$es.obs)[a_dn, copd.Cancer])

plot(exprs(gsvaResults$es.obs)[a_dn, ] ~ gsvaResults$es.obs$indicator, col=copdca_colors, main="Metagene Score for Genes Decreasing With ATF4 Overexpression", names=c("No Disease", "Cancer", "COPD", "COPD+Cancer"), ylab="Metagene Score")

plot(exprs(gsvaResults$es.obs)[a_up, ] ~ gsvaResults$es.obs$indicator, col=copdca_colors, main="Metagene Score for Genes Increasing With ATF4 Overexpression", names=c("No Disease", "Cancer", "COPD", "COPD+Cancer"), ylab="Metagene Score")

```
When tested across all patients, atf4 does not differ by COPD  
When tested across only patients without cancer both are made more significant  
  + atf4-down genes p < 0.01; COPD increased in these genes  
  
### Does ATF4 correlate with function, PFTS?
```{r atf4CorPFTs}
# Test the correlation between function and atf4
# Hypothesis: as atf4 up goes up, better function (protective), PFTs increase
# corrolary: as atf4 dn goes up, worse function (damaged mechanism), PFTs decrease
cor.test(exprs(gsvaResults$es.obs)[a_up, noCancer], gsvaResults$es.obs$RATIOc[noCancer])
cor.test(exprs(gsvaResults$es.obs)[a_dn, noCancer], gsvaResults$es.obs$RATIOc[noCancer])

# test in cancer patients - nothing significant.
# almost as expected; if the transition is based on field selection, then PFTs would correlate with ATF4 up until the point of transformation
cor.test(exprs(gsvaResults$es.obs)[a_up, cancer], gsvaResults$es.obs$FEV1c[cancer])
cor.test(exprs(gsvaResults$es.obs)[a_dn, cancer], gsvaResults$es.obs$FEV1c[cancer])
```
When tested in patients without cancer atf4-down genes significantly, negatively correlated with RATIO, as expected  

### Does ATF4 up or down differ by cancer status?
```{r atf4byCancer}
# Test the ATF4 scores in all patients by cancer status, regardless of COPD status
t.test(exprs(gsvaResults$es.obs)[a_up, noCancer],exprs(gsvaResults$es.obs)[a_up, cancer])
t.test(exprs(gsvaResults$es.obs)[a_dn, noCancer],exprs(gsvaResults$es.obs)[a_dn, cancer])

# Test the ATF4 scores by cancer status, only in no COPD
t.test(exprs(gsvaResults$es.obs)[a_up, noCopd.noCancer],exprs(gsvaResults$es.obs)[a_up, noCopd.Cancer])
t.test(exprs(gsvaResults$es.obs)[a_dn, noCopd.noCancer],exprs(gsvaResults$es.obs)[a_dn, noCopd.Cancer])

# Test the ATF4 scores by cancer status, only in COPD patients
t.test(exprs(gsvaResults$es.obs)[a_up, copd.noCancer],exprs(gsvaResults$es.obs)[a_up, copd.Cancer])
t.test(exprs(gsvaResults$es.obs)[a_dn, copd.noCancer],exprs(gsvaResults$es.obs)[a_dn, copd.Cancer])


```
+ When tested across all patients, atf4 up tends towards increase in cancer, and atf4 down tends towards decrease in cancer  
+ When tested across only noCOPD patients, there are no differences  
+ When tested across only COPD patients, there are differences in both atf4 up and down genes. atf4-up increased in cancer patients with COPD. atf4-down genes decreased in cancer patients with copd  



Furthermore, Bishawi, Moore, and Bilfinger (2013) support the idea that NSCLC occurs more readily in areas of increased emphysematous damage, which is supported by this claim. This means that if we look at the lm signatures we should see correlation between the signatures of ATF4 and LM (emphysema severity)  
  + in copd patients without cancer (deficient ATF4 response, cells dying, increasing emphysema). As the deficiency increases we should see worse emphysema and increased lm up genes signatures.  
  + in noCancer patients, we should see a correlation between ATF4 and LM. specifically, as atf4-dn goes up, we should see lm-up go up (positive correlation)  
  + as atf4-up goes down, we should see lm-up go up (negative correlation)
  + the lower the score for atf4 genes that are supposed to be up, the higher the score for lm genes that go up and the lower the score for the lm genes that go down
  
### Correlation between ATF4 and LM Scores
```{r lmCorATF4}
# Test and plot the no cancer patients
# ATF4 down genes
# LM up genes
cor.test(exprs(gsvaResults$es.obs)[a_dn, noCancer], exprs(gsvaResults$es.obs)[lm_up, noCancer])
plot(exprs(gsvaResults$es.obs)[a_dn, noCancer], exprs(gsvaResults$es.obs)[lm_up, noCancer], col=copdca_colors[gsvaResults$es.obs$indicator[noCancer]])
# LM down genes
cor.test(exprs(gsvaResults$es.obs)[a_dn, noCancer], exprs(gsvaResults$es.obs)[lm_dn, noCancer])
plot(exprs(gsvaResults$es.obs)[a_dn, noCancer], exprs(gsvaResults$es.obs)[lm_dn, noCancer], col=copdca_colors[gsvaResults$es.obs$indicator[noCancer]])

# ATF4 up genes
# LM up genes
cor.test(exprs(gsvaResults$es.obs)[a_up, noCancer], exprs(gsvaResults$es.obs)[lm_up, noCancer])
plot(exprs(gsvaResults$es.obs)[a_up, noCancer], exprs(gsvaResults$es.obs)[lm_up, noCancer], col=copdca_colors[gsvaResults$es.obs$indicator[noCancer]])
# LM down genes
cor.test(exprs(gsvaResults$es.obs)[a_up, noCancer], exprs(gsvaResults$es.obs)[lm_dn, noCancer])
plot(exprs(gsvaResults$es.obs)[a_up, noCancer], exprs(gsvaResults$es.obs)[lm_dn, noCancer], col=copdca_colors[gsvaResults$es.obs$indicator[noCancer]])

# Test and plot the cancer patients
# ATF4 down genes
# LM up genes
cor.test(exprs(gsvaResults$es.obs)[a_dn, cancer], exprs(gsvaResults$es.obs)[lm_up, cancer])
plot(exprs(gsvaResults$es.obs)[a_dn, cancer], exprs(gsvaResults$es.obs)[lm_up, cancer], col=copdca_colors[gsvaResults$es.obs$indicator[cancer]])
# LM down genes
cor.test(exprs(gsvaResults$es.obs)[a_dn, cancer], exprs(gsvaResults$es.obs)[lm_dn, cancer])
plot(exprs(gsvaResults$es.obs)[a_dn, cancer], exprs(gsvaResults$es.obs)[lm_dn, cancer], col=copdca_colors[gsvaResults$es.obs$indicator[cancer]])

# ATF4 up genes
# LM up genes
cor.test(exprs(gsvaResults$es.obs)[a_up, cancer], exprs(gsvaResults$es.obs)[lm_up, cancer])
plot(exprs(gsvaResults$es.obs)[a_up, cancer], exprs(gsvaResults$es.obs)[lm_up, cancer], col=copdca_colors[gsvaResults$es.obs$indicator[cancer]])
# LM down genes
cor.test(exprs(gsvaResults$es.obs)[a_up, cancer], exprs(gsvaResults$es.obs)[lm_dn, cancer])
plot(exprs(gsvaResults$es.obs)[a_up, cancer], exprs(gsvaResults$es.obs)[lm_dn, cancer], col=copdca_colors[gsvaResults$es.obs$indicator[cancer]])

# Test and plot all patients
# ATF4 down genes
# LM up genes
cor.test(exprs(gsvaResults$es.obs)[a_dn, ], exprs(gsvaResults$es.obs)[lm_up, ])
plot(exprs(gsvaResults$es.obs)[a_dn, ], exprs(gsvaResults$es.obs)[lm_up, ], col=copdca_colors[gsvaResults$es.obs$indicator])
# LM down genes
cor.test(exprs(gsvaResults$es.obs)[a_dn, ], exprs(gsvaResults$es.obs)[lm_dn, ])
plot(exprs(gsvaResults$es.obs)[a_dn, ], exprs(gsvaResults$es.obs)[lm_dn, ], col=copdca_colors[gsvaResults$es.obs$indicator])

# ATF4 up genes
# LM up genes
cor.test(exprs(gsvaResults$es.obs)[a_up,], exprs(gsvaResults$es.obs)[lm_up, ])
plot(exprs(gsvaResults$es.obs)[a_up, ], exprs(gsvaResults$es.obs)[lm_up, ], col=copdca_colors[gsvaResults$es.obs$indicator])
# LM down genes
cor.test(exprs(gsvaResults$es.obs)[a_up,], exprs(gsvaResults$es.obs)[lm_dn, ])
plot(exprs(gsvaResults$es.obs)[a_up, ], exprs(gsvaResults$es.obs)[lm_dn, ], col=copdca_colors[gsvaResults$es.obs$indicator])


```

### Intersection between ATF4 and LM genesets
```{r atf4LMOverlap}
intersect(genesets[["atf4dn"]], genesets[["lm127dn"]])
intersect(genesets[["atf4dn"]], genesets[["lm127up"]])
intersect(genesets[["atf4up"]], genesets[["lm127dn"]])
intersect(genesets[["atf4up"]], genesets[["lm127up"]])

```

### Does EMT differ between COPD and COPD+Cancer patients
```{r emtGSVA}
# test between copd and copd+Cancer
t.test(exprs(gsvaResults$es.obs)[emtu,copd.noCancer], exprs(gsvaResults$es.obs)[emtu,copd.Cancer])
t.test(exprs(gsvaResults$es.obs)[emtd,copd.noCancer], exprs(gsvaResults$es.obs)[emtd,copd.Cancer])

# test between copd and no copd
t.test(exprs(gsvaResults$es.obs)[emtu,noCopd.noCancer], exprs(gsvaResults$es.obs)[emtu,copd.noCancer])
t.test(exprs(gsvaResults$es.obs)[emtd,noCopd.noCancer], exprs(gsvaResults$es.obs)[emtd,copd.noCancer])


```


### Does EMT correlate with ATF4 or LM
```{r emtCorrelations}
# first check for relationship between EMT clusters
cor.test(exprs(gsvaResults$es.obs)[emtu,], exprs(gsvaResults$es.obs)[emtd,])
plot(exprs(gsvaResults$es.obs)[emtu,], exprs(gsvaResults$es.obs)[emtd,], col=copdca_colors[gsvaResults$es.obs$indicator])

# emtu with atf4 up, all and copd only
cor.test(exprs(gsvaResults$es.obs)[emtu,], exprs(gsvaResults$es.obs)[a_up,])
plot(exprs(gsvaResults$es.obs)[emtu,], exprs(gsvaResults$es.obs)[a_up,])
cor.test(exprs(gsvaResults$es.obs)[emtu,copd], exprs(gsvaResults$es.obs)[a_up,copd])
plot(exprs(gsvaResults$es.obs)[emtu,copd], exprs(gsvaResults$es.obs)[a_up,copd])

# emtd with atf4 up, all and copd only
cor.test(exprs(gsvaResults$es.obs)[emtd,], exprs(gsvaResults$es.obs)[a_up,])
plot(exprs(gsvaResults$es.obs)[emtd,], exprs(gsvaResults$es.obs)[a_up,])
cor.test(exprs(gsvaResults$es.obs)[emtd,copd], exprs(gsvaResults$es.obs)[a_up,copd])
plot(exprs(gsvaResults$es.obs)[emtd,copd], exprs(gsvaResults$es.obs)[a_up,copd])

# emtu with atf4 down, all and copd only
cor.test(exprs(gsvaResults$es.obs)[emtu,], exprs(gsvaResults$es.obs)[a_dn,])
cor.test(exprs(gsvaResults$es.obs)[emtu,copd], exprs(gsvaResults$es.obs)[a_dn,copd])

# emtd with atf4 down, all and copd only
cor.test(exprs(gsvaResults$es.obs)[emtd,], exprs(gsvaResults$es.obs)[a_dn,])
cor.test(exprs(gsvaResults$es.obs)[emtd,copd], exprs(gsvaResults$es.obs)[a_dn,copd])

cor.test(exprs(gsvaResults$es.obs)[emtd,], exprs(gsvaResults$es.obs)[lm_dn,])
cor.test(exprs(gsvaResults$es.obs)[emtd,], exprs(gsvaResults$es.obs)[lm_up,])



```
+ atf4 up genes correlate with emt genes, more so in COPD patients than all patients

### Background?  
+ We previously demonstrated a link between ATF4 and COPD (Steiling et al., 2013)  
+ ATF4/PERK/UPR linked to cancer  
+ EMT cells constitutively activate PERK/ATF4 pathway (Feng et al., 2014)  
+ Hypoxia induces EMT with increased PERK, ATF4/ATF6 (Shen et al., 2015)  
+ knockdown of ATF4/ATF4/PERK impedes EMT (Shen et al., 2015)  
+ NSCLC more likely in highly emphysematous regions (Bishawi et al., 2013)  

### Results
+ there is a signature for copd+cancer interaction enriched for ATF4 binding sites  