In this experiment I am going to check for separation between various groups based on the EMT gene signature from Byers et al., 2013. This signature includes both mesenchymal and epithelial cell markers, so combined should be able to distinguish between the cell types (of note: this signature was developed in tumor data)
  
```{r defineFunctions, include=FALSE}
foldChange <- function(eset, groups, genes){
  apply(exprs(eset)[genes, groups==0], 1, mean) - 
    apply(exprs(eset)[genes, groups==1], 1, mean)
}

medianFilter <- function(eset){
  md <- median(exprs(eset))
  passFilter <- logical(featureNumber(eset))
  for(i in 1:featureNumber(eset)){
    passFilter[i] <- sum(exprs(eset)[i, ] > md) > 0
  }
  
  eset <- eset[passFilter, ]
  return(eset)
  
}

saveGeneList <- function(analysis, filename=paste(Sys.Date(), "Temp", analysis$name, ".txt", sep="")){
  
  write.table(analysis$geneSymbols, file=filename, quote=FALSE,row.names=FALSE, col.names=FALSE)

}

saveRankedList <- function(fit, varColInd, filename = "tempRankedList"){
  filename = paste(paste(Sys.Date(), colnames(fit$t)[varColInd], filename, sep="_"), "rnk", sep=".")
  print(filename)
  temp <- data.frame(geneSymbols = fit$genes$Symbol, ts = fit$t[, varColInd])
  write.table(temp, file=filename, row.names=FALSE, quote=FALSE, col.names=TRUE, sep="\t", eol="\n")
  
  
}

saveGeneSet <- function(fit, inds, filename = "tempGeneSet", geneSetName="GeneSet", description=""){
  filename = paste(paste(Sys.Date(), filename, sep="_"), "gmx", sep=".")
  print(filename)
  temp <- data.frame(geneSymbols=c(geneSetName, description, fit$genes$Symbol[inds]))
  write.table(temp, file=filename, row.names=FALSE, quote=FALSE, col.names=FALSE)
  
}


```

### Setup
```{r setup, eval=TRUE, echo=TRUE, include=FALSE, results='hide'}

setwd("/protected/projects/pulmarray/Allegro/COPD_Cancer/experiments/Expmt_0040_EMT_ATF4_Clustering/")
source("/protected/projects/pulmarray/Allegro/COPD_Cancer/scripts/AllegroSetup.R")
#source("../2015-05-03_CBM_ATS_2015/plotGSEA.R")
# fix the one patient with wonky data
# eventually this should just be saved in the RDS file
holdEset <- eset
holdEset$FEV1Pc[holdEset$FEV1Pc==89.2] <- 0.892

emt_genes <- read.csv("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/emt_gene_symbols_byers_2013.csv", header=FALSE)
temp <- as.factor(unique(as.character(emt_genes$V1)))
emt_genes <- data.frame("EMT_Genes" = temp)

GOLD_colors = c("0" = "gray100", "1" = "gray75", "2" = "gray50", "3" = "gray25", "4" = "black")
greyblack <- colorRampPalette(c("black", "white", "grey"))(256)


# load ATF4 BEASB2 gene perturbation data for analysis
atf4_data <- readRDS("/protected/projects/pulmarray/Allegro/ATF4/ATF4data_2015MAR17_ExpressionSet.rds")

lam_sign <- read.delim("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/katie98genesEntrez.txt")
lam_sign[, 3] <- paste(lam_sign[, 2], "_at", sep="")

lamGMT <- read.csv("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/100525_lam_98overlapgenes_covars.gmt", sep="\t", as.is=TRUE, header=FALSE)
lamUP <- as.character(lamGMT[1, 3:56])
lamDN <- as.character(lamGMT[2, 3:46])

joshLM127dn <- read.csv("/protected/projects/pulmarray/HoggCOPD/A1ATGrifols/experiments/Expmt_0001_LM_Parenchyma/2015-06-08_Joshdn127.gmx", skip=1, header=TRUE)

joshLM127up <- read.csv("/protected/projects/pulmarray/HoggCOPD/A1ATGrifols/experiments/Expmt_0001_LM_Parenchyma/2015-06-08_Joshup127.gmx", skip=1, header=TRUE)

# from parenchymal REAL data
cleLMup <- read.csv("/protected/projects/pulmarray/HoggCOPD/A1ATGrifols/experiments/Expmt_0003_LM_Parenchyma_CLE/2015-06-24_parenchyma_cle_geneset_cluster2_fdr_0.05.gmx")

cleLMdn <- read.csv("/protected/projects/pulmarray/HoggCOPD/A1ATGrifols/experiments/Expmt_0003_LM_Parenchyma_CLE/2015-06-24_parenchyma_cle_geneset_cluster1_fdr_0.05.gmx")

natMed <- read.csv("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/Nature_Medicine_gmx_files/top_100_genes_symbols_quality_adjusted.gmx", sep="\t", as.is=TRUE)
natMedUp <- natMed[2:65, 1]
natMedDn <- natMed[2:37, 2]

tgfB <- read.csv("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/msigdb_C2_CGP_genesets/BIOCARTA_TGFB_PATHWAY.gmx", skip=1)
tgfB <- as.character(tgfB$X..TGF.beta.signaling.pathway)



#atf4_targets <- read.delim("ATF4_Targets_Enrichr_TRANSFACJASPAR_270Genes.txt", sep=",", as.is=TRUE, header=FALSE)
#atf4_targets <- as.character(atf4_targets)

#a4LEup <- read.delim("ATF4_targets_cluster1_270Genes_LeadingEdge_Up.csv", header=FALSE)
#a4LEup <- as.character(a4LEup$V1)

#a4LEdn <- read.delim("ATF4_targets_cluster2_270Genes_LeadingEdge_Down.csv", header=FALSE)
#a4LEdn <- as.character(a4LEdn$V1)
library(znorm)
library(diptest)
library(GSVA)
library(annotate)
library(org.Hs.eg.db)
library(corrplot)
library(ggplot2)
library(beeswarm)


```

### Data Cleaning
```{r cleanup}

# Remaining question - should we keep or remove the COPD GOLD Status 1 patients? i.e. mild disease
# What about patients using or their charts indicating they have used corticosteroids?
esetClean <- holdEset
esetClean <- cleanNAForAnalysis(esetClean, "COPD2_R7")
esetClean <- cleanNAForAnalysis(esetClean, "RATIOc")
esetClean <- cleanNAForAnalysis(esetClean, "PYc")
esetClean <- cleanNAForAnalysis(esetClean, "RIN")
esetClean <- removeFactorLevel(esetClean, "COPD2_R7", "DK")
esetClean <- removeFactorLevel(esetClean, "FinalCaDXc", "DK")
esetClean <- removeFactorLevel(esetClean, "COPD2_R7", "DK")
esetClean <- removeFactorLevel(esetClean, "GENDERc", "DK")

esetClean <- calcIndicator(esetClean, "FinalCaDXc", "COPD2_R7")
esetClean$smkindic <- as.numeric(as.character(esetClean$indicator))
esetClean$smkindic[esetClean$SMKc==2] <- esetClean$smkindic[esetClean$SMKc==2] + 4
esetClean$smkindic <- as.factor(esetClean$smkindic)

# Define the GOLD Stage appropriate for all patients
esetClean$GOLD <- factor(x=rep(0, sampleNumber(esetClean)), levels=c("0","1","2","3","4"))
# GOLD Stage 1 Ratio < 0.7, FEV1 >= 8
esetClean$GOLD[esetClean$RATIOc < 0.7 & esetClean$FEV1Pc >= 0.8] <- 1
# GOLD Stage 2 Ratio < 0.7, 50 <= FEV1 < 80
esetClean$GOLD[esetClean$RATIOc < 0.7 & esetClean$FEV1Pc < 0.8] <- 2
# GOLD Stage 3 Ratio < 0.7, 30 <= FEV1 <= 50
esetClean$GOLD[esetClean$RATIOc < 0.7 & esetClean$FEV1Pc < 0.5] <- 3
# GOLD Stage 4 Ratio < 0.7, FEV1 < 30%
esetClean$GOLD[esetClean$RATIOc < 0.7 & esetClean$FEV1Pc < 0.3] <- 4

esetCleanFormers <- removeFactorLevel(esetClean, "SMKc", "1")
esetCleanCurrents <- removeFactorLevel(esetClean, "SMKc", "2")
# Generate a super clean dataset devoid of all patients with other pulmonary diseases
esetSuperClean <- esetClean
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroPNAc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroTBc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroSarcoidc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroIPFc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroIPFc", "DK")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroAtelectasisc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroAtelectasisc", "DK")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroAsthmac", "DK")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroAsthmac", "1")

esetSuperCleanICS <- removeFactorLevel(esetSuperClean, "AllegroCorticosteroidsc", "1")
esetSuperCleanICS <- removeFactorLevel(esetSuperCleanICS, "AllegroCorticosteroidsc", "DK")

# clean up ATF4 data
varLabels(atf4_data) <- "SampleID"
atf4_data$SampleID[grep("ATF4", atf4_data$SampleID)] <- "ATF4"
atf4_data$SampleID[grep("Ctrl", atf4_data$SampleID)] <- "Control"
atf4_data$SampleID <- as.factor(atf4_data$SampleID)

atf4_colors <- c("ATF4" = "red", "Control" = "black")


```

### ATF4 Models
```{r ATF4models}
# generate ATF4 model(s)
atf4Model <- lmFitWrapper(atf4_data, c("SampleID"),
                          name="ATF4 Perturbation Data", adjust.method="none",
                          p.value=0.05, varOfInterest=1)

clabels <- cbind(atf4_colors[as.character(atf4_data$SampleID)])
heatmap3(exprs(atf4_data)[atf4Model$inds, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = atf4_data$SampleID, ColSideColors = clabels,
         main = "ATF4 Genes",
         labCol="")

hmapClusts <- heatmap3(exprs(atf4_data)[atf4Model$inds,], col = bluered,
                       keep.dendro=TRUE,
                       hclustfun=function(d) hclust(d, method="ward.D"),
                       col.clustering = atf4_data$SampleID, ColSideColors=clabels,
                       main="ATF4 Genes", labCol="")

clusters <- cutree(as.hclust(hmapClusts$Rowv), k=2)
atf4Cluster1_dn <- atf4Model$geneSymbols[clusters==1]
atf4Cluster2_up <- atf4Model$geneSymbols[clusters==2]

atf4dn <- names(clusters)[clusters==1]
atf4up <- names(clusters)[clusters==2]
atf4 <- names(clusters)

rlabels <- cbind(smoking_colors[clusters])
heatmap3(exprs(atf4_data)[atf4Model$inds,], col = bluered,
         keep.dendro=TRUE,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = atf4_data$SampleID, ColSideColors=clabels,
         main="ATF4 Genes", labCol="", RowSideColors=rlabels)



```


### Multiplot function definition 
```{r multiplot}
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)
  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)
  numPlots = length(plots)
  
  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                     ncol = cols, nrow = ceiling(numPlots/cols))
    }
  
  if (numPlots==1) {
    print(plots[[1]])
    } else {
      # Set up the page
      grid.newpage()
      pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
      
      # Make each plot, in the correct location
      for (i in 1:numPlots) {
        # Get the i,j matrix positions of the regions that contain this subplot
        matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
        print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                        layout.pos.col = matchidx$col))
        }
      }
  }

```

### EMT separation between current and former healthy patients
```{r emtNeitherSeparation}
eset <- esetClean
# remove the cancer patients
eset <- removeFactorLevel(eset, "FinalCaDXc", "1")
# remove the COPD patients
eset <- removeFactorLevel(eset, "COPD2_R7", "1")

# set the gene symbols as the row names
gnsbs <- getSYMBOL(return_entrez(featureNames(eset)), data='org.Hs.eg')
#eset2 <- eset[!is.na(temp), ]
#featureNames(eset2) <- getSYMBOL(return_entrez(featureNames(eset2)), data='org.Hs.eg')

# fix the emt gene signature so it matches the names pulled into gnsbs
# this was done all online via ncbi - below are the necessary 5 changes
emtGns <- as.character(emt_genes$EMT_Genes)
emtGns[match("MTAC2D1",emtGns)] <- "TC2N"
emtGns[match("C3orf21",emtGns)] <- "XXYLT1"
emtGns[match("LRRC54",emtGns)] <- "TSKU"
emtGns[match("RBM35A",emtGns)] <- "ESRP1"
emtGns[match("TACSTD1",emtGns)] <- "EPCAM"

inds <- match(emtGns, gnsbs)
# PCA plot generation code from Adam's QC Pipeline
emtPCAsmoking2 <- prcomp(na.omit(t(scale(t(exprs(eset)[inds,])))), center=FALSE, scale=FALSE);
percent.variance <- summary(emtPCAsmoking2)$importance["Proportion of Variance", ] * 100;
# Create a PDF of a plot of PC2 vs PC1
#pdf(output.filename);
smk_color = c("1" = "black", "2" = "grey")
pcCols <- smk_color[as.character(eset$SMKc)]
# Note: 10% extra space is alloted in order to give more room for the sample names
plot(emtPCAsmoking2$rotation[,c("PC1","PC2")], pch=21, cex=1, col=pcCols, bg=pcCols,
     xlim=1.1 * range(emtPCAsmoking2$rotation[,"PC1"]),
     ylim=1.1 * range(emtPCAsmoking2$rotation[,"PC2"]),
     xlab=sprintf("PC1 (%2.0f%%)", percent.variance["PC1"]),
     ylab=sprintf("PC2 (%2.0f%%)",percent.variance["PC2"]),
     main="Principal Component Analysis (PCA)\nacross EMT genes in healthy samples"
  )
legend("topright", c("Current", "Former"), col=c("black", "grey"), pch=21, pt.bg=c("black", "grey"), title="Smoking Status", inset=0.01)

# generate a heatmap of the EMT genes, unsupervised, hierarchical clustering
clabels <- cbind(smk_color[as.character(eset$SMKc)])
heatmap3(exprs(eset)[inds, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D2"),
         col.clustering = "unsupervised", ColSideColors = clabels,
         main = "EMT Genes in Non-Diseased Patients",
         labCol="", labRow=emtGns)

# check for differential expression among the EMT genes by Smoking status
smkModel <- lmFitWrapper(eset[inds,], c("SMKc"), 1, "bonferroni", 0.05, "EMT Genes Different by Smoking Status")
heatmap3(exprs(eset)[smkModel$inds, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = "semisupervised", ColSideColors = clabels,
         main = "DE EMT Genes in Non-Diseased Patients",
         labCol="", labRow=smkModel$geneSymbols)


dd <- cbind(t(exprs(eset)[smkModel$inds,]), pData(eset))
names(dd)[1:length(smkModel$inds)] <- smkModel$geneSymbols
p1 <- ggplot(dd, aes(SMKc, GALNT5)) + geom_boxplot(aes(fill=SMKc)) + geom_jitter()
p2 <- ggplot(dd, aes(SMKc, CARD6)) +  geom_boxplot(aes(fill=SMKc)) + geom_jitter()
p3 <- ggplot(dd, aes(SMKc, TMEM45B)) +  geom_boxplot(aes(fill=SMKc)) + geom_jitter()
p4 <- ggplot(dd, aes(SMKc, SERINC2)) +  geom_boxplot(aes(fill=SMKc)) + geom_jitter()
p5 <- ggplot(dd, aes(SMKc, MAPK13)) +  geom_boxplot(aes(fill=SMKc)) + geom_jitter()
p6 <- ggplot(dd, aes(SMKc, GALNT3)) +  geom_boxplot(aes(fill=SMKc)) + geom_jitter()
p7 <- ggplot(dd, aes(SMKc, SPINT2)) +  geom_boxplot(aes(fill=SMKc)) + geom_jitter()
p8 <- ggplot(dd, aes(SMKc, TACSTD2)) +  geom_boxplot(aes(fill=SMKc)) + geom_jitter()
p9 <- ggplot(dd, aes(SMKc, PRSS8)) +  geom_boxplot(aes(fill=SMKc)) + geom_jitter()

multiplot(p1, p2, p3, p4, p5, p6, p7, p8, p9, cols=3)

#SPINT2 acts as a tumor suppressor - here it is decreased significantly in smokers
# see Epigenetic inactivation of SPINT2 is associated with tumor suppressive function in esophageal squamous cell carcinoma Yue et al., 2014
```

### EMT separation between former smokers with and without COPD
```{r emtCOPDSeparation}
# remove the biological replicates
eset <- esetCleanFormers
# remove the cancer patients
eset <- removeFactorLevel(eset, "FinalCaDXc", "1")

# use gnsbs, emtGns, and inds as above

# PCA plot generation code from Adam's QC Pipeline
emtPCACOPD <- prcomp(na.omit(t(scale(t(exprs(eset)[inds,])))), center=FALSE, scale=FALSE);
percent.variance <- summary(emtPCACOPD)$importance["Proportion of Variance", ] * 100;
# Create a plot of PC2 vs PC1
pcCols <- copd_colors[as.character(eset$COPD2_R7)]
# Note: 10% extra space is alloted in order to give more room for the sample names
plot(emtPCACOPD$rotation[,c("PC1","PC2")], pch=21, cex=1, col=pcCols, bg=pcCols,
     xlim=1.1 * range(emtPCAsmoking2$rotation[,"PC1"]),
     ylim=1.1 * range(emtPCAsmoking2$rotation[,"PC2"]),
     xlab=sprintf("PC1 (%2.0f%%)", percent.variance["PC1"]),
     ylab=sprintf("PC2 (%2.0f%%)",percent.variance["PC2"]),
     main="Principal Component Analysis (PCA)\nacross EMT genes in former smokers with or without COPD"
  )
legend("topright", c("no COPD", "COPD"), col=c("bisque", "blue"), pch=21, pt.bg=c("bisque", "blue"), title="COPD Status", inset=0.01)

# generate a heatmap of the EMT genes, unsupervised, hierarchical clustering
clabels <- cbind(copd_colors[as.character(eset$COPD2_R7)])
heatmap3(exprs(eset)[inds, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D2"),
         col.clustering = "unsupervised", ColSideColors = clabels,
         main = "EMT Genes in Former Smokers with or without COPD",
         labCol="", labRow=emtGns)

# check for differential expression among the EMT genes by Smoking status
copdModel <- lmFitWrapper(eset[inds,], c("COPD2_R7"), 1, "bonferroni", 0.05, "EMT Genes Different by COPD Status among Former Smokers")
heatmap3(exprs(eset)[copdModel$inds, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = "semisupervised", ColSideColors = clabels,
         main = "DE EMT Genes in former smokers with or without COPD",
         labCol="", labRow=copdModel$geneSymbols)


dd <- cbind(t(exprs(eset)[copdModel$inds,]), pData(eset))
names(dd)[1:length(smkModel$inds)] <- copdModel$geneSymbols
p1 <- ggplot(dd, aes(COPD2_R7, ANTXR2)) + geom_boxplot(aes(fill=COPD2_R7)) + geom_jitter()
p2 <- ggplot(dd, aes(COPD2_R7, HNMT)) + geom_boxplot(aes(fill=COPD2_R7)) + geom_jitter()
p3 <- ggplot(dd, aes(COPD2_R7, PRSS22)) + geom_boxplot(aes(fill=COPD2_R7)) + geom_jitter()
p4 <- ggplot(dd, aes(COPD2_R7, SH3YL1)) + geom_boxplot(aes(fill=COPD2_R7)) + geom_jitter()
p5 <- ggplot(dd, aes(COPD2_R7, SERINC2)) + geom_boxplot(aes(fill=COPD2_R7)) + geom_jitter()
p6 <- ggplot(dd, aes(COPD2_R7, GRHL1)) + geom_boxplot(aes(fill=COPD2_R7)) + geom_jitter()
p7 <- ggplot(dd, aes(COPD2_R7, C1orf172)) + geom_boxplot(aes(fill=COPD2_R7)) + geom_jitter()
p8 <- ggplot(dd, aes(COPD2_R7, MPZL2)) + geom_boxplot(aes(fill=COPD2_R7)) + geom_jitter()
p9 <- ggplot(dd, aes(COPD2_R7, INADL)) + geom_boxplot(aes(fill=COPD2_R7)) + geom_jitter()

multiplot(p1, p2, p3)
multiplot(p4, p5, p6)
multiplot(p7, p8, p9)

# upregulated in COPD
# 64063=PRSS22, protease, serine, 22: participates in ECM remodeling and overexpression enhanced tumor cell migration and invasion (Chen et al. 2014 Thyroid hormone enhanced human hepatoma cell motility involved PRSS22 activation via ERK signaling); additionally mild association between expression and metastases was observed in the same paper

# 347735=SERINC2/TDE2 is highly expressed in NSCLC Player et al., 2003 (Identification of TDE2 gene and its expression in NSCLC)

# 126695=KDF1/C1orf172: blocks proliferation and promotes differentiation in epidermal progenitor cells (Lee, Kong, and Weatherbee 2013 Forward genetics identifies...)

# downregulated in COPD
# ANTXR2 anthrax toxin receptor 2: binds collagen IV and laminin, suggesting it may be involved in ECM adhesion. aka capillary morphogenesis gene 2 (CMG2), overexpression enhanced adhesion of endothelial cells to ECM but was negatively associated with cell migration (more gene, less migration; more gene more adhesion - suggestive of role in EMT; Ye et al., Therapeutic potential of CMG2 in tumour-related angiogenesis)... overexpression also inhibited angiogenesis
# Ye et al. 2014 CMG2 inhibits growth of breast cancer cells and is inversely correlated with the disease progression and prognosis

# HNMT histamine N-methyltransferase: nothing sticks out as relevant here

# SH3YL1: Dock4 forms a complex with SH3YL1 and regulates cancer cell migration (opposite of expected result here)

# GRHL1 grainyhead-like 1 acts like a TF during development: in combination with Nkx2.1 forms a positive transcriptional loop to coordinate morphogenesis and differentiation of lung epithelial cells (Varma et al., 2012). varma et al., 2014 "Grhl2...maintains alveolar epithelial cell integrity by directly regulating components of adherens and tight junctions" (NOTE - these refer to Grhl2 not Grhl1); Grhl1 acts as a tumor suppressor in neuroblastoma and is negatively regulated by MYCN and HDAC3 (Fabian et al., 2014); cadherin family genes are direct targets of the grainyhead-like TFs (Wilanowski et al., 2008 - Perturbed desmosomal cadherin expression in Grhl1 null mice)

# MPZL2 myelin protein zero like 2: mediates cell-cell adhesion

# INADL: InaD-like, aka PATJ, localizes to tight junctions and to the apical membrane of epithelial cells; PATJ regulates directional migration of mammalian epithelial cells (Shin, Wang, Margolis 2007)
```

### EMT separation between former smokers with COPD +/- Cancer
```{r emtCOPDCancerSeparation}
eset <- esetCleanFormers
# remove the cancer patients
eset <- removeFactorLevel(eset, "COPD2_R7", "0")

# use gnsbs, emtGns, and inds as above

# PCA plot generation code from Adam's QC Pipeline
emtPCACOPD <- prcomp(na.omit(t(scale(t(exprs(eset)[inds,])))), center=FALSE, scale=FALSE);
percent.variance <- summary(emtPCACOPD)$importance["Proportion of Variance", ] * 100;
# Create a plot of PC2 vs PC1
pcCols <- copdca_colors[as.character(eset$indicator)]
# Note: 10% extra space is alloted in order to give more room for the sample names
plot(emtPCACOPD$rotation[,c("PC1","PC2")], pch=21, cex=1, col=pcCols, bg=pcCols,
     xlim=1.1 * range(emtPCAsmoking2$rotation[,"PC1"]),
     ylim=1.1 * range(emtPCAsmoking2$rotation[,"PC2"]),
     xlab=sprintf("PC1 (%2.0f%%)", percent.variance["PC1"]),
     ylab=sprintf("PC2 (%2.0f%%)",percent.variance["PC2"]),
     main="Principal Component Analysis (PCA)\nacross EMT genes in former smokers with COPD +/- Cancer"
  )
legend("topright", c("COPD", "COPD+Cancer"), col=c("green", "blue"), pch=21, pt.bg=c("green", "blue"), title="COPD Status", inset=0.01)

# generate a heatmap of the EMT genes, unsupervised, hierarchical clustering
clabels <- cbind(copdca_colors[as.character(eset$indicator)])
heatmap3(exprs(eset)[inds, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D2"),
         col.clustering = "unsupervised", ColSideColors = clabels,
         main = "EMT Genes in Former Smokers with COPD +/- Cancer",
         labCol="", labRow=emtGns)

# check for differential expression among the EMT genes by Smoking status
copdModel <- lmFitWrapper(eset[inds,], c("FinalCaDXc"), 1, "fdr", 0.05, "EMT Genes Different by Cancer Status among Former Smokers with COPD")
heatmap3(exprs(eset)[copdModel$inds, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = "semisupervised", ColSideColors = clabels,
         main = "DE EMT Genes in former smokers with or without COPD",
         labCol="", labRow=copdModel$geneSymbols)


dd <- cbind(t(exprs(eset)[copdModel$inds,]), pData(eset))
names(dd)[1:length(copdModel$inds)] <- copdModel$geneSymbols
p1 <- ggplot(dd, aes(COPD2_R7, MAL2)) + geom_boxplot(aes(fill=FinalCaDXc)) + geom_jitter()
p2 <- ggplot(dd, aes(COPD2_R7, MPZL2)) + geom_boxplot(aes(fill=FinalCaDXc)) + geom_jitter()
p3 <- ggplot(dd, aes(COPD2_R7, EPB41L5)) + geom_boxplot(aes(fill=FinalCaDXc)) + geom_jitter()

multiplot(p1, p2, p3)

# 3 genes up-regulated in COPD+Cancer v COPD
# MAL2
# MPZL2
# EPB41L5 - erythrocyte membrane protein band 4.1 like 5; overexpression in polarized MDCK cells affects tightness of cell junctions and results in disorganization of the tight junction markers ZO-1 and PATJ (above)
```


### EMT separation between COPD and non-COPD among all smokers (no cancer)
```{r emtCOPDCancerSeparation2}
eset <- removeBioReps(esetClean)

# use gnsbs, emtGns, and inds as above

# PCA plot generation code from Adam's QC Pipeline
emtPCACOPD <- prcomp(na.omit(t(scale(t(exprs(eset)[inds,])))), center=FALSE, scale=FALSE);

# remove the cancer patients
#eset <- removeFactorLevel(eset, "FinalCaDXc", "1")

percent.variance <- summary(emtPCACOPD)$importance["Proportion of Variance", ] * 100;
# Create a plot of PC2 vs PC1
pcCols <- copdca_colors[as.character(eset$indicator)]
# Note: 10% extra space is alloted in order to give more room for the sample names
plot(emtPCACOPD$rotation[eset$FinalCaDXc==0,c("PC1","PC2")], pch=21, cex=1, col=pcCols[eset$FinalCaDXc==0], bg=pcCols[eset$FinalCaDXc==0],
     xlim=1.1 * range(emtPCACOPD$rotation[,"PC1"]),
     ylim=1.1 * range(emtPCACOPD$rotation[,"PC2"]),
     xlab=sprintf("PC1 (%2.0f%%)", percent.variance["PC1"]),
     ylab=sprintf("PC2 (%2.0f%%)",percent.variance["PC2"]),
     main="Principal Component Analysis (PCA)\nacross EMT genes in samples without cancer +/- COPD"
  )
legend("topright", c("no COPD", "COPD"), col=c("red", "green"), pch=21, pt.bg=c("red", "green"), title="COPD Status", inset=0.01)

# generate a heatmap of the EMT genes, unsupervised, hierarchical clustering
clabels <- cbind(copdca_colors[as.character(eset$indicator)])
heatmap3(exprs(eset)[inds, eset$FinalCaDXc==0], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D2"),
         col.clustering = "unsupervised", ColSideColors = clabels[eset$FinalCaDXc==0],
         main = "EMT Genes in All Smokers +/- COPD",
         labCol="", labRow=emtGns)

# check for differential expression among the EMT genes by Smoking status
copdModel <- lmFitWrapper(eset[inds,eset$FinalCaDXc==0], c("COPD2_R7", "AllegroCorticosteroidsc"), 1, "bonferroni", 0.05, "EMT Genes Different by COPD Status among non-Cancer patients")
heatmap3(exprs(eset)[copdModel$inds, eset$FinalCaDXc==0], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = "semisupervised", ColSideColors = clabels[eset$FinalCaDXc==0],
         main = "DE EMT Genes in all smokers with or without COPD",
         labCol="", labRow=copdModel$geneSymbols)


dd <- cbind(t(exprs(eset)[copdModel$inds,eset$FinalCaDXc==0]), pData(eset[, eset$FinalCaDXc==0]))
names(dd)[1:length(copdModel$inds)] <- copdModel$geneSymbols
#plots <- list()
#for(i in 1:length(copdModel$inds)){
  
#  plots[[i]] <- ggplot(dd, aes(COPD2_R7, [copdModel$geneSymbols[i]])) + geom_boxplot(aes(fill=COPD2_R7)) + geom_jitter() + labs(title=copdModel$geneSymbols[i], y=copdModel$geneSymbols[i])
  
#}


p1 <- ggplot(dd, aes(COPD2_R7, GALNT5)) + geom_boxplot(aes(fill=FinalCaDXc)) + geom_jitter()
p2 <- ggplot(dd, aes(COPD2_R7, HNMT)) + geom_boxplot(aes(fill=FinalCaDXc)) + geom_jitter()
p3 <- ggplot(dd, aes(COPD2_R7, EPB41L5)) + geom_boxplot(aes(fill=FinalCaDXc)) + geom_jitter()


multiplot(plots[[1]], plots[[2]], plots[[3]], plots[[4]], cols=2)

```