Allegro COPD*Cancer: Changing ratio cutoffs  
========================================================
2014-06-09
-------------------------------------------
I will be running an analysis for the Allegro CODP+Cancer project. 
In this analysis I will be using only patients with PFTs and defining COPD by lowering the ratio threshold from < 0.7 to less than 0.6, 0.5, etc. Alternatively, I could include FEV1 or the ratio as a continuous measure.

Thoughts for other analyses:  
  * split out into cancer groups - specifically look at squamous cell subtype
  * use SVA  
  * use forward selection for covariates
  * include cancer subtype  
  
  
To do:  
  * recode functions to work with apply (e.g. return_lm)  


```{r Running AllegroSetup.R}
# AllegroSetup.R uses an RDS file with 800+ samples (biological replicates) and without QC
source("/protected/projects/pulmarray/Allegro/COPD_Cancer/scripts/AllegroSetup.R")
```

### Analysis in R starts here:
```{r Preparing data for analysis}

# find the samples with PFT information
pftInds <- which(!is.na(exprData$RATIOc))
# subset the dataset to include only those patients with PFTs
exprDataPFT <- exprData[,pftInds]

# calculate a binary COPD field that strictly follows the Ratio < 0.7 = COPD
COPD_strict <- numeric(sampleNumber(exprDataPFT))
COPD_strict[which(exprDataPFT$RATIOc<0.7)] <- 1
COPD_strict[which(exprDataPFT$RATIOc==0.7)] <- 2
# add the field to the expressionSet
exprDataPFT$COPD_strict <- COPD_strict
# remove the subjects with R=0.7
exprDataPFT <- exprDataPFT[,exprDataPFT$RATIO!=0.7]
# make COPD_strict a factor
exprDataPFT$COPD_strict <- as.factor(exprDataPFT$COPD_strict)

# calculate a COPD severity field
# Ratio: 0.6-0.7 = mild, 0.5-0.6 = moderate, 0.4-0.5 = severe, < 0.4 = very severe
COPD_severity <- numeric(sampleNumber(exprDataPFT))
COPD_severity[which(exprDataPFT$RATIOc<0.7)] <- 1
COPD_severity[which(exprDataPFT$RATIOc<0.6)] <- 2
COPD_severity[which(exprDataPFT$RATIOc<0.5)] <- 3
COPD_severity[which(exprDataPFT$RATIOc<0.4)] <- 4
# add the field to the expressionSet
# make sure the factor is ordered
exprDataPFT$COPD_severity <- factor(COPD_severity, ordered=TRUE)

# calculate/create an indicator column
exprDataPFT <- calcIndicator(exprDataPFT, "COPD_strict")

# remove the DKs from COPD2_R7
exprDataPFT <- removeFactorLevel(exprDataPFT, "FinalCaDXc", "DK")
```



```{r Analyzing data}
# define covariate lists to use for analysis
covarsR7 <- list("SMKc", "COPD_strict", "FinalCaDXc", "COPD_strict*FinalCaDXc", "SMKc*COPD_strict", "SMKc*FinalCaDXc")
testPFT1 <- analyze_covariate_lists(exprDataPFT, covarsR7)
summary(testPFT1$cutoffs_fdr[["0.25"]])
summary(testPFT1$cutoffs_p[["0.05"]])

# testPFTs has no significant by FDR genes for COPD_strict*Cancer but many for p
# will try filtering data down to more sever subset and see what happens

# remove patients with Ratio < 0.7 & >= 0.6
exprDataPFTR2 <- exprDataPFT[,exprDataPFT$COPD_severity!=1]
testPFT2 <- analyze_covariate_lists(exprDataPFTR2, covarsR7)
summary(testPFT2$cutoffs_fdr[["0.25"]])
summary(testPFT2$cutoffs_p[["0.05"]])

# remove patients with Ratio < 0.6 & >= 0.5
exprDataPFTR3 <- exprDataPFTR2[, exprDataPFTR2$COPD_severity!=2]
exprDataPFTR3 <- filtering(exprDataPFTR3)
testPFT3 <- analyze_covariate_lists(exprDataPFTR3$unfiltered, covarsR7)
summary(testPFT3$cutoffs_fdr[["0.25"]])
summary(testPFT3$cutoffs_p[["0.05"]])

testPFT4 <- analyze_covariate_lists(exprDataPFTR3$mn.cv, covarsR7)
summary(testPFT4$cutoffs_fdr[["0.25"]])
summary(testPFT4$cutoffs_p[["0.05"]])

# Try a different, simpler modeling approach
covarsSimple <- list("SMKc", "COPD_strict", "FinalCaDXc", "COPD_strict*FinalCaDXc")

exprDataPFT <- filtering(exprDataPFT)
testPFT5 <- analyze_covariate_lists(exprDataPFT$cv, covarsSimple)
summary(testPFT5$cutoffs_fdr[["0.25"]])
summary(testPFT5$cutoffs_p[["0.05"]])

# Try checking only those patients with squamous cell subtype cancer or no cancer
# In doing this, many values of CA_SUB_CELL_TYPE are NA
# these NA values returned an NA value in the function "removeFactorLevel"
# which in turn set all of the values in the pData subsequently to NA, which is wrong
exprDataSqPFT <- exprDataPFT$unfiltered
exprDataSqPFT$CA_SUB_CELL_TYPE <- factor(exprDataSqPFT$CA_SUB_CELL_TYPE, levels=c(levels(exprDataSqPFT$CA_SUB_CELL_TYPE), 0))
exprDataSqPFT$CA_SUB_CELL_TYPE[which(is.na(exprDataSqPFT$CA_SUB_CELL_TYPE))] <- 0

exprDataSqPFT <- exprDataSqPFT[,which(exprDataSqPFT$CA_SUB_CELL_TYPE=="Squamous" | exprDataSqPFT$FinalCaDXc==0)]


# force CA_SUB_CELL_TYPE==NA to 0 (i.e. no cancer)
# first add '0' to list of levels of factor CA_SUB_CELL_TYPE
exprDataSqPFT <- removeFactorLevel(exprDataSqPFT, "CA_SUB_CELL_TYPE", "Adeno")
exprDataSqPFT <- removeFactorLevel(exprDataSqPFT, "CA_SUB_CELL_TYPE", "Large Cell")
exprDataSqPFT <- removeFactorLevel(exprDataSqPFT, "CA_SUB_CELL_TYPE", "#N/A")
exprDataSqPFT <- removeFactorLevel(exprDataSqPFT, "CA_SUB_CELL_TYPE", "NSC UN")
exprDataSqPFT <- removeFactorLevel(exprDataSqPFT, "CA_SUB_CELL_TYPE", "Uncertain")

# filter the data
exprDataSqPFT <- filtering(exprDataSqPFT)

# run the analysis
testSqPFT <- analyze_covariate_lists(exprDataSqPFT$cv, covarsSimple)

# create an expression set of squamous cancer and COPD with R < 0.6
exprDataSqPFT2 <- removeFactorLevel(exprDataSqPFT$unfiltered,"COPD_severity", "1")
exprDataSqPFT2 <- filtering(exprDataSqPFT2)

# run analyses with different covariate lists
testSqPFT2 <- analyze_covariate_lists(exprDataSqPFT2$cv, covarsSimple)
testSqPFT3 <- analyze_covariate_lists(exprDataSqPFT2$mn.cv, covarsR7)


clabels <- cbind("Group"=copdca_colors[exprDataSqPFT2$mn.cv$indicator],
                 "Cancer"=cancer_colors[exprDataSqPFT2$mn.cv$FinalCaDXc],
                 "COPD"=copd_colors[exprDataSqPFT2$mn.cv$COPD2_R7])


```


```{r Creating Figures, fig.width=7, fig.height=6}
# find the indices to print in the heatmap
inds <- testSqPFT3$cutoffs_p[[2]]$'COPD_strict1:FinalCaDXc1'
  
heatmap3(exprs(exprDataSqPFT2$mn.cv)[inds,], col=bluered, ColSideColors=clabels, col.clustering=exprDataSqPFT2$mn.cv$indicator, main=paste(length(inds), "genes at p < 0.01 for COPD*Cancer interaction\nSquamous Cell and Ratio < 0.6"))

heatmap3(exprs(exprDataSqPFT2$mn.cv)[inds,], col=bluered, ColSideColors=clabels, main=paste(length(inds), "genes at p < 0.01 for COPD*Cancer interaction\nSquamous Cell and Ratio < 0.6"))



```

For Next Time:  
=====================

```{r Analyzing sample clusters and their significant genes}
#cl1 <- return_cluster(inds, exprDataSqPFT2$cv.mn)
#cl1 <- factor(cl1)
#exprDataSqPFT2$cv.mn$cluster <- cl1
#testSqPFT4 <- analyze_covariate_lists(exprDataSqPFT2, list("cl1"))


  
```
