Allegro COPD*Cancer: Changing ratio cutoffs  
========================================================
2014-06-12
-------------------------------------------
I will be running an analysis for the Allegro CODP+Cancer project. 
In this analysis I will be using lmFit in the R package 'limma' as it solves all of the  
linear models simultaneously through the use of model matrices

Thoughts for other analyses:  
  * split out into cancer groups - specifically look at squamous cell subtype
  * use SVA  
  * use forward selection for covariates
  * include cancer subtype  
  * split out into separate smoking groups  
  
  
  
To do:  
  * recode functions to work with apply (e.g. return_lm) and limma lmFit


```{r RunningAllegroSetup}
# AllegroSetup.R uses an RDS file with 800+ samples (biological replicates) and without QC
source("/protected/projects/pulmarray/Allegro/COPD_Cancer/scripts/AllegroSetup.R")
library(limma)
```

### Analysis in R starts here:
```{r PreparingData}

# find the samples with PFT information
pftInds <- which(!is.na(exprData$RATIOc))
# subset the dataset to include only those patients with PFTs
exprDataPFT <- exprData[,pftInds]

# calculate a binary COPD field that strictly follows the Ratio < 0.7 = COPD
COPD_strict <- numeric(sampleNumber(exprDataPFT))
COPD_strict[which(exprDataPFT$RATIOc<0.7)] <- 1
COPD_strict[which(exprDataPFT$RATIOc==0.7)] <- 2
# add the field to the expressionSet
exprDataPFT$COPD_strict <- COPD_strict
# remove the subjects with R=0.7
exprDataPFT <- exprDataPFT[,exprDataPFT$RATIO!=0.7]
# make COPD_strict a factor
exprDataPFT$COPD_strict <- as.factor(exprDataPFT$COPD_strict)

# calculate a COPD severity field
# Ratio: 0.6-0.7 = mild, 0.5-0.6 = moderate, 0.4-0.5 = severe, < 0.4 = very severe
COPD_severity <- numeric(sampleNumber(exprDataPFT))
COPD_severity[which(exprDataPFT$RATIOc<0.7)] <- 1
COPD_severity[which(exprDataPFT$RATIOc<0.6)] <- 2
COPD_severity[which(exprDataPFT$RATIOc<0.5)] <- 3
COPD_severity[which(exprDataPFT$RATIOc<0.4)] <- 4
# add the field to the expressionSet
# make sure the factor is ordered
exprDataPFT$COPD_severity <- factor(COPD_severity, ordered=TRUE)

# calculate/create an indicator column
exprDataPFT <- calcIndicator(exprDataPFT, "COPD_strict")

# remove the DKs from COPD2_R7
exprDataPFT <- removeFactorLevel(exprDataPFT, "FinalCaDXc", "DK")
```



```{r limmaBlock, fig.width=7, fig.height=6}
# define the data to be worked with
eset <- exprDataPFT
eset <- removeFactorLevel(eset, "COPD2_R7", "DK")
eset <- cleanNAForAnalysis(eset, "COPD2_R7")

# define a new COPD factor with appropriate labels
copd <- eset$COPD2_R7
copd <- factor(copd, levels=c(0,1,"COPD", "noCOPD"))
copd[copd==1] <- "COPD"
copd[copd==0] <- "noCOPD"
copd <- factor(copd, levels=c("COPD", "noCOPD"))

# define a new Cancer factor with appropriate labels
cancer <- eset$FinalCaDXc
cancer <- factor(cancer, levels=c(0,1,"Cancer", "noCancer"))
cancer[cancer==1] <- "Cancer"
cancer[cancer==0] <- "noCancer"
cancer <- factor(cancer, levels=c("Cancer","noCancer"))

# define a new COPD.Cancer factor with appropriate labels
CC <- paste(copd,cancer,sep=".")
CC <- factor(CC, levels=c("noCOPD.noCancer", "noCOPD.Cancer", "COPD.noCancer", "COPD.Cancer"))

# define the model matrix to solve
design <- model.matrix(~0+CC)

# set the names of the design matrix to simply be the factor levels
colnames(design) <- levels(CC)

# fit the model to the data
fit <- lmFit(eset, design)

# setup the contrasts of interest
# here we've setup all of the possible contrasts
cont.matrixF <- makeContrasts(HealthyVsCancer = noCOPD.noCancer - noCOPD.Cancer,
                             HealthyVsCOPD = noCOPD.noCancer - COPD.noCancer,
                             HealthyVsCC = noCOPD.noCancer - COPD.Cancer,
                             CancerVsCOPD = noCOPD.Cancer - COPD.noCancer,
                             CancerVsCC = noCOPD.Cancer - COPD.Cancer,
                             COPDvsCC = COPD.noCancer - COPD.Cancer, 
                             Diff1 = ((COPD.Cancer - COPD.noCancer) - 
                                       (noCOPD.Cancer - noCOPD.noCancer)),
                             levels=designF)

# calculate the contrasts
fitF2 <- contrasts.fit(fitF, cont.matrixF)
fitF2 <- eBayes(fitF2)

resultsF <- decideTests(fitF2)
#save_entrez(1:21, names(inds), filename="entrez.txt")
eset <- calcIndicator(eset,"FinalCaDXc", "COPD2_R7")

vennDiagram(results[,1:5])

inds <- which(results[,2]!=0 & results[,3]!=0 & results[,5]!=0 & results[,4]!=0)
generate_heatmap(inds, eset)

inds2 <- which(results[,2]!=0 & results[,3]!=0 & results[,5]!=0 & results[,4]!=0)
generate_heatmap(inds2, eset)

inds3 <- which(results[,2]!=0 & results[,3]!=0 & results[,5]!=0)
generate_heatmap(inds3, eset)
# smoking status is a problem here - need to correct for SMKc

# relabel SMKc and GENDERc to be correct names
# SMKc 1=current, 2=former
# GENDERc 0=male, 1=female
smk <- factor(eset$SMKc,levels=c(1,2,"C", "F"))
gdr <- factor(eset$GENDERc, levels=c(0,1,"M","F"))
smk[smk==1] <- "C"
smk[smk==2] <- "F"
gdr[gdr==0] <- "M"
gdr[gdr==1] <- "F"
smk <- factor(smk, levels=c("F","C"))
gdr <- factor(gdr, levels=c("F", "M"))
eset$SMKj<- smk
eset$GENDERj <- gdr


design2a <- model.matrix(~0+CC+smk)
colnames(design2a) <- c(levels(CC), "smkc")

contrast.matrix2 <- makeContrasts(HealthyVsCancer=noCOPD.noCancer-noCOPD.Cancer, HealthyVsCOPD=noCOPD.noCancer-COPD.noCancer, HealthyVsCC=noCOPD.noCancer-COPD.Cancer, CancerVsCOPD=noCOPD.Cancer-COPD.noCancer, CancerVsCC=noCOPD.Cancer-COPD.Cancer, COPDvsCC=COPD.noCancer-COPD.Cancer, levels=design2a)

fit2a <- lmFit(eset, design2a)

fit2b <- contrasts.fit(fit2a, contrast.matrix2)
fit2b <- eBayes(fit2b)

results2a <- decideTests(fit2b)
summary(results2a)
vennDiagram(results2a[,1:5])

inds1a <- which(results2a[,2]!=0 & results2a[,3]!=0 & results2a[,5]!=0 & results2a[,4]==0)
generate_heatmap(inds1a, eset)

inds2a <- which(results2a[,2]!=0 & results2a[,3]!=0 & results2a[,5]!=0 & results2a[,4]!=0)
generate_heatmap(inds2a, eset)

inds3a <- which(results2a[,2]!=0 & results2a[,3]!=0 & results2a[,5]!=0)
generate_heatmap(inds3a, eset)

```

```{r limmaBlock2, eval=FALSE}

esetHold <- eset
esetCurSM <- eset[,eset$SMKc==1]
esetForSM <- eset[,eset$SMKc==2]
eset <- esetCurSM

```


```{r CreatingFigures, fig.width=7, fig.height=6, eval=FALSE}
# find the indices to print in the heatmap
inds <- testSqPFT3$cutoffs_p[[2]]$'COPD_strict1:FinalCaDXc1'
  
heatmap3(exprs(exprDataSqPFT2$mn.cv)[inds,], col=bluered, ColSideColors=clabels, col.clustering=exprDataSqPFT2$mn.cv$indicator, main=paste(length(inds), "genes at p < 0.01 for COPD*Cancer interaction\nSquamous Cell and Ratio < 0.6"))

heatmap3(exprs(exprDataSqPFT2$mn.cv)[inds,], col=bluered, ColSideColors=clabels, main=paste(length(inds), "genes at p < 0.01 for COPD*Cancer interaction\nSquamous Cell and Ratio < 0.6"))



```