In this experiment I am going to check for separation between various groups based on the EMT gene signature from Byers et al., 2013. This signature includes both mesenchymal and epithelial cell markers, so combined should be able to distinguish between the cell types (of note: this signature was developed in tumor data)
  
```{r defineFunctions, include=FALSE}
foldChange <- function(eset, groups, genes){
  apply(exprs(eset)[genes, groups==0], 1, mean) - 
    apply(exprs(eset)[genes, groups==1], 1, mean)
}

medianFilter <- function(eset){
  md <- median(exprs(eset))
  passFilter <- logical(featureNumber(eset))
  for(i in 1:featureNumber(eset)){
    passFilter[i] <- sum(exprs(eset)[i, ] > md) > 0
  }
  
  eset <- eset[passFilter, ]
  return(eset)
  
}

saveGeneList <- function(analysis, filename=paste(Sys.Date(), "Temp", analysis$name, ".txt", sep="")){
  
  write.table(analysis$geneSymbols, file=filename, quote=FALSE,row.names=FALSE, col.names=FALSE)

}

saveRankedList <- function(fit, varColInd, filename = "tempRankedList"){
  filename = paste(paste(Sys.Date(), colnames(fit$t)[varColInd], filename, sep="_"), "rnk", sep=".")
  print(filename)
  temp <- data.frame(geneSymbols = fit$genes$Symbol, ts = fit$t[, varColInd])
  write.table(temp, file=filename, row.names=FALSE, quote=FALSE, col.names=TRUE, sep="\t", eol="\n")
  
  
}

saveGeneSet <- function(fit, inds, filename = "tempGeneSet", geneSetName="GeneSet", description=""){
  filename = paste(paste(Sys.Date(), filename, sep="_"), "gmx", sep=".")
  print(filename)
  temp <- data.frame(geneSymbols=c(geneSetName, description, fit$genes$Symbol[inds]))
  write.table(temp, file=filename, row.names=FALSE, quote=FALSE, col.names=FALSE)
  
}


```

### Setup
```{r setup, eval=TRUE, echo=TRUE, include=FALSE, results='hide'}

setwd("/protected/projects/pulmarray/Allegro/COPD_Cancer/experiments/Expmt_0042_ATF4_Confirmation/")
source("/protected/projects/pulmarray/Allegro/COPD_Cancer/scripts/AllegroSetup.R")
#source("../2015-05-03_CBM_ATS_2015/plotGSEA.R")
# fix the one patient with wonky data
# eventually this should just be saved in the RDS file
holdEset <- eset
holdEset$FEV1Pc[holdEset$FEV1Pc==89.2] <- 0.892

emt_genes <- read.csv("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/emt_gene_symbols_byers_2013.csv", header=FALSE)
temp <- as.factor(unique(as.character(emt_genes$V1)))
emt_genes <- data.frame("EMT_Genes" = temp)

GOLD_colors = c("0" = "gray100", "1" = "gray75", "2" = "gray50", "3" = "gray25", "4" = "black")
greyblack <- colorRampPalette(c("black", "white", "grey"))(256)


# load ATF4 BEASB2 gene perturbation data for analysis
atf4_data <- readRDS("/protected/projects/pulmarray/Allegro/ATF4/ATF4data_2015MAR17_ExpressionSet.rds")

lam_sign <- read.delim("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/katie98genesEntrez.txt")
lam_sign[, 3] <- paste(lam_sign[, 2], "_at", sep="")

lamGMT <- read.csv("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/100525_lam_98overlapgenes_covars.gmt", sep="\t", as.is=TRUE, header=FALSE)
lamUP <- as.character(lamGMT[1, 3:56])
lamDN <- as.character(lamGMT[2, 3:46])

joshLM127dn <- read.csv("/protected/projects/pulmarray/HoggCOPD/A1ATGrifols/experiments/Expmt_0001_LM_Parenchyma/2015-06-08_Joshdn127.gmx", skip=1, header=TRUE)

joshLM127up <- read.csv("/protected/projects/pulmarray/HoggCOPD/A1ATGrifols/experiments/Expmt_0001_LM_Parenchyma/2015-06-08_Joshup127.gmx", skip=1, header=TRUE)

# from parenchymal REAL data
cleLMup <- read.csv("/protected/projects/pulmarray/HoggCOPD/A1ATGrifols/experiments/Expmt_0003_LM_Parenchyma_CLE/2015-06-24_parenchyma_cle_geneset_cluster2_fdr_0.05.gmx")

cleLMdn <- read.csv("/protected/projects/pulmarray/HoggCOPD/A1ATGrifols/experiments/Expmt_0003_LM_Parenchyma_CLE/2015-06-24_parenchyma_cle_geneset_cluster1_fdr_0.05.gmx")

natMed <- read.csv("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/Nature_Medicine_gmx_files/top_100_genes_symbols_quality_adjusted.gmx", sep="\t", as.is=TRUE)
natMedUp <- natMed[2:65, 1]
natMedDn <- natMed[2:37, 2]

tgfB <- read.csv("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/msigdb_C2_CGP_genesets/BIOCARTA_TGFB_PATHWAY.gmx", skip=1)
tgfB <- as.character(tgfB$X..TGF.beta.signaling.pathway)



#atf4_targets <- read.delim("ATF4_Targets_Enrichr_TRANSFACJASPAR_270Genes.txt", sep=",", as.is=TRUE, header=FALSE)
#atf4_targets <- as.character(atf4_targets)

#a4LEup <- read.delim("ATF4_targets_cluster1_270Genes_LeadingEdge_Up.csv", header=FALSE)
#a4LEup <- as.character(a4LEup$V1)

#a4LEdn <- read.delim("ATF4_targets_cluster2_270Genes_LeadingEdge_Down.csv", header=FALSE)
#a4LEdn <- as.character(a4LEdn$V1)
library(znorm)
library(diptest)
library(GSVA)
library(annotate)
library(org.Hs.eg.db)
library(corrplot)
library(ggplot2)
library(beeswarm)


```

### Data Cleaning
```{r cleanup}

# Remaining question - should we keep or remove the COPD GOLD Status 1 patients? i.e. mild disease
# What about patients using or their charts indicating they have used corticosteroids?
esetClean <- holdEset
esetClean <- cleanNAForAnalysis(esetClean, "COPD2_R7")
esetClean <- cleanNAForAnalysis(esetClean, "RATIOc")
esetClean <- cleanNAForAnalysis(esetClean, "PYc")
esetClean <- cleanNAForAnalysis(esetClean, "RIN")
esetClean <- removeFactorLevel(esetClean, "COPD2_R7", "DK")
esetClean <- removeFactorLevel(esetClean, "FinalCaDXc", "DK")
esetClean <- removeFactorLevel(esetClean, "COPD2_R7", "DK")
esetClean <- removeFactorLevel(esetClean, "GENDERc", "DK")

esetClean <- calcIndicator(esetClean, "FinalCaDXc", "COPD2_R7")
esetClean$smkindic <- as.numeric(as.character(esetClean$indicator))
esetClean$smkindic[esetClean$SMKc==2] <- esetClean$smkindic[esetClean$SMKc==2] + 4
esetClean$smkindic <- as.factor(esetClean$smkindic)

# Define the GOLD Stage appropriate for all patients
esetClean$GOLD <- factor(x=rep(0, sampleNumber(esetClean)), levels=c("0","1","2","3","4"))
# GOLD Stage 1 Ratio < 0.7, FEV1 >= 8
esetClean$GOLD[esetClean$RATIOc < 0.7 & esetClean$FEV1Pc >= 0.8] <- 1
# GOLD Stage 2 Ratio < 0.7, 50 <= FEV1 < 80
esetClean$GOLD[esetClean$RATIOc < 0.7 & esetClean$FEV1Pc < 0.8] <- 2
# GOLD Stage 3 Ratio < 0.7, 30 <= FEV1 <= 50
esetClean$GOLD[esetClean$RATIOc < 0.7 & esetClean$FEV1Pc < 0.5] <- 3
# GOLD Stage 4 Ratio < 0.7, FEV1 < 30%
esetClean$GOLD[esetClean$RATIOc < 0.7 & esetClean$FEV1Pc < 0.3] <- 4

esetCleanFormers <- removeFactorLevel(esetClean, "SMKc", "1")
esetCleanCurrents <- removeFactorLevel(esetClean, "SMKc", "2")
# Generate a super clean dataset devoid of all patients with other pulmonary diseases
esetSuperClean <- esetClean
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroPNAc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroTBc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroSarcoidc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroIPFc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroIPFc", "DK")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroAtelectasisc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroAtelectasisc", "DK")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroAsthmac", "DK")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroAsthmac", "1")

esetSuperCleanICS <- removeFactorLevel(esetSuperClean, "AllegroCorticosteroidsc", "1")
esetSuperCleanICS <- removeFactorLevel(esetSuperCleanICS, "AllegroCorticosteroidsc", "DK")

# clean up ATF4 data
varLabels(atf4_data) <- "SampleID"
atf4_data$SampleID[grep("ATF4", atf4_data$SampleID)] <- "ATF4"
atf4_data$SampleID[grep("Ctrl", atf4_data$SampleID)] <- "Control"
atf4_data$SampleID <- as.factor(atf4_data$SampleID)

atf4_colors <- c("ATF4" = "red", "Control" = "black")


```

### ATF4 Models
```{r ATF4models}
# generate ATF4 model(s)
atf4Model <- lmFitWrapper(atf4_data, c("SampleID"),
                          name="ATF4 Perturbation Data", adjust.method="none",
                          p.value=0.05, varOfInterest=1)

clabels <- cbind(atf4_colors[as.character(atf4_data$SampleID)])
heatmap3(exprs(atf4_data)[atf4Model$inds, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = atf4_data$SampleID, ColSideColors = clabels,
         main = "ATF4 Genes",
         labCol="")

hmapClusts <- heatmap3(exprs(atf4_data)[atf4Model$inds,], col = bluered,
                       keep.dendro=TRUE,
                       hclustfun=function(d) hclust(d, method="ward.D"),
                       col.clustering = atf4_data$SampleID, ColSideColors=clabels,
                       main="ATF4 Genes", labCol="")

clusters <- cutree(as.hclust(hmapClusts$Rowv), k=2)
atf4Cluster1_dn <- atf4Model$geneSymbols[clusters==1]
atf4Cluster2_up <- atf4Model$geneSymbols[clusters==2]

atf4dn <- names(clusters)[clusters==1]
atf4up <- names(clusters)[clusters==2]
atf4 <- names(clusters)

rlabels <- cbind(smoking_colors[clusters])
heatmap3(exprs(atf4_data)[atf4Model$inds,], col = bluered,
         keep.dendro=TRUE,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = atf4_data$SampleID, ColSideColors=clabels,
         main="ATF4 Genes", labCol="", RowSideColors=rlabels)



```


### Multiplot function definition 
```{r multiplot}
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)
  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)
  numPlots = length(plots)
  
  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                     ncol = cols, nrow = ceiling(numPlots/cols))
    }
  
  if (numPlots==1) {
    print(plots[[1]])
    } else {
      # Set up the page
      grid.newpage()
      pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
      
      # Make each plot, in the correct location
      for (i in 1:numPlots) {
        # Get the i,j matrix positions of the regions that contain this subplot
        matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
        print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                        layout.pos.col = matchidx$col))
        }
      }
  }

```

### Plot ATF4 leading edge genes from Steiling 2013 in ATF4 and AEGIS data non-Cancer
```{r plot}

atf4_data$SampleID <- relevel(atf4_data$SampleID,ref="Control")
LEgeneList = c("3934_at", "2353_at", "1847_at", "9982_at", "5743_at", "3371_at", 
             "10643_at", "6289_at", "26154_at", "1836_at", "629_at", "11213_at", 
             "6288_at")

# generate figure of the ATF4 genes in the ATF4 data
inds <- match(LEgeneList, featureNames(atf4_data))

clabels <- cbind(atf4_colors[as.character(atf4_data$SampleID)])
heatmap3(exprs(atf4_data)[inds, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = atf4_data$SampleID, ColSideColors=clabels,
         main="ATF4 Leading Edge Genes", labCol="", row.clustering="supervised")

# generate figure of the ATF4 genes in the AEGIS data
eset <- removeFactorLevel(removeFactorLevel(removeBioReps(esetClean), "FinalCaDXc", "1"), "AllegroCorticosteroidsc", "DK")
inds2 <- match(LEgeneList, featureNames(eset))
clabels <- cbind(copd_colors[as.character(eset$COPD2_R7)], smoking_colors[as.character(eset$SMKc)])

heatmap3(exprs(eset)[inds2, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = eset$COPD2_R7, ColSideColors=clabels,
         main="ATF4 Leading Edge Genes in AEGIS non-Cancer", labCol="",
         row.clustering="supervised")

atf4LEtest <- lmFitWrapper(eset[inds2, ], c("COPD2_R7", "AllegroCorticosteroidsc", "GENDERc", "AGEcalc", "SMKc"), 1, "fdr", 0.25)

heatmap3(exprs(eset)[inds2[atf4LEtest$inds], ], col=bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering=eset$COPD2_R7, ColSideColors=clabels,
         main="DE ATF4 LE Genes in AEGIS non-Cancer", labCol="")

```

