Testing R Markdown with Allegro Analysis
========================================================
2014-06-05
-------------------------------------------
This is an R Mardown document. I will be running an analysis for the Allegro CODP+Cancer project. 
In this analysis I will be using only patients with PFTs and using the definition of COPD as R < 0.7
I will also hope to include analysis using only PFTs with R < 0.6, < 0.5, etc. to determine
if there is an effect of changing the definition of COPD. I can also later try to use LLN as a
different definition or alternatively use FEV1 or some other continuous measure


```{r Running AllegroSetup.R}
# AllegroSetup.R uses an RDS file with 800+ samples (biological replicates) and without QC
source("/protected/projects/pulmarray/Allegro/COPD_Cancer/scripts/AllegroSetup.R")
```

### Analysis in R starts here:
```{r Analysis in R starts here:}

# find the samples with PFT information
pftInds <- which(!is.na(exprData$RATIOc))
# subset the dataset to include only those patients with PFTs
exprDataPFT <- exprData[,pftInds]

# calculate/create an indicator column
exprDataPFT <- calcIndicator(exprDataPFT, "COPD2_R7")

# remove the NAs from COPD2_R7
exprDataPFT <- cleanNAForAnalysis(exprDataPFT, "COPD2_R7")

# remove the DKs from COPD2_R7
exprDataPFT <- removeFactorLevel(exprDataPFT, "COPD2_R7", "DK")
exprDataPFT <- removeFactorLevel(exprDataPFT, "FinalCaDXc", "DK")

covarsR7 <- list("SMKc", "COPD2_R7", "FinalCaDXc", "COPD2_R7*FinalCaDXc", "SMKc*COPD2_R7", 
                 "SMKc*FinalCaDXc")
testPFT1 <- analyze_covariate_lists(exprDataPFT, covarsR7)

# testPFTs has no significant by FDR genes for COPD*Cancer but many for p
# will try filtering data down to smaller subset and see what happens
exprDataPFT <- removeFactorLevel(exprDataPFT, "indicator", "0")
exprDataPFT <- filtering(exprDataPFT)
testPFT2 <- analyze_covariate_lists(exprDataPFT$cv, covarsR7)


clabels <- cbind("Group"=copdca_colors[exprDataPFT$cv$indicator],
                 "Cancer"=cancer_colors[exprDataPFT$cv$FinalCaDXc],
                 "COPD"=copd_colors[exprDataPFT$cv$COPD2_R7])


```


```{r fig.width=7, fig.height=6}
# find the indices to print in the heatmap
inds <- testPFT2$cutoffs_p[[3]]$'COPD2_R71:FinalCaDXc1'
  
heatmap3(exprs(exprDataPFT$cv)[inds,], col=bluered, ColSideColors=clabels, col.clustering=exprDataPFT$cv$indicator, main=paste(length(inds), "genes at p < 0.005 for COPD*Cancer interaction"))



```

I'd also like to try running this by including only those patients with severe COPD  
i.e. decreasing the cutoffs by which we define COPD  

Done for the day though.  
For tomorrow:  
  Try running with LLN definition of COPD    
  Try running with different thresholds for R defined COPD  
    
For later:  
  Use SVA  
  Use forward selection  

