Work Done Post Conversation with Isaac
========================================================
`r Sys.Date()`
---------------------------------------------

<h4>To Do from Marc:</h4>

1. Nature exploration with a few genes
2. Explore further visualization: PCA, scatter plots, residuals, smoking
3. Inflammation and cancer datasets
4. Tumor vs normal comparisons


<h4>Thoughts from the paper investigating EMT/UPR in gastric cancer</h4>

1. check out vimentin (VIM/7431)
2. check out e-cadherin (CDH1/999)

<h4>Some thoughts after conversation with Isaac</h4>
Maybe we're not asking the question quite right with the interaction term. Also, there is a lot of noise in the data due to the presence of other non-COPD and non-malignancy pulmonary diseases - TB, atelectasis, pneumonia, asthma, etc. How about we remove these other diseases? Also, let's try seeing what these contrasts reveal: 1) COPD-Cancer, 2) COPD+Cancer - Cancer, 3) COPD+Cancer - COPD, 4) (COPD+Cancer - COPD) - (Cancer - COPD) = COPD+Cancer - Cancer

side note: smoker=2 is former; smoker=1 is current
  
```{r defineFunctions, include=FALSE}
foldChange <- function(eset, groups, genes){
  apply(exprs(eset)[genes, groups==0], 1, mean) - 
    apply(exprs(eset)[genes, groups==1], 1, mean)
}

medianFilter <- function(eset){
  md <- median(exprs(eset))
  passFilter <- logical(featureNumber(eset))
  for(i in 1:featureNumber(eset)){
    passFilter[i] <- sum(exprs(eset)[i, ] > md) > 0
  }
  
  eset <- eset[passFilter, ]
  return(eset)
  
}

saveGeneList <- function(analysis, filename=paste(Sys.Date(), "Temp", analysis$name, ".txt", sep="")){
  
  write.table(analysis$geneSymbols, file=filename, quote=FALSE,row.names=FALSE, col.names=FALSE)

}

default.filename <- function(){ return(paste(Sys.Date(), "Temp", sep="-"))}
```

```{r setup}

setwd("/protected/projects/pulmarray/Allegro/COPD_Cancer/experiments/2015-04-08-New-Perspective-On-Interaction/")
source("/protected/projects/pulmarray/Allegro/COPD_Cancer/scripts/AllegroSetup.R")
# fix the one patient with wonky data
holdEset <- eset
holdEset$FEV1Pc[holdEset$FEV1Pc==89.2] <- 0.892

# load the Lam (n=238) data
#lamData <- readRDS("/protected/projects/pulmarray/Allegro/COPD_Cancer/experiments/lam/LamCOPD_Validate_2014-12-01_ExpressionSet.rds")

esetClean <- cleanNAForAnalysis(holdEset, "COPD2_R7")
esetClean <- removeFactorLevel(esetClean, "FinalCaDXc", "DK")
esetClean <- removeFactorLevel(esetClean, "COPD2_R7", "DK")
esetClean <- calcIndicator(esetClean, "FinalCaDXc", "COPD2_R7")
esetClean$smkindic <- as.numeric(as.character(esetClean$indicator))
esetClean$smkindic[esetClean$SMKc==1] <- esetClean$smkindic[esetClean$SMKc==1] + 4

smkindic_nms <- c("F--","F+-", "F-+", "F++", "C--", "C+-", "C-+", "C++")


```


```{r gastricEMTGenes}

eset <- esetClean
vim <- "999_at"; vimID <- match(vim, rownames(eset))
cdh1 <- "7431_at"; cdh1ID <- match(cdh1, rownames(eset))

# create the smoking X disease indicator

boxplot(exprs(eset)[vimID,] ~ eset$smkindic,
        names=smkindic_nms, main="VIM", col=copdca_colors)

boxplot(exprs(eset)[cdh1ID,] ~ eset$smkindic,
        names=smkindic_nms, main="CDH1", col=copdca_colors)

```

```{r threeWayInteraction}

eset <- esetClean
threeWay <- lmFitWrapper(eset, c("SMKc*FinalCaDXc*COPD2_R7"), 1, "fdr", 0.05, "ThreeWayInteraction")

ccIntrx <- which(p.adjust(threeWay$fit$p.value[, 7], method="fdr") < 0.05)
threeWayIntrx <- which(p.adjust(threeWay$fit$p.value[, 8], method="fdr") < 0.05)

boxplot(exprs(eset)[threeWayIntrx[2],] ~ eset$smkindic)

generate_heatmap(threeWayIntrx, esetClean)

```


```{r cancerVBoth}
eset <- esetClean
eset <- removeFactorLevel(eset, "COPD2_R7", "0")
copdvccc <- lmFitWrapper(eset, c("FinalCaDXc"), 1,"fdr", 0.05, "COPDvCC")

pdf("COPDvCC.pdf")
for(i in 1:54)
  {
  boxplot(exprs(eset)[copdvccc$inds[1],] ~ eset$FinalCaDXc, col=cancer_colors,
          main=copdvccc$geneSymbols[1], notch=TRUE)
  
  }


```

```{r cleanerDataAndNewPerspective}

# we're going to run a new analysis instead of the classic interaction term
# the classic: (COPD+Cancer - COPD) - (Cancer - 'Normal')
# However, the 'Normal' group isn't really normal, but rather is people with
# suspicious pulmonary nodules who are smokers. 
# On Isaac's suggestion, we are going to look at:
#   (COPD+Cancer - COPD) - (Cancer - COPD)
# This should allow us to find different patterns than what we've been finding,
# which has been mostly dominated by the cancer signal, and which may actually
# be what that analysis is designed to find.

eset <- medianFilter(esetClean)
eset <- eset
eset <- removeFactorLevel(eset, "AllegroPNAc", "1")
eset <- cleanNAForAnalysis(eset, "COPD2_R7")
eset <- removeFactorLevel(eset, "AllegroTBc", "1")
eset <- removeFactorLevel(eset, "AllegroSarcoidc", "1")
eset <- removeFactorLevel(eset, "AllegroIPFc", "1")
eset <- removeFactorLevel(eset, "AllegroIPFc", "DK")
eset <- removeFactorLevel(eset, "AllegroAtelectasisc", "1")
eset <- removeFactorLevel(eset, "AllegroAtelectasisc", "DK")

eset$Cancer <- as.character(eset$FinalCaDXc)
eset$Cancer[eset$FinalCaDXc==1] <- "Cancer"
eset$Cancer[eset$FinalCaDXc==0] <- "noCancer"

eset$COPD <- as.character(eset$COPD2_R7)
eset$COPD[eset$COPD2_R7==1] <- "COPD"
eset$COPD[eset$COPD2_R7==0] <- "noCOPD"

eset$CC <- as.factor(paste(eset$Cancer, eset$COPD, sep="."))

eset <- removeFactorLevel(eset, "indicator", "1")
eset <- removeFactorLevel(eset, "CC", "noCancer.noCOPD")

design <- model.matrix(~0 + eset$CC)
colnames(design) <- levels(eset$CC)

fit <- lmFit(eset, design)

# of interest: (COPD+Cancer - COPD) - (Cancer - COPD) = COPD+Cancer - Cancer
cont.matrix <- makeContrasts(
  #Difference=(Cancer.COPD-noCancer.COPD)-(Cancer.noCOPD-noCancer.COPD),
  COPDvCancer=Cancer.noCOPD-noCancer.COPD,
  COPDCavCOPD=Cancer.COPD-noCancer.COPD,
  COPDCavCancer=Cancer.COPD-Cancer.noCOPD,
  levels=design)

fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)

fdr <- 0.05
results <- decideTests(fit2, adjust.method="fdr", p.value=fdr)
vennDiagram(results)
COPDCavCancer.Only <- which(results[, 3] != 0 & results[, 1] == 0 & results[, 2] == 0)
g1 <- fit2$genes$Symbol[COPDCavCancer.Only]

# careful with this list (does not compare contrasts at same FDR values)
fdr <- 0.1
results2 <- decideTests(fit2, adjust.method="fdr", p.value=fdr)
vennDiagram(results2)
COPDCavCancer.Only2 <- which(results2[, 3] != 0 & results[, 1] == 0 & results[, 2] == 0)
g2 <- fit2$genes$Symbol[COPDCavCancer.Only2]

# save the gene list
filename <- paste(default.filename(), "COPDCavCancer.Only2.GeneList.fdrMix.txt", sep="-")
write.table(fit2$genes$Symbol[COPDCavCancer.Only2], file=filename, quote=FALSE,row.names=FALSE, col.names=FALSE)

# save these gene boxplots
filename <- paste(default.filename(), "COPDCavCancer.Only2.Boxplots.fdrMix.pdf")
pdf(filename)
for(i in 1:length(COPDCavCancer.Only2)){
  ind <- COPDCavCancer.Only2[i]
  boxplot(exprs(eset)[ind, ] ~ eset$CC,
          names=c("COPD+Cancer", "Cancer", "COPD"),
          col=c("purple", "red", "blue"), notch=TRUE,
          main=fit2$genes$Symbol[ind],
          xlab="Group", ylab="Gene Expression")
  }
dev.off()









# X genes at fdr < 0.05
# these genes are different between COPD and Cancer
# they also differ between COPD+Cancer and COPD
# and lastly, and most importantly, the differences are different












COPDCavCancer.Only <- which(results[, 3] != 0 & results[, 1] == 0 & results[, 2] == 0)
# save this gene list
filename <- paste(default.filename(), "COPDCavCancer.Only.GeneList.fdr1.txt", sep="-")
write.table(fit2$genes$Symbol[COPDCavCancer.Only], file=filename, quote=FALSE,row.names=FALSE, col.names=FALSE)

# save these gene boxplots
filename <- paste(default.filename(), "COPDCavCancer.Only.Boxplots.fdr1.pdf")
pdf(filename)
for(i in 1:length(COPDCavCancer.Only)){
  ind <- COPDCavCancer.Only[i]
  boxplot(exprs(eset)[ind, ] ~ eset$CC,
          names=c("COPD+Cancer", "Cancer", "COPD"),
          col=c("purple", "red", "blue"), notch=TRUE,
          main=fit2$genes$Symbol[ind],
          xlab="Group", ylab="Gene Expression")
  }
dev.off()


COPDCavCancer <- which(results[, 2] == 0 & results[, 3] != 0)
COPDCavCancer.Cancer.Down <- which(results[, 2] == 0 & results[, 3] < 0)
COPDCavCancer.Cancer.Up <- which(results[, 2] == 0 & results[, 3] > 0)

COPDCavCancer.All.Overlap <- which(results[, 1] != 0 & results[, 2] != 0 & results[, 3] != 0)
# save this genelist
filename <- paste(default.filename(), "COPDCavCancer.All.Overlap.GeneList.txt")
write.table(fit2$genes$Symbol[COPDCavCancer.All.Overlap], file=filename, quote=FALSE,row.names=FALSE, col.names=FALSE)

# save these boxplots
filename <- paste(default.filename(), "COPDCavCancer.All.Overlap.Boxplots.pdf")
pdf(filename)
for(i in 1:length(COPDCavCancer.All)){
  ind <- COPDCavCancer.All.Overlap[i]
  boxplot(exprs(eset)[ind, ] ~ eset$CC,
          names=c("COPD+Cancer", "Cancer", "COPD"),
          col=c("purple", "red", "blue"), notch=TRUE,
          main=fit2$genes$Symbol[ind],
          xlab="Group", ylab="Gene Expression")
  }
dev.off()

COPDCavCancer.All <- which(results[, 3] != 0)
# save this genelist
filename <- paste(default.filename(), "COPDCavCancer.All.GeneList.txt")
write.table(fit2$genes$Symbol[COPDCavCancer.All], file=filename, quote=FALSE,row.names=FALSE, col.names=FALSE)
# save these boxplots
filename <- paste(default.filename(), "COPDCavCancer.All.Boxplots.pdf", sep="-")
pdf(filename)
for(i in 1:length(COPDCavCancer.All)){
  ind <- COPDCavCancer.All[i]
  boxplot(exprs(eset)[ind, ] ~ eset$CC,
          names=c("COPD+Cancer", "Cancer", "COPD"),
          col=c("purple", "red", "blue"), notch=TRUE,
          main=fit2$genes$Symbol[ind],
          xlab="Group", ylab="Gene Expression")
  }
dev.off()



contrastOfInt <- which(results[, 3] != 0 & results[, 2] == 0 & results[, 1] == 0)

filename <- paste(Sys.Date(), "new-perspective-COPDCancer-v-Cancer.pdf", sep="-")
pdf(filename)
for (i in 1:length(COPDCavCancer)) {
  ind <- COPDCavCancer[i]
  boxplot(exprs(eset)[ind, ] ~ eset$CC,
          names=c("COPD+Cancer", "Cancer", "COPD"),
          col=c("purple", "red", "blue"), notch=TRUE,
          main=fit2$genes$Symbol[ind],
          xlab="Group", ylab="Gene Expression")
  }
dev.off()

# save all of the genes, regardless of direction
filename <- paste(Sys.Date(), "COPDCavCancer", length(COPDCavCancer),"genes.txt", sep="-")
write.table(fit2$genes$Symbol[COPDCavCancer], file=filename, quote=FALSE,row.names=FALSE, col.names=FALSE)

# save the genes up in cancer vs COPD+Cancer
filename <- paste(Sys.Date(), "COPDCavCancer-Cancer-Up", length(COPDCavCancer.Cancer.Up),"genes.txt", sep="-")
write.table(fit2$genes$Symbol[COPDCavCancer.Cancer.Up], file=filename, quote=FALSE,row.names=FALSE, col.names=FALSE)

# save the genes down in cancer vc COPD+Cancer
filename <- paste(Sys.Date(), "COPDCavCancer-Cancer-Down", length(COPDCavCancer.Cancer.Down),"genes.txt", sep="-")
write.table(fit2$genes$Symbol[COPDCavCancer.Cancer.Down], file=filename, quote=FALSE,row.names=FALSE, col.names=FALSE)

# print the boxplot for EGR1 - the top hit TF from Enrichr for the 300 genes
ind <- match("EGR1", fit2$genes$Symbol)
boxplot(exprs(eset)[ind, ] ~ eset$CC,
        names=c("COPD+Cancer", "Cancer", "COPD"),
        col=c("purple", "red", "blue"), notch=TRUE,
        main=fit2$genes$Symbol[ind],
        xlab="Group", ylab="Gene Expression")
```

