Work Done Post Conversation with Isaac
========================================================
`r Sys.Date()`
---------------------------------------------

<h4>Some thoughts after conversation with Isaac</h4>
Maybe we're not asking the question quite right with the interaction term. Also, there is a lot of noise in the data due to the presence of other non-COPD and non-malignancy pulmonary diseases - TB, atelectasis, pneumonia, asthma, etc. How about we remove these other diseases? Also, let's try seeing what these contrasts reveal: 1) COPD-Cancer, 2) COPD+Cancer - Cancer, 3) COPD+Cancer - COPD, 4) (COPD+Cancer - COPD) - (Cancer - COPD) = COPD+Cancer - Cancer

side note: smoker=2 is former; smoker=1 is current
  
```{r defineFunctions, include=FALSE}
foldChange <- function(eset, groups, genes){
  apply(exprs(eset)[genes, groups==0], 1, mean) - 
    apply(exprs(eset)[genes, groups==1], 1, mean)
}

medianFilter <- function(eset){
  md <- median(exprs(eset))
  passFilter <- logical(featureNumber(eset))
  for(i in 1:featureNumber(eset)){
    passFilter[i] <- sum(exprs(eset)[i, ] > md) > 0
  }
  
  eset <- eset[passFilter, ]
  return(eset)
  
}

saveGeneList <- function(analysis, filename=paste(Sys.Date(), "Temp", analysis$name, ".txt", sep="")){
  
  write.table(analysis$geneSymbols, file=filename, quote=FALSE,row.names=FALSE, col.names=FALSE)

}

default.filename <- function(filename=""){ return(paste(Sys.Date(), "Temp", filename,sep="-"))}

outputGenes <- function(eset, fit, inds, filename=default.filename(), groups=eset$CC, nms=c("COPD+Cancer", "Cancer", "COPD"), cols=c("purple", "red", "blue")){
  fn <- default.filename(filename)
  write.table(fit$genes$Symbol[inds], file=paste(fn, "txt", sep="."), quote=FALSE, row.names=FALSE,col.names=FALSE)
  
  pdf(paste(fn, "pdf", sep="."))
  
  for(i in 1:length(inds)){
    ind <- inds[i]
    boxplot(exprs(eset)[ind, ] ~ groups,
            names=nms,
            col=cols, notch=TRUE,
            main=fit$genes$Symbol[ind],
            xlab="Group", ylab="Gene Expression")
    }
  dev.off()
  
  }
```

```{r setup}

setwd("/protected/projects/pulmarray/Allegro/COPD_Cancer/experiments/2015-04-08-New-Perspective-On-Interaction/")
source("/protected/projects/pulmarray/Allegro/COPD_Cancer/scripts/AllegroSetup.R")
# fix the one patient with wonky data
holdEset <- eset
holdEset$FEV1Pc[holdEset$FEV1Pc==89.2] <- 0.892

# load the Lam (n=238) data
#lamData <- readRDS("/protected/projects/pulmarray/Allegro/COPD_Cancer/experiments/lam/LamCOPD_Validate_2014-12-01_ExpressionSet.rds")

esetClean <- cleanNAForAnalysis(holdEset, "COPD2_R7")
esetClean <- removeFactorLevel(esetClean, "FinalCaDXc", "DK")
esetClean <- removeFactorLevel(esetClean, "COPD2_R7", "DK")
esetClean <- calcIndicator(esetClean, "FinalCaDXc", "COPD2_R7")
esetClean$smkindic <- as.numeric(as.character(esetClean$indicator))
esetClean$smkindic[esetClean$SMKc==2] <- esetClean$smkindic[esetClean$SMKc==2] + 4
esetClean$smkindic <- as.factor(esetClean$smkindic)

smkindic_nms <- c("C--", "C+-", "C-+", "C++","F--","F+-", "F-+", "F++")


```


```{r cleanerDataAndNewPerspective}

# we're going to run a new analysis instead of the classic interaction term
# the classic: (COPD+Cancer - COPD) - (Cancer - 'Normal')
# However, the 'Normal' group isn't really normal, but rather is people with
# suspicious pulmonary nodules who are smokers. 
# On Isaac's suggestion, we are going to look at:
#   (COPD+Cancer - COPD) - (Cancer - COPD)
# This should allow us to find different patterns than what we've been finding,
# which has been mostly dominated by the cancer signal, and which may actually
# be what that analysis is designed to find.

# 2015-04-13 noticed that medianFilter should be applied after the 
# sample set is determined/cleaned
#eset <- medianFilter(esetClean)
eset <- esetClean
eset <- removeFactorLevel(eset, "AllegroPNAc", "1")
eset <- cleanNAForAnalysis(eset, "COPD2_R7")
eset <- removeFactorLevel(eset, "AllegroTBc", "1")
eset <- removeFactorLevel(eset, "AllegroSarcoidc", "1")
eset <- removeFactorLevel(eset, "AllegroIPFc", "1")
eset <- removeFactorLevel(eset, "AllegroIPFc", "DK")
eset <- removeFactorLevel(eset, "AllegroAtelectasisc", "1")
eset <- removeFactorLevel(eset, "AllegroAtelectasisc", "DK")

eset$Cancer <- as.character(eset$FinalCaDXc)
eset$Cancer[eset$FinalCaDXc==1] <- "Cancer"
eset$Cancer[eset$FinalCaDXc==0] <- "noCancer"

eset$COPD <- as.character(eset$COPD2_R7)
eset$COPD[eset$COPD2_R7==1] <- "COPD"
eset$COPD[eset$COPD2_R7==0] <- "noCOPD"

eset$CC <- as.factor(paste(eset$Cancer, eset$COPD, sep="."))

# apply a gene filter
eset <- medianFilter(eset)

# removing the no-disease patients should not have an effect on the
# analysis COPD+Cancer - Cancer (I don't think)
#eset <- removeFactorLevel(eset, "indicator", "1")
#eset <- removeFactorLevel(eset, "CC", "noCancer.noCOPD")

design <- model.matrix(~0 + eset$CC)
colnames(design) <- levels(eset$CC)

fit <- lmFit(eset, design)

# of interest: (COPD+Cancer - COPD) - (Cancer - COPD) = COPD+Cancer - Cancer
# realistically we want a few groups of the genes significant for these terms
# 1. the genes differing between COPD.Cancer and Cancer
# 2. the (1) genes and those also changing between COPD.Cancer and COPD
cont.matrix <- makeContrasts(
  #Difference=(Cancer.COPD-noCancer.COPD)-(Cancer.noCOPD-noCancer.COPD),
  COPDvCancer=Cancer.noCOPD-noCancer.COPD,
  COPDCavCOPD=Cancer.COPD-noCancer.COPD,
  COPDCavCancer=Cancer.COPD-Cancer.noCOPD,
  levels=design)

fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)

fdr <- 0.05
results <- decideTests(fit2, adjust.method="fdr", p.value=fdr)
vennDiagram(results)
COPDCavCancer.Only <- which(results[, 3] != 0 & results[, 1] == 0 & results[, 2] == 0)
COPDCavCancer.All <- which(results[, 3] != 0)
COPDCavCancervCOPD <- which(results[, 3] != 0 & results[, 2] != 0 & results[, 1] != 0)
g1.Only <- fit2$genes$Symbol[COPDCavCancer.Only]
g1.All <- fit2$genes$Symbol[COPDCavCancer.All]
g1.triple <- fit2$genes$Symbol[COPDCavCancervCOPD]

# careful with this list (does not compare contrasts at same FDR values)
# i.e. we try to increase the FDR for just the contrast of interest, 
# still using the fdr < 0.5 decideTest results for the other contrasts.
# this will allow us to increase the FDR and keep all of the same genes
# in the list of results
fdr <- 0.1
results2 <- decideTests(fit2, adjust.method="fdr", p.value=fdr)
vennDiagram(results2)
COPDCavCancer.Only2 <- which(results2[, 3] != 0 & results[, 1] == 0 & results[, 2] == 0)
COPDCavCancer.All2 <- which(results2[, 3] != 0)
COPDCavCancervCOPD.2 <- which(results2[, 3] != 0 & results[, 2] != 0 & results[, 1] != 0)
g2.Only <- fit2$genes$Symbol[COPDCavCancer.Only2]
g2.All <- fit2$genes$Symbol[COPDCavCancer.All2]
g2.triple <- fit2$genes$Symbol[COPDCavCancervCOPD.2]




#outputGenes(eset, fit2, COPDCavCancer.Only, "g1.only.4groups",
#            nms=levels(eset$CC), cols=c("purple", "red", "blue", "black"))
#outputGenes(eset, fit2, COPDCavCancer.Only2, "g2.only.4groups",
#            nms=levels(eset$CC), cols=c("purple", "red", "blue", "black"))
#outputGenes(eset, fit2, COPDCavCancer.All, "g1.all.4groups",
#            nms=levels(eset$CC), cols=c("purple", "red", "blue", "black"))
# too many genes - 1800+
#outputGenes(eset, fit2, COPDCavCancer.All2, "g2.all.4groups",
#           nms=levels(eset$CC), cols=c("purple", "red", "blue", "black"))
#outputGenes(eset, fit2, COPDCavCancervCOPD.2, "g2.triple.4groups",
#            nms=levels(eset$CC), cols=c("purple", "red", "blue", "black"))
#outputGenes(eset, fit2, COPDCavCancervCOPD, "g1.triple.4groups",
#            nms=levels(eset$CC), cols=c("purple", "red", "blue", "black"))

# save out by groups of 8 the list of g1.triple
#outputGenes(eset=eset, fit=fit2, inds=COPDCavCancervCOPD,
#            filename="g1.triple.8groups",
#            groups=eset$smkindic,
#            nms=smkindic_nms,
#            cols=c("purple", "red", "blue","black",
#                   "purple", "red", "blue", "black"))



library(znorm)
pca_g1.Only <- prcomp(znorm(x=exprs(eset)[COPDCavCancer.Only,], margin=1), center=FALSE, scale.=FALSE)

# this shows the overall difference but isn't otherwise informative more than a gene would be
boxplot(pca_g1.Only$rotation[, 1] ~ eset$CC, names=levels(eset$CC),
            col=c("purple", "red", "blue", "black"), notch=TRUE,
            main="PC1 from COPDCavCancer.Only ",
            xlab="Group", ylab="First Principal Component")

# try breaking out some genes by smoking status in a boxplot
MSLNind <- COPDCavCancer.Only[match("MSLN", g1.Only)]
boxplot(exprs(eset)[MSLNind, ] ~ eset$smkindic,
            names=smkindic_nms,
            col=copdca_colors, notch=TRUE,
            main="MSLN",
            xlab="Group", ylab="Gene Expression")

# look at a heatmap of all samples and all genes, regardless of disease/smoking
temp <- eset
exprs(temp) <- znorm(x=exprs(temp), margin=1)
clabels <- cbind("Indicator" = copdca_colors[temp$CC])
colClus <- temp$CC
heatmap3(exprs(temp)[COPDCavCancer.All,], col = bluered, hclustfun=function(d) hclust(d, method="ward.D"), col.clustering = "unsupervised", ColSideColors = clabels)

# try removing the 'normal' patients
temp <- eset
temp <- removeFactorLevel(temp, "indicator", "1")
exprs(temp) <- znorm(x=exprs(temp), margin=1)
clabels <- cbind("Indicator" = copdca_colors[temp$CC],
                 "Smoking" = smoking_colors[temp$SMKc])
colClus <- temp$CC
heatmap3(exprs(temp)[COPDCavCancer.All,], col = bluered, hclustfun=function(d) hclust(d, method="ward.D"), col.clustering = "unsupervised", ColSideColors = clabels)


# try looking at a heatmap
temp <- eset
temp <- removeFactorLevel(eset, "SMKc","1")
exprs(temp) <- znorm(x=exprs(temp), margin=1)
temp$FEV1bin <- temp$FEV1Pc <= 0.8
temp$FEV1bina <- numeric(sampleNumber(temp))
temp$FEV1bina[temp$FEV1bin] <- 1
temp$FEV1bina <- as.factor(temp$FEV1bina)

clabels <- cbind("Indicator" = copdca_colors[temp$CC],
                 "Smoking Status" = smoking_colors[temp$SMKc],
                 "Gender" = gender_colors[temp$GENDERc],
                 "FEV1p" = cancer_colors[temp$FEV1bina])
colClus <- temp$CC

# former smokers only
# the smallest group
heatmap3(exprs(temp)[COPDCavCancer.Only,], col = bluered, hclustfun=function(d) hclust(d, method="average"), col.clustering = colClus, ColSideColors = clabels)

# unclustered smallest group of genes at fdr < 0.05
clabels <- cbind("Indicator" = copdca_colors[temp$CC])
heatmap3(exprs(temp)[COPDCavCancer.Only,], col = bluered, hclustfun=function(d) hclust(d, method="ward.D"), col.clustering = "unsupervised", ColSideColors = clabels)

# unclustered smallest group of genes at fdr < 0.10
heatmap3(exprs(temp)[COPDCavCancer.Only2,], col = bluered, hclustfun=function(d) hclust(d, method="ward.D"), col.clustering = "unsupervised", ColSideColors = clabels)

# the largest group
heatmap3(exprs(temp)[COPDCavCancer.All,], col = bluered, hclustfun=function(d) hclust(d, method="average"), col.clustering = colClus, ColSideColors = clabels)

# the largest group unclustered with ward
clabels <- cbind("Indicator" = copdca_colors[temp$CC],
                 "Ratio" = smoking_colors[temp$FEV1bina])
heatmap3(exprs(temp)[COPDCavCancer.All,], col = bluered, hclustfun=function(d) hclust(d, method="ward.D"), col.clustering = "unsupervised", ColSideColors = clabels)

# largest group clustered, but with different color bars than above
heatmap3(exprs(temp)[COPDCavCancer.All,], col = bluered, hclustfun=function(d) hclust(d, method="ward.D"), col.clustering = colClus, ColSideColors = clabels)

# remove the 'normal' patients
temp <- removeFactorLevel(temp, "indicator", "1") 
clabels <- cbind("Indicator" = copdca_colors[temp$CC],
                 "Ratio" = smoking_colors[temp$FEV1bina])
colClus <- temp$CC
heatmap3(exprs(temp)[COPDCavCancer.All,], col = bluered, hclustfun=function(d) hclust(d, method="ward.D"), col.clustering = colClus, ColSideColors = clabels)

# try unclustering
# this is my favorite
clabels <- cbind("Indicator" = copdca_colors[temp$CC])
heatmap3(exprs(temp)[COPDCavCancer.All,], col = bluered, hclustfun=function(d) hclust(d, method="ward.D"), col.clustering = "unsupervised", ColSideColors = clabels)

# let's look at current smokers
temp <- eset
temp <- removeFactorLevel(eset, "SMKc","2")
exprs(temp) <- znorm(x=exprs(temp), margin=1)
temp$FEV1bin <- temp$FEV1Pc < 0.7
temp$FEV1bina <- numeric(sampleNumber(temp))
temp$FEV1bina[temp$FEV1bin] <- 1
temp$FEV1bina <- as.factor(temp$FEV1bina)

clabels <- cbind("Indicator" = copdca_colors[temp$CC],
                 "Smoking Status" = smoking_colors[temp$SMKc],
                 "Gender" = gender_colors[temp$GENDERc],
                 "FEV1p" = cancer_colors[temp$FEV1bina])

colClus <- temp$CC

# the largest group unclustered with ward
heatmap3(exprs(temp)[COPDCavCancer.All,], col = bluered, hclustfun=function(d) hclust(d, method="ward.D"), col.clustering = "unsupervised", ColSideColors = clabels)

# same heatmap but with fewer column indicator bars
clabels <- cbind("Indicator" = copdca_colors[temp$CC])
heatmap3(exprs(temp)[COPDCavCancer.All,], col = bluered, hclustfun=function(d) hclust(d, method="ward.D"), col.clustering = "unsupervised", ColSideColors = clabels)

# try removing the 'normal' patients
temp <- removeFactorLevel(temp, "indicator", 1)
clabels <- cbind("Indicator" = copdca_colors[temp$CC])
colClus <- temp$CC
heatmap3(exprs(temp)[COPDCavCancer.All,], col = bluered, hclustfun=function(d) hclust(d, method="ward.D"), col.clustering = "unsupervised", ColSideColors = clabels)

```

