07-16-2014
========================================================

### Changing the COPD definition to include only moderate-severe disease

On Yaron's suggestion I will try changing the COPD definition from strictly Ratio < 0.7 to Ratio < 0.7 and FEV1P < 0.5. There are approximately 70 patients with these PFT values.

I'll also look at PCA within each group to check for strange outliers.

**Note**: there is one patients marked with an FEV1P of 89 - should either be removed or set to 0.89.

Some terms to include:  
  * history of cancer  
  * time since quit
  * age  
  * pack years
  * cancer cell type
  * cancer sub cell type 
  * corticosteroids  - correlate with SVA?
  
*Add email communications regarding this project*

  
```{r GlobalVariables, include=FALSE}
cacheOption = FALSE

```


```{r defineFunctions, include=FALSE}
foldChange <- function(eset, groups, genes){
  apply(exprs(eset)[genes, groups==0], 1, mean) - 
    apply(exprs(eset)[genes, groups==1], 1, mean)
}

medianFilter <- function(eset){
  md <- median(exprs(eset))
  passFilter <- logical(featureNumber(eset))
  for(i in 1:featureNumber(eset)){
    passFilter[i] <- sum(exprs(eset)[i, ] > md) > 0
  }
  
  eset <- eset[passFilter, ]
  return(eset)
  
}

```

```{r LoadData, echo=FALSE, cache=cacheOption, include=FALSE}
# set the working directory to the directory of this script
setwd("/restricted/projectnb/pulmarray/LinGA_protected/Allegro/COPD_Cancer/experiments/2014-07-16")

# load the data using a script that removes patients with:
#  Cancer = NA
#  SMK    = 3
source("/protected/projects/pulmarray/Allegro/COPD_Cancer/scripts/AllegroSetup.R")

# remove the patients who have a DK for Cancer status as this is the a
# phenotype of interest (along with COPD)
holdEset <- removeBioReps(eset)
eset <- holdEset

#eset <- removeFactorLevel(eset, "AllegroCOPDc", "DK")
#eset <- cleanNAForAnalysis(eset, "COPD2_R7")
#eset <- removeFactorLevel(eset, "COPD2_R7", "DK")
#eset <- cleanNAForAnalysis(eset, "AGEcalc")
#eset <- removeBioReps(exprData=eset)
#eset <- calcIndicator(eset, "FinalCaDXc", "AllegroCOPDc")
#eset <- removeFactorLevel(eset, "GENDERc", "DK")
#eset <- calcIndicator(eset, "FinalCaDXc", "COPD2_R7")

library(znorm)

```


```{r SVAcancer, eval=FALSE, include=FALSE}
library(sva)

eset <- removeFactorLevel(eset, "FinalCaDXc", "DK")
eset <- removeBioReps(eset)
mod1 <- model.matrix(~ 1 + FinalCaDXc, data=pData(eset))
mod0 <- model.matrix(~ 1, data=pData(eset))


# Construct the surrogate variables and return them

sva_obj <- sva(exprs(eset), mod1, mod0)


# check the relationship of each of the surrogate variables with the possible phenotype variables available in eset
# character pheonotypes will be ignored
# integer, numeric phenotypes will be tested by correlation
# factor phenotypes will be tested by t.test

# PROBLEM: need to clean to remove DKs and NAs
# When removing those, need to also remove those samples from surrogate variables
tempHold <- eset
pData(tempHold) <- pData(tempHold)[, -2]

for(i in 1:dim(sva_obj$sv)[2]){
  print(paste("i=", i))
  for(j in 1:dim(pData(tempHold))[2]){
    print(paste("j=", j, varLabel))
    if(is.factor(pData(tempHold)[, j])){
      temp <- cleanNAForAnalysis(tempHold, varLabels(tempHold)[j])
      temp <- removeFactorLevel(tempHold, varLabels(tempHold)[j], "DK")
      tSVA <- sva_obj$sv[which(!is.na(match(sampleNames(tempHold), sampleNames(temp)))), ]
      t.test(tSVA[, i] ~ pData(temp)[, j])
    }
    else if (is.numeric(pData(tempHold)[, j])) {
      temp <- cleanNAForAnalysis(eset, pData(tempHold)[, j])
      tSVA <- sva_obj$sv[which(!is.na(match(sampleNames(tempHold), sampleNames(temp)))), ]
      cor(sva_obj$sv[, i], pData(temp)[, j])
    }
  }
}

# let's look at just those variables we're interested in:
# RIN, AGEcalc, GENDERc, SMKc, PYc, 

```

### Defining COPD

Re-define COPD as moderate-severe disease, i..e. patients wtih ratio < 0.7 and FEV1P < 0.5
```{r definingCOPD}
# clean the data for analysis
eset <- holdEset
eset <- removeFactorLevel(eset, "FinalCaDXc", "DK")
eset <- cleanNAForAnalysis(eset, c("RATIOc", "FEV1Pc"))
# change the FEV1Pc that is 89.20 to 0.892 or alternatively remote it
eset$FEV1Pc[which(eset$FEV1Pc==89.2)] <- 0.89

#define COPD as RATIO < 0.7, FEV1Pc < 0.5
eset$COPDmodsev <- rep(2, sampleNumber(eset))
eset$COPDmodsev[eset$RATIOc < 0.7 & eset$FEV1Pc < 0.5] <- 1
# define no COPD as RATIO > 0.7, FEV1Pc > 0.5
eset$COPDmodsev[eset$RATIOc > 0.7 & eset$FEV1Pc >= 0.8] <- 0

# set COPDmodsev as factor (from numeric)
eset$COPDmodsev <- factor(eset$COPDmodsev)
# remove those patients fitting neither COPD nor healthy criteria
eset <- removeFactorLevel(eset, "COPDmodsev", "2")

eset <- calcIndicator(eset, "FinalCaDXc", "COPDmodsev")

# remove patients with IPF, which has its own signal
eset <- removeFactorLevel(eset, "AllegroIPFc", "1")
tapply(eset$indicator, eset$SMKc, summary)

esetMod <- eset
```

### Analyze data with new COPD term
```{r analyseNewCOPDTerm}

# Model 1: simplest model including just cancer and COPD terms
eset <- medianFilter(esetMod)
designNewCOPD <- model.matrix(~ 1 + FinalCaDXc*COPDmodsev, data=eset)
fitNewCOPD <- lmFit(eset, designNewCOPD)
fitNewCOPD <- eBayes(fitNewCOPD)
resultsNewCOPD <- decideTests(fitNewCOPD, adjust.method="fdr", p.value=0.25)
summary(resultsNewCOPD)

#generate_heatmap(which(resultsNewCOPD[, 4] != 0), eset, tp="indicator")

# Model 2: model 1 + SMK
designNewCOPD2 <- model.matrix(~ 1 + FinalCaDXc*COPDmodsev + SMKc, data=eset)
fitNewCOPD2 <- lmFit(eset, designNewCOPD2)
fitNewCOPD2 <- eBayes(fitNewCOPD2)
resultsNewCOPD2 <- decideTests(fitNewCOPD2, adjust.method="fdr", p.value=0.2)
summary(resultsNewCOPD2)

#generate_heatmap(which(resultsNewCOPD2[, 5] != 0), eset, tp="indicator")

# Model 3: model 2 + SMK interactions
designNewCOPD3 <- model.matrix(~ 1 + FinalCaDXc*COPDmodsev + 
                                 FinalCaDXc*SMKc + COPDmodsev*SMKc, data=eset)
fitNewCOPD3 <- lmFit(eset, designNewCOPD3)
fitNewCOPD3 <- eBayes(fitNewCOPD3)
resultsNewCOPD3 <- decideTests(fitNewCOPD3, adjust.method="fdr", p.value=0.1)
summary(resultsNewCOPD3)

#generate_heatmap(which(resultsNewCOPD3[, 5] != 0), eset, tp="indicator")

# Model 4: model 3 + RIN
eset <- cleanNAForAnalysis(eset, "RIN")
designNewCOPD4 <- model.matrix(~ 1 + RIN + FinalCaDXc*COPDmodsev + 
                                 FinalCaDXc*SMKc + COPDmodsev*SMKc, data=eset)
fitNewCOPD4 <- lmFit(eset, designNewCOPD4)
fitNewCOPD4 <- eBayes(fitNewCOPD4)
resultsNewCOPD4 <- decideTests(fitNewCOPD4, adjust.method="fdr", p.value=0.25)
summary(resultsNewCOPD4)

#generate_heatmap(which(resultsNewCOPD4[, 6] != 0), eset, tp="indicator")

# Model 5: interaction just in former smokers
eset <- esetMod
eset <- removeFactorLevel(eset, "SMKc", "1")
eset <- medianFilter(eset)
designFormer <- model.matrix(~ 1 + FinalCaDXc*COPDmodsev, data=eset)
fitFormer <- lmFit(eset, designFormer)
fitFormer <- eBayes(fitFormer)
resultsFormer <- decideTests(fitFormer, adjust.method="fdr", p.value=0.25)
summary(resultsFormer)

#generate_heatmap(which(resultsFormer[, 4] != 0), eset, tp="indicator")

# Model 6: model 5 + GENDERc
eset <- removeFactorLevel(eset, "GENDERc", "DK")
designFormer2 <- model.matrix(~ 1 + FinalCaDXc*COPDmodsev + GENDERc, data=eset)
fitFormer2 <- lmFit(eset, designFormer2)
fitFormer2 <- eBayes(fitFormer2)
resultsFormer2 <- decideTests(fitFormer2, adjust.method="fdr", p.value=0.25)
summary(resultsFormer2)

#generate_heatmap(which(resultsFormer2[, 5] != 0), eset, tp="indicator")

# Model 7: compare COPD to COPD+Cancer in all patients
eset <- esetMod
eset <- medianFilter(eset)
eset <- removeFactorLevel(eset, "indicator", "1")
eset <- removeFactorLevel(eset, "indicator", "2")

designCOPDonly <- model.matrix(~ 1 + FinalCaDXc, data=eset)
fitCOPDonly <- lmFit(eset, designCOPDonly)
fitCOPDonly <- eBayes(fitCOPDonly)
resultsCOPDonly <- decideTests(fitCOPDonly, adjust.method="fdr", p.value=0.25)
summary(resultsCOPDonly)

# Model 8: model 7 plus some terms
designCOPDonly2 <- model.matrix(~ 1 + FinalCaDXc + SMKc, data=eset)
fitCOPDonly2 <- lmFit(eset, designCOPDonly2)
fitCOPDonly2 <- eBayes(fitCOPDonly2)
resultsCOPDonly2 <- decideTests(fitCOPDonly2, adjust.method="fdr", p.value=0.001)
summary(resultsCOPDonly2)

# sanity check - plot the smoking genes unsupervised and make sure they look good
heatmap3(exprs(eset)[which(resultsCOPDonly2[, 3] != 0), ], col = bluered,
           hclustfun=function(d) hclust(d, method="average"),
           col.clustering = "unsupervised", ColSideColors=smoking_colors[eset$SMKc])


# Model 9: in all patients try to find signal for cancer in the context of COPD
eset <- removeFactorLevel(removeFactorLevel(holdEset, "FinalCaDXc", "DK"), "AllegroCOPDc", "DK")
eset <- calcIndicator(eset, "FinalCaDXc", "AllegroCOPDc")
eset <- removeFactorLevel(removeFactorLevel(eset, "indicator", "1"), "indicator", "2")
eset <- removeFactorLevel(eset, "GENDERc", "DK")
COPDeset <- eset
eset <- medianFilter(eset)

designCOPDonly3 <- model.matrix(~ 1 + FinalCaDXc + SMKc + GENDERc, data=eset)
fitCOPDonly3 <- lmFit(eset, designCOPDonly3)
fitCOPDonly3 <- eBayes(fitCOPDonly3)
resultsCOPDonly3 <- decideTests(fitCOPDonly3, adjust.method="fdr", p.value=0.1)
summary(resultsCOPDonly3)

generate_heatmap(which(resultsCOPDonly3[, 2] != 0), eset, tp="indicator")

heatmap3(exprs(eset)[which(resultsCOPDonly3[, 2] != 0), ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward"),
         col.clustering = "unsupervised",
         ColSideColors=cbind(copdca_colors[eset$indicator], smoking_colors[eset$SMKc]))

o <- sort(abs(fitCOPDonly3$t[, 2]), decreasing=TRUE, index.return=TRUE)
heatmap3(exprs(eset)[o$ix[1:10], ], col = bluered,
         hclustfun=function(d) hclust(d, method="average"),
         col.clustering = "unsupervised",
         ColSideColors=cbind(copdca_colors[eset$indicator], smoking_colors[eset$SMKc]))

generate_heatmap(o$ix[1:10], eset, tp="indicator")


```

I think some of the problems I've been encountering stem from the fact that both COPD and lung cancer in this cohort are heterogenous. Especially concering, is that I haven't been able to find a lot of genes that differ between COPD and COPD+Cancer. The last heatmap above illustrates those genes that are statistically different between these two groups (based on all patients, allegro COPD definition), but no clear pattern is demonstrated.

I would like to investigate the nature-medicine cancer signature in this cohort and potentially use a PC1 to score the probability of cancer/premalignant lesions in each of the patients. Perhaps this can shine some light on the cancer heterogeneity. Ditto for the blue journal COPD signature and the observed COPD heterogeneity. First, complete the thoughts below, looking at different terms of interest.

### Model Previously Unused Terms - Are they helpful?

Try modeling some different terms:  
  * LENGTHSMKQUITYRc
  * HXCAc
  * ANATOMYc - look for dysplasia
  * AllegroCorticosteroidsc
  * PYc
  * CA_CELL_TYPE
  * CA_SUB_CELL_TYPE
  
```{r newTerms}
eset <- holdEset
eset$LENGTHSMKQUITYRc[eset$SMKc==1] <- 0
eset$LENGTHSMKQUITMOc[eset$SMKc==1] <- 0
eset <- cleanNAForAnalysis(eset, "LENGTHSMKQUITYRc")
eset <- cleanNAForAnalysis(eset, "LENGTHSMKQUITMOc")
eset <- cleanNAForAnalysis(eset, "PYc")
eset <- medianFilter(eset)
eset$QUITSMKMOcalc <- eset$LENGTHSMKQUITYRc*12 + eset$LENGTHSMKQUITMOc

source("131210_continuous_colorbar.R")

# Model 1: find genes for months since quit (current = 0)
designYrsQuit <- model.matrix(~ 1 + eset$QUITSMKMOcalc*PYc, data=eset)
fitYrsQuit <- lmFit(eset, designYrsQuit)
fitYrsQuit <- eBayes(fitYrsQuit)
resultsYrsQuit <- decideTests(fitYrsQuit, adjust.method="fdr", p.value=0.001)
summary(resultsYrsQuit)

# better off picking the top 100 than trying to find an fdr value
o <- sort(abs(fitYrsQuit$t[, 2]), decreasing=TRUE, index.return=TRUE)
  
heatmap3(exprs(eset)[o$ix[1:200], ], col = bluered,
         hclustfun=function(d) hclust(d, method="average"),
         col.clustering = "unsupervised",
         ColSideColors=cbind(smoking_colors[eset$SMKc]))

# try to explain the variance in the heatmap above
# there appear to be 3 main sample clusters:
#   1) currents, 2) formers, and 3) a mix of both
# make a quartile PY variable (1-4)
eset$PYquart <- rep(4, sampleNumber(eset))
PYqs <- quantile(eset$PYc)
eset$PYquart[eset$PYc <= PYqs[4]] <- 3
eset$PYquart[eset$PYc <= PYqs[3]] <- 2
eset$PYquart[eset$PYc <= PYqs[2]] <- 1

eset$PYquart <- as.factor(eset$PYquart)

heatmap3(exprs(eset)[o$ix[1:200], ], col = bluered,
         hclustfun=function(d) hclust(d, method="average"),
         col.clustering = "unsupervised",
         ColSideColors=cbind(smoking_colors[eset$SMKc], copdca_colors[eset$PYquart]))

# visualize PY*timeSinceQuit genes
o <- sort(abs(fitYrsQuit$t[, 4]), decreasing=TRUE, index.return=TRUE)

heatmap3(exprs(eset)[o$ix[1:200], ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward"),
         col.clustering = "semisupervised",
         ColSideColors=cbind(smoking_colors[eset$SMKc], copdca_colors[eset$PYquart]))

# Model 2: Find genes for history of previous cancer
eset <- removeFactorLevel(holdEset, "HXCAc", "DK")
eset <- medianFilter(eset)
eset$HXCA2 <- rep(2, sampleNumber(eset))
eset$HXCA2[eset$HXCAc=="0"] <- 0
eset$HXCA2[eset$HXCAc!="0"] <- 1
eset$HXCA2 <- as.factor(eset$HXCA2)

designHxCa2 <- model.matrix(~ 1 + HXCA2, data=eset)
fitHxCa2 <- lmFit(eset, designHxCa2)
fitHxCa2 <- eBayes(fitHxCa2)
resultsHxCa2 <- decideTests(fitHxCa2, adjust.method="fdr", p.value=0.25)
summary(resultsHxCa2)

# Model 3: Find genes for abnormal anatomy discovered on bronchoscopy
# specifically look for "dysplasia"
eset <- cleanNAForAnalysis(holdEset, "ANATOMYc")
eset$dysp <- numeric(sampleNumber(eset))
eset$dysp[grepl("dysp", eset$ANATOMYc, ignore.case=TRUE)] <- 1
eset$dysp <- as.factor(eset$dysp)
eset <- medianFilter(eset)
eset <- removeFactorLevel(eset, "GENDERc" ,"DK")

designDysp <- model.matrix(~ 1 + dysp + GENDERc, data=eset)
fitDysp <- lmFit(eset, designDysp)
fitDysp <- eBayes(fitDysp)
resultsDysp <- decideTests(fitDysp, adjust.method="none", p.value=0.005)
summary(resultsDysp)

o <- sort(abs(fitDysp$t[, 2]), decreasing=TRUE, index.return=TRUE)
heatmap3(exprs(eset)[o$ix[1:50], ], col = bluered,
         hclustfun=function(d) hclust(d, method="average"),
         col.clustering = "unsupervised",
         ColSideColors=cbind(smoking_colors[eset$dysp]))

heatmap3(exprs(eset)[which(resultsDysp[, 2] != 0), ], col = bluered,
         hclustfun=function(d) hclust(d, method="average"),
         col.clustering = "semisupervised",
         ColSideColors=cbind(smoking_colors[eset$dysp], cancer_colors[eset$FinalCaDXc]))

# Model 4: Look for effects of any abnormal anatomy
eset <- cleanNAForAnalysis(holdEset, "ANATOMYc")
eset$abnormal <- numeric(sampleNumber(eset))
eset$abnormal[eset$ANATOMY != 0] <- 1
eset$abnormal <- as.factor(eset$abnormal)


designAbnormal <- model.matrix(~ 1 + abnormal, data=eset)
fitAbnormal <- lmFit(eset, designAbnormal)
fitAbnormal <- eBayes(fitAbnormal)
resultsAbnormal <- decideTests(fitAbnormal, adjust.method="fdr", p.value=0.05)
summary(resultsAbnormal)

heatmap3(exprs(eset)[which(resultsAbnormal[, 2] != 0), ], col = bluered,
         hclustfun=function(d) hclust(d, method="average"),
         col.clustering = "semisupervised",
         ColSideColors=cbind(smoking_colors[eset$abnormal]))

# Model 5: corticosteroids
eset <- removeFactorLevel(holdEset, "AllegroCorticosteroidsc", "DK")
eset <- removeFactorLevel(eset, "AllegroCOPDc", "DK")
eset <- medianFilter(eset)

designCortico <- model.matrix(~ 1 + AllegroCorticosteroidsc, data=eset)
fitCortico <- eBayes(lmFit(eset, designCortico))
resultsCortico <- decideTests(fitCortico, adjust.method="fdr", p.value=0.005)
summary(resultsCortico)


eset <- calcIndicator(eset, "AllegroCorticosteroidsc", "AllegroCOPDc")
generate_heatmap(which(resultsCortico[, 2] != 0), eset, tp="indicator")
heatmap3(exprs(eset)[which(resultsCortico[, 2] != 0), ], col = bluered,
         hclustfun=function(d) hclust(d, method="average"),
         col.clustering = "unsupervised",
         ColSideColors=cbind(smoking_colors[eset$AllegroCorticosteroidsc],
                             copd_colors[eset$AllegroCOPDc]))


# Model 6: Cancer sub types and cell sub types (CA_CELL_TYPE, CA_SUB_CELL_TYPE)
eset <- holdEset
eset$CA_CELL_TYPE <- factor(eset$CA_CELL_TYPE, levels=c(levels(eset$CA_CELL_TYPE), "NoCancer"))
eset$CA_CELL_TYPE[eset$FinalCaDXc==0] <- "NoCancer"
eset <- cleanNAForAnalysis(eset, "CA_CELL_TYPE")
eset <- removeFactorLevel(eset, "CA_CELL_TYPE", "Uncertain")
eset <- removeFactorLevel(eset, "CA_CELL_TYPE", "#N/A")

eset$CA_SUB_CELL_TYPE <- factor(eset$CA_SUB_CELL_TYPE, levels=c(levels(eset$CA_SUB_CELL_TYPE), "NoCancer", "SCLC"))

eset$CA_SUB_CELL_TYPE[eset$FinalCaDXc==0] <- "NoCancer"
eset$CA_SUB_CELL_TYPE[eset$CA_CELL_TYPE=="SCLC"] <- "SCLC"
eset <- removeFactorLevel(eset, "CA_SUB_CELL_TYPE", "#N/A")
eset <- removeFactorLevel(eset, "CA_SUB_CELL_TYPE", "Uncertain")
eset$CA_SUB_CELL_TYPE <- factor(eset$CA_SUB_CELL_TYPE, levels(eset$CA_SUB_CELL_TYPE)[c(5, 1, 4, 2, 3, 6)])

designSub <- model.matrix(~ 1 + CA_SUB_CELL_TYPE, data=eset)
fitSub <- lmFit(eset, designSub)
fitSub <- eBayes(fitSub)
resultsSub <- decideTests(fitSub, adjust.method="fdr", p.value=0.05)
summary(resultsSub)

heatmap3(exprs(eset)[which(resultsSub[, 3] != 0), ], col = bluered,
         hclustfun=function(d) hclust(d, method="average"),
         col.clustering = "unsupervised",
         ColSideColors=as.numeric(eset$CA_SUB_CELL_TYPE))

```

Go back and take a look at the genes in the heatmap representing COPD vs COPD+Cancer. This actually looks pretty good.

### Model Disease with New Terms

Also look at a model including monthsSinceQuit, PY (maybe quartiles?), corticosteroids, gender, age (necessary with PY?), cancer, COPD
  * try this in COPD vs COPD+CANCER

```{r COPDwithTerms}
library(pROC)

eset <- COPDeset
eset$LENGTHSMKQUITYRc[eset$SMKc==1] <- 0
eset$LENGTHSMKQUITMOc[eset$SMKc==1] <- 0
eset <- cleanNAForAnalysis(eset, "LENGTHSMKQUITYRc")
eset <- cleanNAForAnalysis(eset, "LENGTHSMKQUITMOc")
eset <- cleanNAForAnalysis(eset, "PYc")
eset <- removeFactorLevel(eset, "AllegroCorticosteroidsc", "DK")
eset <- medianFilter(eset)
eset$QUITSMKMOcalc <- eset$LENGTHSMKQUITYRc*12 + eset$LENGTHSMKQUITMOc

eset$PYquart <- rep(4, sampleNumber(eset))
PYqs <- quantile(eset$PYc)
eset$PYquart[eset$PYc <= PYqs[4]] <- 3
eset$PYquart[eset$PYc <= PYqs[3]] <- 2
eset$PYquart[eset$PYc <= PYqs[2]] <- 1
eset$PYquart <- as.factor(eset$PYquart)

designCOPDTerms <- model.matrix(~ 1 + AllegroCorticosteroidsc + PYquart + QUITSMKMOcalc + 
                           FinalCaDXc + GENDERc, data=eset)
fitCOPDTerms <- lmFit(eset, designCOPDTerms)
fitCOPDTerms <- eBayes(fitCOPDTerms)
resultsCOPDTerms <- decideTests(fitCOPDTerms, adjust.method="none", p.value=0.05)
summary(resultsCOPDTerms)

o <- sort(abs(fitCOPDTerms$t[, 7]), decreasing=TRUE, index.return=TRUE)
heatmap3(exprs(eset)[o$ix[1:50], ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward"),
         col.clustering = "unsupervised",
         ColSideColors=cbind(cancer_colors[eset$FinalCaDXc]))
generate_heatmap(o$ix[1:50], eset, tp="indicator")

# use these genes as a predictor just to see how it works

pcasCancerInCOPD <- prcomp(znorm(exprs(eset)[o$ix[1:50], ], margin=1), center=FALSE, scale.=FALSE)
plot(roc(response=eset$FinalCaDXc, predictor=pcasCancerInCOPD$rotation[, 1]), 
         print.auc=TRUE, main="Cancer in COPD predicted by \nPC1 of top 50 Cancer genes")

```


### Nature Medicine in Allegro

Check the Nature Medicine signature in this data set. See if there is a reliable signal for cancer or if there is a difference between cancer and cancer+COPD, or between COPD and COPD+Cancer

```{r natureMedicine}
nmGenes <- read.table("Nature_Medicine_gmx_files/top_100_genes_symbols_quality_adjusted.gmx", header=FALSE, sep="", blank.lines.skip=TRUE, skip=2, fill=TRUE, as.is=TRUE)
nmGenes <- c(nmGenes[, 1], nmGenes[, 2])
nmGenes <- nmGenes[nmGenes != ""]

designBlank <- model.matrix(~ 1, data=holdEset)
fitBlank <- lmFit(holdEset, designBlank)

nmGenesInd <- match(nmGenes, fitBlank$genes[, 1])
nmGenesInd <- nmGenesInd[!is.na(nmGenesInd)]

nmEset <- holdEset[nmGenesInd, ]
eset <- removeFactorLevel(nmEset, "FinalCaDXc", "DK")
eset <- cleanNAForAnalysis(eset, "COPD2_R7")
eset <- removeFactorLevel(eset, "COPD2_R7", "DK")

eset <- calcIndicator(eset, "FinalCaDXc", "COPD2_R7")
generate_heatmap(1:featureNumber(eset), eset, tp="indicator")

designCCnm <- model.matrix(~ 1 + FinalCaDXc*COPD2_R7, data=eset)
fitCCnm <- lmFit(eset, designCCnm)
fitCCnm <- eBayes(fitCCnm)
resultsCCnm <- decideTests(fitCCnm, adjust.method="none", p.value=0.05)
summary(resultsCCnm)

generate_heatmap(which(resultsCCnm[, 2] != 0), eset, tp="indicator",
                 mn="Nature Medicine: Cancer genes p < 0.05")
generate_heatmap(which(resultsCCnm[, 3] != 0), eset, tp="indicator",
                 mn="Nature Medicine: COPD genes p < 0.05")
generate_heatmap(which(resultsCCnm[, 4] != 0), eset, tp="indicator",
                 mn="Nature Medicine: Interaction genes p < 0.05")

heatmap3(exprs(eset)[which(resultsCCnm[, 2] != 0), ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward"),
         col.clustering = "unsupervised",
         ColSideColors=cbind(cancer_colors[eset$FinalCaDXc]),
         main="Unsupervised Nature Medicine:\nCancer genes p < 0.05")

heatmap3(exprs(eset)[which(resultsCCnm[, 3] != 0), ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward"),
         col.clustering = "unsupervised",
         ColSideColors=cbind(copd_colors[eset$COPD2_R7]),
         main="Unsupervised Nature Medicine:\nCOPD genes p < 0.05")

heatmap3(exprs(eset)[which(resultsCCnm[, 4] != 0), ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward"),
         col.clustering = "unsupervised",
         ColSideColors=cbind(copdca_colors[eset$indicator]),
         main="Unsupervised Nature Medicine:\nInteraction genes p < 0.05")

heatmap3(exprs(eset), col=bluered, hclustfun=function(d) hclust(d, method="ward"),
         col.clustering="unsupervised",
         ColSideColors=cbind(copdca_colors[eset$indicator]),
         main="Unsupervised Nature Medicine:\n All genes")
```


### Partial AUC - Picking Genes For Sample Subsets

```{r pAUC}
require(genefilter)

# here try a loop going through 10 increments of the value sent to 
# rowpAUCs (i.e. 0.1:1, by 0.1); for each increment plot the ROC for the 
# maximum pAUC; then create a subset of genes with the top 10 genes from
# that run and union it with the top 10 from each of the previous runs.
# Use the subset of genes (likely > 10 genes) to run PCA and then predict
# cancer based on the PC1; also plot a heatmap of the genes
medEset <- medianFilter(holdEset)
medEset <- removeFactorLevel(medEset, "FinalCaDXc", "DK")
testpauc1 <- rowpAUCs(medEset, medEset$FinalCaDXc, 0.1)
atestpauc1 <- area(testpauc1)
o1 <- sort(atestpauc1, decreasing=TRUE, index.return=TRUE)
generate_heatmap(o$ix[1:50], medEset, tp="FinalCaDXc")


testpauc2 <- rowpAUCs(medEset, medEset$FinalCaDXc, 0.2)
atestpauc2 <- area(testpauc2)
o2 <- sort(atestpauc2, decreasing=TRUE, index.return=TRUE)
plot(testpauc2[which(atestpauc2==max(atestpauc2))])

testpauc3 <- rowpAUCs(medEset, medEset$FinalCaDXc, 0.3)
atestpauc3 <- area(testpauc3)
o3 <- sort(atestpauc3, decreasing=TRUE, index.return=TRUE)
plot(testpauc3[which(atestpauc3==max(atestpauc3))])

gSet <- medEset[union(o1$ix[1:10], union(o2$ix[1:10], o3$ix[1:10])),]
pcaGSet <- prcomp(znorm(exprs(gSet), margin=1), center=FALSE, scale.=FALSE)

plot(roc(response=medEset$FinalCaDXc, predictor=pcaGSet$rotation[, 1]),
     print.auc=TRUE,
     main="Cancer predicted by PC1 from top Cancer pAUC genes")

generate_heatmap(1:18, gSet, tp="FinalCaDXc")
boxplot(pcaGSet$rotation[, 1] ~ FinalCaDXc, data=medEset)

# should try this in:
#   1. cancer vs healthy
#   2. COPD vs healthy
#   3. cancer+COPD vs healthy
#   4. cancer vs COPD
#   5. cancer vs cancer+COPD
#   6. COPD vs cancer+COPD
```

### Do pairwise comparisons between groups
```{r pairwiseComparisons}
# MODEL 1
# try removing the COPD group and collapsing the cancer and healthy groups into one group, then using the differentially expressed genes between COPD+Cancer and (Cancer and Healthy) to predict group membership
eset <- holdEset
eset <- removeFactorLevel(eset, "FinalCaDXc", "DK")
eset <- cleanNAForAnalysis(eset, "COPD2_R7")
eset <- removeFactorLevel(eset, "COPD2_R7", "DK")
eset <- calcIndicator(eset, "FinalCaDXc", "COPD2_R7")
eset <- removeFactorLevel(eset, "indicator", "3")
eset$indicator[eset$indicator==2] <- 1
eset <- removeFactorLevel(eset, "indicator", "2")

eset <- medianFilter(eset)

designPW1 <- model.matrix(~ 1 + SMKc + indicator + AGEcalc, data=eset)
fitPW1 <- lmFit(eset, designPW1)
fitPW1 <- eBayes(fitPW1)
resultsPW1 <- decideTests(fitPW1, adjust.method="fdr", p.value=0.05, lfc=.25)
summary(resultsPW1)

generate_heatmap(which(resultsPW1[, 3] != 0), eset, tp="indicator")

require(znorm)
require(pROC)
pcaCancerInCOPD1 <- prcomp(znorm(exprs(eset)[which(resultsPW1[, 3] != 0), ], margin=1), center=FALSE, scale.=FALSE)
plot(roc(response=eset$indicator, predictor=pcaCancerInCOPD1$rotation[, 1]), 
         print.auc=TRUE, main="Cancer in COPD\npredicted by PC1")


##################################################################
##################################################################
##################################################################
#MODEL 2
#try the same as above but without removing the COPD group
# in this case, put together the COPD, Cancer, and healthy groups
eset <- holdEset
eset <- removeFactorLevel(eset, "FinalCaDXc", "DK")
eset <- cleanNAForAnalysis(eset, "COPD2_R7")
eset <- removeFactorLevel(eset, "COPD2_R7", "DK")
eset <- calcIndicator(eset, "FinalCaDXc", "COPD2_R7")
eset$indicator[eset$indicator==3] <- 1
eset$indicator[eset$indicator==2] <- 1
eset <- removeFactorLevel(eset, "indicator", "3")
eset <- removeFactorLevel(eset, "indicator", "2")

eset <- medianFilter(eset)

designPW2 <- model.matrix(~ 1 + SMKc + indicator, data=eset)
fitPW2 <- lmFit(eset, designPW2)
fitPW2 <- eBayes(fitPW2)
resultsPW2 <- decideTests(fitPW2, adjust.method="fdr", p.value=0.05, lfc=.25)
summary(resultsPW2)

generate_heatmap(which(resultsPW2[, 3] != 0), eset, tp="indicator")

require(znorm)
require(pROC)
pcaCancerInCOPD2 <- prcomp(znorm(exprs(eset)[which(resultsPW2[, 3] != 0), ], margin=1), center=FALSE, scale.=FALSE)
plot(roc(response=eset$indicator, predictor=pcaCancerInCOPD2$rotation[, 1]), 
         print.auc=TRUE, main="Cancer in COPD\npredicted by PC1")

##################################################################
##################################################################
##################################################################
# MODEL 3
#try the same as above but this time removing COPD and Cancer only groups
# in this case, comparing the COPD+Cancer to the disease free group only
eset <- holdEset
eset <- removeFactorLevel(eset, "FinalCaDXc", "DK")
eset <- cleanNAForAnalysis(eset, "COPD2_R7")
eset <- removeFactorLevel(eset, "COPD2_R7", "DK")
eset <- calcIndicator(eset, "FinalCaDXc", "COPD2_R7")
tempEset <- eset
eset <- removeFactorLevel(eset, "indicator", "3")
eset <- removeFactorLevel(eset, "indicator", "2")

eset <- medianFilter(eset)

designPW3 <- model.matrix(~ 1 + SMKc + indicator, data=eset)
fitPW3 <- lmFit(eset, designPW3)
fitPW3 <- eBayes(fitPW3)
resultsPW3 <- decideTests(fitPW3, adjust.method="fdr", p.value=0.05, lfc=.25)
summary(resultsPW3)

generate_heatmap(which(resultsPW3[, 3] != 0), eset, tp="indicator")

require(znorm)
require(pROC)
pcaCancerInCOPD3 <- prcomp(znorm(exprs(eset)[which(resultsPW3[, 3] != 0), ], margin=1), center=FALSE, scale.=FALSE)
plot(roc(response=eset$indicator, predictor=pcaCancerInCOPD3$rotation[, 1]), 
         print.auc=TRUE, main="Cancer in COPD\npredicted by PC1")

# try these genes, which so differentiate disease free from COPD+Cancer, 
# in all patients

tempEset$indicator[tempEset$indicator==2] <- 1
tempEset$indicator[tempEset$indicator==3] <- 1
tempEset <- removeFactorLevel(tempEset, "indicator", "2")
tempEset <- removeFactorLevel(tempEset, "indicator", "3")

pcaCancerInCOPD4 <- prcomp(znorm(exprs(tempEset)[which(resultsPW3[, 3] != 0), ],
                                 margin=1), center=FALSE, scale.=FALSE)
plot(roc(response=tempEset$indicator,
         predictor=pcaCancerInCOPD4$rotation[, 1]), 
     print.auc=TRUE,
     main="Cancer in COPD\npredicted by PC1")


heatmap3(exprs(holdEset)[smokingIDs, ], col=bluered,
hclustfun=function(d) hclust(d, method="ward"),
col.clustering="unsupervised", ColSideColors=smoking_colors[holdEset$SMKc])

```

