Testing R Markdown with Allegro Analysis
========================================================
2014-06-04
-------------------------------------------
This is my first R Mardown document. I will be attempting to run and create a simple analyses
for the Allegro CODP+Cancer project. 

When you click the **Knit HTML** button a web page will be generated that includes both content as well as the output of any embedded R code chunks within the document.

```{r Running one model: BATCH + RIN + AGEcalc}
# AllegroSetup.R uses an RDS file with 800+ samples (biological replicates) and without QC
source("/protected/projects/pulmarray/Allegro/COPD_Cancer/scripts/AllegroSetup.R")

# I can put my analysis here:

# Define the covariate to use
covars <- list("SMKc", "COPD_EM_Bronchitisc", "FinalCaDXc", "FinalCaDXc*COPD_EM_Bronchitisc",
               "SMKc*COPD_EM_Bronchitisc", "SMKc*FinalCaDXc")

exprData <- removeFactorLevel(exprData, "COPD_EM_Bronchitisc", "DK")
exprData <- removeFactorLevel(exprData, "FinalCaDXc", "DK")

# define an indicator/group factor in the phenotype data (values 1-4, healthy, copd, cancer, both)
exprData <- calcIndicator(exprData, "COPD_EM_Bronchitisc")

exprDataF.mncv <- removeFactorLevel(exprDataFiltered$mn.cv, "COPD_EM_Bronchitisc", "DK")
exprDataF.mncv <- removeFactorLevel(exprDataF.mncv, "FinalCaDXc", "DK")
exprDataF.mncv <- calcIndicator(exprDataF.mncv, "COPD_EM_Bronchitisc")

exprDataF.cv <- removeFactorLevel(exprDataFiltered$cv, "COPD_EM_Bronchitisc", "DK")
exprDataF.cv <- removeFactorLevel(exprDataF.cv, "FinalCaDXc", "DK")
exprDataF.cv <- calcIndicator(exprDataF.cv, "COPD_EM_Bronchitisc")

test1 <- analyze_covariate_lists(exprData, covars)

testcm <- analyze_covariate_lists(exprDataF.mncv, covars)
indsCM <- testcm$cutoffs_p[[2]]$'COPD_EM_Bronchitisc1:FinalCaDXc1'

testc <- analyze_covariate_lists(exprDataF.cv, covars)



```

A heatmap representing the genes significant for COPD*Cancer from a model that also includes
smoking status (former or current):

```{r TestFigure, fig.width=7, fig.height=6}
# define the indices of the genes of interest to print in a heatmap

# there are 480 genes that survive FDR correction at FDR < 0.2
inds2 <- test1$cutoffs_fdr[[2]]$'COPD_EM_Bronchitisc1:FinalCaDXc1'

# there are 127 genes that survive FDR correction at FDR < 0.1
inds3 <- test1$cutoffs_fdr[[3]]$'COPD_EM_Bronchitisc1:FinalCaDXc1'


clabels <- cbind("Group"=copdca_colors[exprDataF.mncv$indicator],
                 "Cancer"=cancer_colors[exprDataF.mncv$FinalCaDXc],
                 "COPD"=copd_colors[exprDataF.mncv$COPD_EM_Bronchitisc])

#heatmap3(exprs(exprData)[inds,], col=bluered, colSideColors=clabels, col.clustering="semisupervised")
#heatmap3(exprs(exprData)[inds2,], col=bluered, ColSideColors=clabels, col.clustering=exprData$indicator, main=paste(length(inds2), "genes at FDR < 0.2 for COPD*Cancer interaction"))

#heatmap3(exprs(exprData)[inds3,], col=bluered, ColSideColors=clabels, col.clustering=exprData$indicator, main=paste(length(inds3), "genes at FDR < 0.1 for COPD*Cancer interaction"))

heatmap3(exprs(exprDataF.mncv)[indsCM,], col=bluered, ColSideColors=clabels, col.clustering=exprDataF.mncv$indicator, main = paste(length(indsCM), "genes at p<0.01 for COPD*Cancer interaction"))



```


### I'd like to try running this with just patients with PFTs

```{r Running model on only patients with PFT data available}

# find the samples with PFT information
pftInds <- which(!is.na(exprData$RATIOc))
# subset the dataset to include only those patients with PFTs
exprDataPFT <- exprData[,pftInds]

# calculate/create an indicator column
exprDataPFT <- calcIndicator(exprDataPFT, "COPD2_R7")

# remove the NAs from COPD2_R7
exprDataPFT <- cleanNAForAnalysis(exprDataPFT, "COPD2_R7")

# remove the DKs from COPD2_R7
exprDataPFT <- removeFactorLevel(exprDataPFT, "COPD2_R7", "DK")

covarsR7 <- list("SMKc", "COPD2_R7", "FinalCaDXc", "COPD2_R7*FinalCaDXc", "SMKc*COPD2_R7", 
                 "SMKc*FinalCaDXc")
testPFT1 <- analyze_covariate_lists(exprDataPFT, covarsR7)

indsIntr <- testPFT$cutoffs_fdr[[3]]$'COPD2_R71:FinalCaDXc1'

clabels <- cbind("Group"=copdca_colors[exprDataPFT$indicator],
                 "Cancer"=cancer_colors[exprDataPFT$FinalCaDXc],
                 "COPD"=copd_colors[exprDataPFT$COPD2_R7])
```

```{r PFT patients, fig.width=7, fig.height=6}
heatmap3(exprs(exprDataPFT)[indsIntr,], col=bluered, ColSideColors=clabels, col.clustering=exprDataPFT$indicator, main=paste(length(indsIntr), "genes at FDR < 0.1 for COPDR7*Cancer interaction"))

```

I'd also like to try running this by including only those patients with severe COPD
i.e. decreasing the cutoffs by which we define COPD

