Re-Expanding the COPD definition for In-Person Meeting
========================================================
2014-07-15
---------------------------------------------

Avi and Marc are at the NCI in-person meeting and they've asked me to do several things. Marc had asked that I run some simulation data through my code to make sure that nothing was broken. That's been done and is in 2014-07-13. Now Avi has asked that I re-expand the COPD term back to the Allegro term and try a model that compares COPD+Cancer to everyone else, like what I've been doing in the slimmed down, PFT-only data set, to see if we can more-or-less wipe out the COPD-dominant effect. 

I'll also examine PCA and color the COPD patients to check if there are any odd patterns so that we may remove any appropriate outliers. 

```{r GlobalVariables, include=FALSE}
cacheOption = FALSE

```


```{r defineFunctions, include=FALSE}
foldChange <- function(eset, groups, genes){
  apply(exprs(eset)[genes, groups==0], 1, mean) - 
    apply(exprs(eset)[genes, groups==1], 1, mean)
}

medianFilter <- function(eset){
  md <- median(exprs(eset))
  passFilter <- logical(featureNumber(eset))
  for(i in 1:featureNumber(eset)){
    passFilter[i] <- sum(exprs(eset)[i, ] > md) > 0
  }
  
  eset <- eset[passFilter, ]
  return(eset)
  
}

```


```{r LoadData, echo=FALSE, cache=cacheOption, include=FALSE}
# set the working directory to the directory of this script
setwd("/restricted/projectnb/pulmarray/LinGA_protected/Allegro/COPD_Cancer/experiments/2014-07-15")

# load the data using a script that removes patients with:
#  Cancer = NA
#  SMK    = 3
source("/protected/projects/pulmarray/Allegro/COPD_Cancer/scripts/AllegroSetup.R")

# remove the patients who have a DK for Cancer status as this is the a
# phenotype of interest (along with COPD)
eset <- removeFactorLevel(eset, "FinalCaDXc", "DK")
eset <- removeFactorLevel(eset, "AllegroCOPDc", "DK")
#eset <- cleanNAForAnalysis(eset, "COPD2_R7")
#eset <- removeFactorLevel(eset, "COPD2_R7", "DK")
eset <- cleanNAForAnalysis(eset, "AGEcalc")
eset <- removeBioReps(exprData=eset)
eset <- calcIndicator(eset, "FinalCaDXc", "AllegroCOPDc")
#eset <- removeFactorLevel(eset, "GENDERc", "DK")
#eset <- calcIndicator(eset, "FinalCaDXc", "COPD2_R7")

library(znorm)

ind1 <- sum(eset$indicator==1)
ind2 <- sum(eset$indicator==2)
ind3 <- sum(eset$indicator==3)
ind4 <- sum(eset$indicator==4)

esetC <- eset[, eset$SMKc==1]
esetF <- eset[, eset$SMKc==2]
esetPY <- cleanNAForAnalysis(eset, "PYc")

# generate the values to put in the table
ns <- summary(eset$indicator)
# Age mean and standard deviation
ageM <- round(tapply(eset$AGEcalc, eset$indicator, mean))
ageS <- round(tapply(eset$AGEcalc, eset$indicator, sd))
ageP <- paste(ageM, ageS, sep="+/-")

# Pack years mean and standard deviation
pyM <- round(tapply(eset$PYc, eset$indicator, mean, na.rm=TRUE))
pyS <- round(tapply(eset$PYc, eset$indicator, sd, na.rm=TRUE))
pyP <- paste(pyM, pyS, sep="+/-")

# Smoking status breakdown
smkM <- tapply(eset$SMKc, eset$indicator, summary)

# Cancer subtype breakdown (i.e. NSCLC vs SCLC)
casubM <- tapply(eset$CA_CELL_TYPE, eset$indicator, summary)

# FEV1% mean and standard deviation
fev1pM <- round(tapply(eset$FEV1Pc, eset$indicator, mean),2)
fev1pS <- round(tapply(eset$FEV1Pc, eset$indicator, sd),2)
fev1pP <- paste(fev1pM, fev1pS, sep="+/-")

# FEV1/FVC mean and standard deviation
ratioM <- round(tapply(eset$RATIOc, eset$indicator, mean, na.rm=TRUE), 2)
ratioS <- round(tapply(eset$RATIOc, eset$indicator, sd, na.rm=TRUE), 2)
ratioP <- paste(ratioM, ratioS, sep="+/-")

```

                |  Disease Free  |   Cancer        |    COPD         |  Cancer+COPD
--------------- |:--------------:|:---------------:|:---------------:|:------------:
   **n**        |  `r ns[1]`     |   `r ns[2]`     |   `r ns[3]`     |  `r ns[4]`
   **Age**      |  `r ageP[1]`   |    `r ageP[2]`  |   `r ageP[3]`   |  `r ageP[4]`
   **Pack Years** | `r pyP[1]`    |   `r pyP[2]`    |   `r pyP[3]`    |  `r pyP[4]`
   **Smoking(C,F)** | `r smkM[[1]]` | `r smkM[[2]]`    | `r smkM[[3]]`  | `r smkM[[4]]`
   **Subtype(NSCLC, SCLC)**  | NA | `r casubM[[2]][2:3]` | NA  | `r casubM[[4]][2:3]`
   **FEV1%**   |  `r fev1pP[1]` |  `r fev1pP[2]`  |  `r fev1pP[3]`   |  `r fev1pP[4]`
   **FEV1/FVC**   |  `r ratioP[1]` |  `r ratioP[2]`  |  `r ratioP[3]`   |  `r ratioP[4]`
Table 1: demographics for samples with available PFT data. COPD here is the given Allegro term (not PFT based). 



### GOAL 1: Use Allegro COPD Term in COPD+Cancer vs Everything Analysis
#### Specifically this will be in the former smokers (exlude currents)
```{r allegroAllOrNone, cache=cacheOption, fig.width=8, fig.height=7, fig.align='center'}


holdEset <- eset
eset <- medianFilter(esetF)
# create a binary factor that indicates COPD+Cancer or not
cc <- numeric(sampleNumber(eset))
cc[eset$indicator==4] <- 1
eset$cc <- factor(cc)

# create and fit the all-or-none model
designAllOrNone <- model.matrix(~ 1 + cc, data=eset)

fitAllOrNone <- lmFit(eset, designAllOrNone)
fitAllOrNone <- eBayes(fitAllOrNone)

resultsAllOrNone <- decideTests(fitAllOrNone, adjust.method="fdr", p.value=0.05)
summary(resultsAllOrNone)

temp <- eset
exprs(temp) <- znorm(exprs(temp), margin=1)

# plot the top 50  by t-value
o <- sort(abs(fitAllOrNone$t[, 2]), decreasing=TRUE, index.return=TRUE)
inds <- o$ix[1:50]

#pdf("top50SMKFormerfinal.pdf")
generate_heatmap(inds, temp, tp="indicator")
#dev.off()


```


### GOAL 1A: Make PCA plots for genes of interest; plot metagene
```{r secondaryImages, cache=cacheOption}

pca <- prcomp(exprs(temp)[inds, ], scale.=FALSE, center=FALSE, )
pdf("pcaPlot.pdf")
plot(pca$rotation[, 1], pca$rotation[, 2], xlab="PC1", ylab="PC2",
     col=copdca_colors,
     main="PCA Plot for Top 50 Genes",)
dev.off()

pdf("metagenePlot.pdf")
boxplot(pca$rotation[eset$indicator==1, 1],
        pca$rotation[eset$indicator==2, 1],
        pca$rotation[eset$indicator==3, 1], 
        pca$rotation[eset$indicator==4, 1],
        col=copdca_colors, names=c("DiseaseFree", "Cancer" , "COPD", "Cancer+COPD"),
        main="Metagene of Top 50 Genes")
dev.off()
```

### GOAL 1B: Generate boxplots of the top 50 Former smoker genes
```{r formerSmoker50Boxplots, inclue=FALSE}

pdf("top50FormerBoxplots.pdf")
for(i in 1:length(inds)){
  ind <- inds[i]
 
  gn <- paste(fitAllOrNone$genes[ind,1], ":", fitAllOrNone$genes[ind,2],  sep="")
 
  boxplot(exprs(temp[ind, ])[1, ] ~ temp$indicator,
          names=c("Disease Free", "Cancer", "COPD", "Cancer+COPD"),
          col=c("green", "red", "blue", "yellow"),
          xlab="Group", ylab="Expression",
          main=gn)
}
dev.off()


# output the gene symbols from the Former smoker analysis
genes <- data.frame(fitAllOrNone$genes[inds, 1])
colnames(genes) <- "GeneSymbols"

write.table(genes, file="FormerSmoker50GeneSymbols", sep=",", row.names=FALSE, quote=FALSE)



```

### GOAL 2: Use Allegro COPD Term in Interaction Analysis
```{r allegroInteraction, cache=cacheOption, fig.width=8, fig.height=7, fig.align='center'}
# create and fit the interaction model
designIntrx <- model.matrix(~ 1 + AllegroCOPDc*FinalCaDXc, data=eset)

fitIntrx <- lmFit(eset, designIntrx)
fitIntrx <- eBayes(fitIntrx)

resultsIntrx <- decideTests(fitIntrx, adjust.method="fdr", p.value=0.1)
summary(resultsIntrx)

inds <- which(resultsIntrx[, 4] != 0)
generate_heatmap(inds, temp, tp="indicator")

```


