In this script I am going to attemt to summarize as much of the work I have done as is possible to pull into one place. See headers below for general categories of analyses. 
  
```{r defineFunctions, include=FALSE}
foldChange <- function(eset, groups, genes){
  apply(exprs(eset)[genes, groups==0], 1, mean) - 
    apply(exprs(eset)[genes, groups==1], 1, mean)
}

medianFilter <- function(eset){
  md <- median(exprs(eset))
  passFilter <- logical(featureNumber(eset))
  for(i in 1:featureNumber(eset)){
    passFilter[i] <- sum(exprs(eset)[i, ] > md) > 0
  }
  
  eset <- eset[passFilter, ]
  return(eset)
  
}

saveGeneList <- function(analysis, filename=paste(Sys.Date(), "Temp", analysis$name, ".txt", sep="")){
  
  write.table(analysis$geneSymbols, file=filename, quote=FALSE,row.names=FALSE, col.names=FALSE)

}

saveRankedList <- function(fit, varColInd, filename = "tempRankedList"){
  filename = paste(paste(Sys.Date(), colnames(fit$t)[varColInd], filename, sep="_"), "rnk", sep=".")
  print(filename)
  temp <- data.frame(geneSymbols = fit$genes$Symbol, ts = fit$t[, varColInd])
  write.table(temp, file=filename, row.names=FALSE, quote=FALSE, col.names=TRUE, sep="\t", eol="\n")
  
  
}

saveGeneSet <- function(fit, inds, filename = "tempGeneSet", geneSetName="GeneSet", description=""){
  filename = paste(paste(Sys.Date(), filename, sep="_"), "gmx", sep=".")
  print(filename)
  temp <- data.frame(geneSymbols=c(geneSetName, description, fit$genes$Symbol[inds]))
  write.table(temp, file=filename, row.names=FALSE, quote=FALSE, col.names=FALSE)
  
}

# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)
  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)
  numPlots = length(plots)
  
  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                     ncol = cols, nrow = ceiling(numPlots/cols))
    }
  
  if (numPlots==1) {
    print(plots[[1]])
    } else {
      # Set up the page
      grid.newpage()
      pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
      
      # Make each plot, in the correct location
      for (i in 1:numPlots) {
        # Get the i,j matrix positions of the regions that contain this subplot
        matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
        print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                        layout.pos.col = matchidx$col))
        }
      }
  }

genFileName <- function(filename){
  
  return(paste(Sys.Date(), filename, sep="_"))
}
```

### Setup
```{r setup, eval=TRUE, echo=TRUE, include=FALSE, results='hide'}

setwd("/protected/projects/pulmarray/Allegro/COPD_Cancer/experiments/Expmt_0043_SILCC_Summary/")
source("/protected/projects/pulmarray/Allegro/COPD_Cancer/scripts/AllegroSetup.R")
#source("../2015-05-03_CBM_ATS_2015/plotGSEA.R")
# fix the one patient with wonky data
# eventually this should just be saved in the RDS file
holdEset <- eset
holdEset$FEV1Pc[holdEset$FEV1Pc==89.2] <- 0.892

emt_genes <- read.csv("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/emt_gene_symbols_byers_2013.csv", header=FALSE)
temp <- as.factor(unique(as.character(emt_genes$V1)))
emt_genes <- data.frame("EMT_Genes" = temp)

GOLD_colors = c("0" = "gray100", "1" = "gray75", "2" = "gray50", "3" = "gray25", "4" = "black")
greyblack <- colorRampPalette(c("black", "white", "grey"))(256)


# load data, signatures and gene sets
atf4_data <- readRDS("/protected/projects/pulmarray/Allegro/ATF4/ATF4data_2015MAR17_ExpressionSet.rds")

lam_sign <- read.delim("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/katie98genesEntrez.txt")
lam_sign[, 3] <- paste(lam_sign[, 2], "_at", sep="")

lamGMT <- read.csv("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/100525_lam_98overlapgenes_covars.gmt", sep="\t", as.is=TRUE, header=FALSE)
lamUP <- as.character(lamGMT[1, 3:56])
lamDN <- as.character(lamGMT[2, 3:46])

joshLM127dn <- read.csv("/protected/projects/pulmarray/HoggCOPD/A1ATGrifols/experiments/Expmt_0001_LM_Parenchyma/2015-06-08_Joshdn127.gmx", skip=1, header=TRUE)

joshLM127up <- read.csv("/protected/projects/pulmarray/HoggCOPD/A1ATGrifols/experiments/Expmt_0001_LM_Parenchyma/2015-06-08_Joshup127.gmx", skip=1, header=TRUE)

# from parenchymal REAL data
cleLMup <- read.csv("/protected/projects/pulmarray/HoggCOPD/A1ATGrifols/experiments/Expmt_0003_LM_Parenchyma_CLE/2015-06-24_parenchyma_cle_geneset_cluster2_fdr_0.05.gmx")

cleLMdn <- read.csv("/protected/projects/pulmarray/HoggCOPD/A1ATGrifols/experiments/Expmt_0003_LM_Parenchyma_CLE/2015-06-24_parenchyma_cle_geneset_cluster1_fdr_0.05.gmx")

conLMdn <- read.csv("/protected/projects/pulmarray/HoggCOPD/A1ATGrifols/experiments/Expmt_0005_LM_Parenchyma_Control/2015-06-30_parenchyma_control_geneset_cluster1_fdr_0.15.gmx")

conLMup <- read.csv("/protected/projects/pulmarray/HoggCOPD/A1ATGrifols/experiments/Expmt_0005_LM_Parenchyma_Control/2015-06-30_parenchyma_control_geneset_cluster2_fdr_0.15.gmx")

natMed <- read.csv("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/Nature_Medicine_gmx_files/top_100_genes_symbols_quality_adjusted.gmx", sep="\t", as.is=TRUE)
natMedUp <- natMed[2:65, 1]
natMedDn <- natMed[2:37, 2]

tgfB <- read.csv("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/msigdb_C2_CGP_genesets/BIOCARTA_TGFB_PATHWAY.gmx", skip=1)
tgfB <- as.character(tgfB$X..TGF.beta.signaling.pathway)

erStress <- as.character(read.csv("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/er.entrez.txt", header=FALSE)$V1)

cilioGen <- as.character(read.csv("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/cil.entrez.txt", header=FALSE)$V1)

#atf4_targets <- read.delim("ATF4_Targets_Enrichr_TRANSFACJASPAR_270Genes.txt", sep=",", as.is=TRUE, header=FALSE)
#atf4_targets <- as.character(atf4_targets)

#a4LEup <- read.delim("ATF4_targets_cluster1_270Genes_LeadingEdge_Up.csv", header=FALSE)
#a4LEup <- as.character(a4LEup$V1)

#a4LEdn <- read.delim("ATF4_targets_cluster2_270Genes_LeadingEdge_Down.csv", header=FALSE)
#a4LEdn <- as.character(a4LEdn$V1)
library(znorm)
library(diptest)
library(GSVA)
library(annotate)
library(org.Hs.eg.db)
library(corrplot)
library(ggplot2)
library(beeswarm)
library(ASSIGN)

### Analysis Variables
SAVE=FALSE
```

### Data Cleaning
```{r cleanup}

# Remaining question - should we keep or remove the COPD GOLD Status 1 patients? i.e. mild disease
# What about patients using or their charts indicating they have used corticosteroids?
esetClean <- holdEset
esetClean <- cleanNAForAnalysis(esetClean, "COPD2_R7")
esetClean <- cleanNAForAnalysis(esetClean, "RATIOc")
esetClean <- cleanNAForAnalysis(esetClean, "PYc")
esetClean <- cleanNAForAnalysis(esetClean, "RIN")
esetClean <- removeFactorLevel(esetClean, "COPD2_R7", "DK")
esetClean <- removeFactorLevel(esetClean, "FinalCaDXc", "DK")
esetClean <- removeFactorLevel(esetClean, "COPD2_R7", "DK")
esetClean <- removeFactorLevel(esetClean, "GENDERc", "DK")

esetClean <- calcIndicator(esetClean, "FinalCaDXc", "COPD2_R7")
esetClean$smkindic <- as.numeric(as.character(esetClean$indicator))
esetClean$smkindic[esetClean$SMKc==2] <- esetClean$smkindic[esetClean$SMKc==2] + 4
esetClean$smkindic <- as.factor(esetClean$smkindic)

# Define the GOLD Stage appropriate for all patients
esetClean$GOLD <- factor(x=rep(0, sampleNumber(esetClean)), levels=c("0","1","2","3","4"))
# GOLD Stage 1 Ratio < 0.7, FEV1 >= 8
esetClean$GOLD[esetClean$RATIOc < 0.7 & esetClean$FEV1Pc >= 0.8] <- 1
# GOLD Stage 2 Ratio < 0.7, 50 <= FEV1 < 80
esetClean$GOLD[esetClean$RATIOc < 0.7 & esetClean$FEV1Pc < 0.8] <- 2
# GOLD Stage 3 Ratio < 0.7, 30 <= FEV1 <= 50
esetClean$GOLD[esetClean$RATIOc < 0.7 & esetClean$FEV1Pc < 0.5] <- 3
# GOLD Stage 4 Ratio < 0.7, FEV1 < 30%
esetClean$GOLD[esetClean$RATIOc < 0.7 & esetClean$FEV1Pc < 0.3] <- 4

esetClean <- removeBioReps(esetClean)
esetCleanFormers <- removeFactorLevel(esetClean, "SMKc", "1")
esetCleanCurrents <- removeFactorLevel(esetClean, "SMKc", "2")
# Generate a super clean dataset devoid of all patients with other pulmonary diseases
esetSuperClean <- esetClean
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroPNAc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroTBc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroSarcoidc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroIPFc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroIPFc", "DK")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroAtelectasisc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroAtelectasisc", "DK")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroAsthmac", "DK")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroAsthmac", "1")

esetSuperCleanICS <- removeFactorLevel(esetSuperClean, "AllegroCorticosteroidsc", "1")
esetSuperCleanICS <- removeFactorLevel(esetSuperCleanICS, "AllegroCorticosteroidsc", "DK")

# clean up ATF4 data
varLabels(atf4_data) <- "SampleID"
atf4_data$SampleID[grep("ATF4", atf4_data$SampleID)] <- "ATF4"
atf4_data$SampleID[grep("Ctrl", atf4_data$SampleID)] <- "Control"
atf4_data$SampleID <- as.factor(atf4_data$SampleID)

atf4_colors <- c("ATF4" = "red", "Control" = "black")


```

### ATF4 Models
```{r ATF4models}
# generate ATF4 model(s)
atf4Model <- lmFitWrapper(atf4_data, c("SampleID"),
                          name="ATF4 Perturbation Data", adjust.method="none",
                          p.value=0.05, varOfInterest=1)

clabels <- cbind(atf4_colors[as.character(atf4_data$SampleID)])
heatmap3(exprs(atf4_data)[atf4Model$inds, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = atf4_data$SampleID, ColSideColors = clabels,
         main = "ATF4 Genes",
         labCol="")

hmapClusts <- heatmap3(exprs(atf4_data)[atf4Model$inds,], col = bluered,
                       keep.dendro=TRUE,
                       hclustfun=function(d) hclust(d, method="ward.D"),
                       col.clustering = atf4_data$SampleID, ColSideColors=clabels,
                       main="ATF4 Genes", labCol="")

clusters <- cutree(as.hclust(hmapClusts$Rowv), k=2)
atf4Cluster1_dn <- atf4Model$geneSymbols[clusters==1]
atf4Cluster2_up <- atf4Model$geneSymbols[clusters==2]

atf4dn <- names(clusters)[clusters==1]
atf4up <- names(clusters)[clusters==2]
atf4 <- names(clusters)

rlabels <- cbind(smoking_colors[clusters])
heatmap3(exprs(atf4_data)[atf4Model$inds,], col = bluered,
         keep.dendro=TRUE,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = atf4_data$SampleID, ColSideColors=clabels,
         main="ATF4 Genes", labCol="", RowSideColors=rlabels)



```

## Interaction Models
### Among All Patients  
```{r intrxAlla}
target <- medianFilter(esetClean)
tName = "intrxAlla"
tempModel <- lmFitWrapper(target, c("COPD2_R7", "FinalCaDXc", "SMKc", "AGEcalc", "RIN", "GENDERc", "COPD2_R7*FinalCaDXc"), 7, "fdr", 0.15, "Interaction Model 1 Among All Patients")

clabels <- cbind(copdca_colors[target$indicator], smoking_colors[as.character(target$SMKc)])
tempHMap <- heatmap3(exprs(target)[tempModel$inds, ], col=bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering=target$indicator, ColSideColors=clabels,
         main=tempModel$hMapName, labCol="", keep.dendro=TRUE)

if(SAVE){
  tempClusters <- cutree(as.hclust(tempHMap$Rowv), k=2)

  # save entire gene list
  write.table(tempModel$geneSymbols, file=genFileName(paste(tName, ".txt", sep="")), quote=FALSE, row.names=FALSE, col.names=FALSE)
  # save cluster 1 gene list
  write.table(tempModel$geneSymbols[tempClusters==1], file=genFileName(paste(tName, "c1.txt", sep="")))
  
  # save cluster 2 gene list
  write.table(tempModel$geneSymbols[tempClusters==2], file=genFileName(paste(tName, "c2.txt", sep="")))

}

intrxAllaResults <- list(model=tempModel, heatmap=tempHMap)
```

### Among All Patients + ICS
```{r intrxAllb}
target <- medianFilter(removeFactorLevel(esetClean, "AllegroCorticosteroidsc", "DK"))
tName = "intrxAllb"
tempModel <- lmFitWrapper(target, c("COPD2_R7", "FinalCaDXc", "SMKc", "AGEcalc", "RIN", "GENDERc", "AllegroCorticosteroidsc", "COPD2_R7*FinalCaDXc"), 8, "fdr", 0.2, "Interaction Model 2 +ICS Among All Patients")

clabels <- cbind(copdca_colors[target$indicator], smoking_colors[as.character(target$SMKc)])
tempHMap <- heatmap3(exprs(target)[tempModel$inds, ], col=bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering=target$indicator, ColSideColors=clabels,
         main=tempModel$hMapName, labCol="", keep.dendro=TRUE)


if(SAVE){
  tempClusters <- cutree(as.hclust(tempHMap$Rowv), k=2)

  # save entire gene list
  write.table(tempModel$geneSymbols, file=genFileName(paste(tName, ".txt", sep="")), quote=FALSE, row.names=FALSE, col.names=FALSE)
  # save cluster 1 gene list
  write.table(tempModel$geneSymbols[tempClusters==1], file=genFileName(paste(tName, "c1.txt", sep="")))
  
  # save cluster 2 gene list
  write.table(tempModel$geneSymbols[tempClusters==2], file=genFileName(paste(tName, "c2.txt", sep="")))

}

intrxAllbResults <- list(model=tempModel, heatmap=tempHMap)
```


### Among Former Smokers  
```{r intrxFormera}
target <- medianFilter(esetCleanFormers)
tName = "intrxFormera"
tempModel <- lmFitWrapper(target, c("COPD2_R7", "FinalCaDXc", "AGEcalc", "RIN", "GENDERc", "COPD2_R7*FinalCaDXc"), 6, "fdr", 0.3, "Interaction Model 1 Among Former Smokers")

clabels <- cbind(copdca_colors[target$indicator], smoking_colors[as.character(target$SMKc)])
tempHMap <- heatmap3(exprs(target)[tempModel$inds, ], col=bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering=target$indicator, ColSideColors=clabels,
         main=tempModel$hMapName, labCol="", keep.dendro=TRUE)

if(SAVE){
  tempClusters <- cutree(as.hclust(tempHMap$Rowv), k=2)

  # save entire gene list
  write.table(tempModel$geneSymbols, file=genFileName(paste(tName, ".txt", sep="")), quote=FALSE, row.names=FALSE, col.names=FALSE)
  # save cluster 1 gene list
  write.table(tempModel$geneSymbols[tempClusters==1], file=genFileName(paste(tName, "c1.txt", sep="")))
  
  # save cluster 2 gene list
  write.table(tempModel$geneSymbols[tempClusters==2], file=genFileName(paste(tName, "c2.txt", sep="")))

}

intrxFormeraResults <- list(model=tempModel, heatmap=tempHMap)
```

### Among Former Smokers + ICS
```{r intrxFormerb}
target <- medianFilter(removeFactorLevel(esetCleanFormers, "AllegroCorticosteroidsc", "DK"))
tName = "intrxFormerb"
tempModel <- lmFitWrapper(target, c("COPD2_R7", "FinalCaDXc", "AGEcalc", "RIN", "GENDERc", "AllegroCorticosteroidsc", "COPD2_R7*FinalCaDXc"), 7, "none", 0.01, "Interaction Model 2 +ICS Among Former Smokers")

clabels <- cbind(copdca_colors[target$indicator], smoking_colors[as.character(target$SMKc)])
tempHMap <- heatmap3(exprs(target)[tempModel$inds, ], col=bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering=target$indicator, ColSideColors=clabels,
         main=tempModel$hMapName, labCol="", keep.dendro=TRUE)


if(SAVE){
  tempClusters <- cutree(as.hclust(tempHMap$Rowv), k=2)

  # save entire gene list
  write.table(tempModel$geneSymbols, file=genFileName(paste(tName, ".txt", sep="")), quote=FALSE, row.names=FALSE, col.names=FALSE)
  # save cluster 1 gene list
  write.table(tempModel$geneSymbols[tempClusters==1], file=genFileName(paste(tName, "c1.txt", sep="")))
  
  # save cluster 2 gene list
  write.table(tempModel$geneSymbols[tempClusters==2], file=genFileName(paste(tName, "c2.txt", sep="")))

}

intrxFormerbResults <- list(model=tempModel, heatmap=tempHMap)
```

### Among Current Smokers   
```{r intrxCurrenta}
target <- medianFilter(esetCleanCurrents)
tName = "intrxCurrenta"
tempModel <- lmFitWrapper(target, c("COPD2_R7", "FinalCaDXc", "AGEcalc", "RIN", "GENDERc", "COPD2_R7*FinalCaDXc"), 6, "none", 0.05, "Interaction Model 1 Among Current Smokers")

clabels <- cbind(copdca_colors[target$indicator], smoking_colors[as.character(target$SMKc)])
tempHMap <- heatmap3(exprs(target)[tempModel$inds, ], col=bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering=target$indicator, ColSideColors=clabels,
         main=tempModel$hMapName, labCol="", keep.dendro=TRUE)

if(SAVE){
  tempClusters <- cutree(as.hclust(tempHMap$Rowv), k=2)

  # save entire gene list
  write.table(tempModel$geneSymbols, file=genFileName(paste(tName, ".txt", sep="")), quote=FALSE, row.names=FALSE, col.names=FALSE)
  # save cluster 1 gene list
  write.table(tempModel$geneSymbols[tempClusters==1], file=genFileName(paste(tName, "c1.txt", sep="")))
  
  # save cluster 2 gene list
  write.table(tempModel$geneSymbols[tempClusters==2], file=genFileName(paste(tName, "c2.txt", sep="")))

}

intrxCurrentaResults <- list(model=tempModel, heatmap=tempHMap)
```

### Among Current Smokers + ICS
```{r intrxCurrentb}
target <- medianFilter(removeFactorLevel(esetCleanCurrents, "AllegroCorticosteroidsc", "DK"))
tName = "intrxCurrentb"
tempModel <- lmFitWrapper(target, c("COPD2_R7", "FinalCaDXc", "AGEcalc", "RIN", "GENDERc", "AllegroCorticosteroidsc", "COPD2_R7*FinalCaDXc"), 7, "none", 0.01, "Interaction Model 2 +ICS Among Current Smokers")

clabels <- cbind(copdca_colors[target$indicator], smoking_colors[as.character(target$SMKc)])
tempHMap <- heatmap3(exprs(target)[tempModel$inds, ], col=bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering=target$indicator, ColSideColors=clabels,
         main=tempModel$hMapName, labCol="", keep.dendro=TRUE)


if(SAVE){
  tempClusters <- cutree(as.hclust(tempHMap$Rowv), k=2)

  # save entire gene list
  write.table(tempModel$geneSymbols, file=genFileName(paste(tName, ".txt", sep="")), quote=FALSE, row.names=FALSE, col.names=FALSE)
  # save cluster 1 gene list
  write.table(tempModel$geneSymbols[tempClusters==1], file=genFileName(paste(tName, "c1.txt", sep="")))
  
  # save cluster 2 gene list
  write.table(tempModel$geneSymbols[tempClusters==2], file=genFileName(paste(tName, "c2.txt", sep="")))

}

intrxCurrentbResults <- list(model=tempModel, heatmap=tempHMap)
```


## 3 v 1 Group Analysis - this is statistically terrible    
### Among All Patients  
### Among Former Smokers  
### Among Current Smokers  


## 3 Model Approach  
### Among All Patients  
```{r tripModAll}

# Katie uses age, sex, smoking, and pack years
# She also attempts to account for corticosteroids by removing them, repeating
# the analysis and checking for enrichment in the overlapping genes

target <- medianFilter(esetClean)
adjust="none"
fdrVals <- c(0.005, 0.01, 0.05)
#fdrVals <- c(0.05, 0.1, 0.2)
intrxA <- list()
intrxB <- list()
intrxC <- list()
intrxGenes <- list()
intrxInds <- list()

for(i in 1:length(fdrVals)){
  eset <- target
	# clean the eset for COPD data and other variables
	intrxA[[i]] <- lmFitWrapper(eset, c("COPD2_R7", "FinalCaDXc", "AGEcalc",
                                     "GENDERc", "PYc", "RIN",
                                     "SMKc", "COPD2_R7*FinalCaDXc"),
                             name="COPD Signal",
                             adjust.method=adjust, p.value=fdrVals[i],
                             varOfInterest=8)
                             
	intrxB[[i]] <- lmFitWrapper(eset, c("FEV1Pc", "FinalCaDXc", "AGEcalc",
                                     "GENDERc", "PYc", "RIN",
                                     "SMKc", "FEV1Pc*FinalCaDXc"),
                             name="COPD Signal",
                             adjust.method=adjust, p.value=fdrVals[i],
                             varOfInterest=8)

	intrxC[[i]] <- lmFitWrapper(eset, c("RATIOc", "FinalCaDXc", "AGEcalc",
                                     "GENDERc", "PYc", "RIN",
                                     "SMKc", "RATIOc*FinalCaDXc"),
                             name="COPD Signal",
                             adjust.method=adjust, p.value=fdrVals[i],
                             varOfInterest=8)

  intrxGenes[[i]] <- intersect(intrxA[[i]]$geneSymbols, intersect(intrxB[[i]]$geneSymbols, intrxC[[i]]$geneSymbols))
  intrxInds[[i]] <- intersect(intrxA[[i]]$inds, intersect(intrxB[[i]]$inds, intrxC[[i]]$inds))

	length(intrxA[[i]]$geneSymbols)
	length(intrxB[[i]]$geneSymbols)
	length(intrxC[[i]]$geneSymbols)
	length(intrxGenes[[i]])
	
	#filename <- paste("OverlappingGenes", as.character(length(overlappingGenes)),
						#"fdr", as.character(fdrVals[i]), ".txt", sep="")
	#write.table(overlappingGenes, file=filename, quote=FALSE,
            #row.names=FALSE, col.names=FALSE)

}

tripleModAll <- list(target=target, adjust=adjust, fdrVals=fdrVals, 
                      intrxA=intrxA, intrxB=intrxB, intrxC=intrxC,
                      intrxGenes=intrxGenes, intrxInds=intrxInds)

```

### Among Former Smokers  
```{r tripModFormers}

# Katie uses age, sex, smoking, and pack years
# She also attempts to account for corticosteroids by removing them, repeating
# the analysis and checking for enrichment in the overlapping genes

target <- medianFilter(esetCleanFormers)
adjust="none"
fdrVals <- c(0.005, 0.01, 0.05)
intrxA <- list()
intrxB <- list()
intrxC <- list()
intrxGenes <- list()
intrxInds <- list()

for(i in 1:length(fdrVals)){
  eset <- target
  # clean the eset for COPD data and other variables
	intrxA[[i]] <- lmFitWrapper(eset, c("COPD2_R7", "FinalCaDXc", "AGEcalc",
                                     "GENDERc", "PYc", "RIN",
                                     "COPD2_R7*FinalCaDXc"),
                             name="COPD Signal",
                             adjust.method=adjust, p.value=fdrVals[i],
                             varOfInterest=7)
                             
	intrxB[[i]] <- lmFitWrapper(eset, c("FEV1Pc", "FinalCaDXc", "AGEcalc",
                                     "GENDERc", "PYc", "RIN",
                                     "FEV1Pc*FinalCaDXc"),
                             name="COPD Signal",
                             adjust.method=adjust, p.value=fdrVals[i],
                             varOfInterest=7)

	intrxC[[i]] <- lmFitWrapper(eset, c("RATIOc", "FinalCaDXc", "AGEcalc",
                                     "GENDERc", "PYc", "RIN",
                                     "RATIOc*FinalCaDXc"),
                             name="COPD Signal",
                             adjust.method=adjust, p.value=fdrVals[i],
                             varOfInterest=7)


	intrxGenes[[i]] <- intersect(intrxA[[i]]$geneSymbols, intersect(intrxB[[i]]$geneSymbols, intrxC[[i]]$geneSymbols))
    intrxInds[[i]] <- intersect(intrxA[[i]]$inds, intersect(intrxB[[i]]$inds, intrxC[[i]]$inds))


	length(intrxA[[i]]$geneSymbols)
	length(intrxB[[i]]$geneSymbols)
	length(intrxC[[i]]$geneSymbols)
	length(intrxGenes[[i]])
	
	#filename <- paste("OverlappingGenes", as.character(length(overlappingGenes)),
						#"fdr", as.character(fdrVals[i]), ".txt", sep="")
	#write.table(overlappingGenes, file=filename, quote=FALSE,
            #row.names=FALSE, col.names=FALSE)

}

tripleModFormers <- list(target=target, adjust=adjust, fdrVals=fdrVals, 
                      intrxA=intrxA, intrxB=intrxB, intrxC=intrxC,
                      intrxGenes=intrxGenes, intrxInds=intrxInds)
```

### Among Current Smokers  
```{r tripModCurrents}

# Katie uses age, sex, smoking, and pack years
# She also attempts to account for corticosteroids by removing them, repeating
# the analysis and checking for enrichment in the overlapping genes

target <- medianFilter(esetCleanCurrents)
adjust="none"
fdrVals <- c(0.005, 0.01, 0.05)
intrxA <- list()
intrxB <- list()
intrxC <- list()
intrxGenes <- list()
intrxInds <- list()

for(i in 1:length(fdrVals)){
  eset <- target
  # clean the eset for COPD data and other variables
  intrxA[[i]] <- lmFitWrapper(eset, c("COPD2_R7", "FinalCaDXc", "AGEcalc",
                                     "GENDERc", "PYc", "RIN",
                                     "COPD2_R7*FinalCaDXc"),
                             name="COPD Signal",
                             adjust.method=adjust, p.value=fdrVals[i],
                             varOfInterest=7)
                             
	intrxB[[i]] <- lmFitWrapper(eset, c("FEV1Pc", "FinalCaDXc", "AGEcalc",
                                     "GENDERc", "PYc", "RIN",
                                     "FEV1Pc*FinalCaDXc"),
                             name="COPD Signal",
                             adjust.method=adjust, p.value=fdrVals[i],
                             varOfInterest=7)

	intrxC[[i]] <- lmFitWrapper(eset, c("RATIOc", "FinalCaDXc", "AGEcalc",
                                     "GENDERc", "PYc", "RIN",
                                     "RATIOc*FinalCaDXc"),
                             name="COPD Signal",
                             adjust.method=adjust, p.value=fdrVals[i],
                             varOfInterest=7)


	intrxGenes[[i]] <- intersect(intrxA[[i]]$geneSymbols, intersect(intrxB[[i]]$geneSymbols, intrxC[[i]]$geneSymbols))
    intrxInds[[i]] <- intersect(intrxA[[i]]$inds, intersect(intrxB[[i]]$inds, intrxC[[i]]$inds))


	length(intrxA[[i]]$geneSymbols)
	length(intrxB[[i]]$geneSymbols)
	length(intrxC[[i]]$geneSymbols)
	length(intrxGenes[[i]])
	
	#filename <- paste("OverlappingGenes", as.character(length(overlappingGenes)),
						#"fdr", as.character(fdrVals[i]), ".txt", sep="")
	#write.table(overlappingGenes, file=filename, quote=FALSE,
            #row.names=FALSE, col.names=FALSE)

}

tripleModCurrents <- list(target=target, adjust=adjust, fdrVals=fdrVals, 
                      intrxA=intrxA, intrxB=intrxB, intrxC=intrxC,
                      intrxGenes=intrxGenes, intrxInds=intrxInds)
```

## GSVA Correlation Analyses
### GSVA Results Generation
```{r GSVAresults}
# generate GSVA for lmCLE signatures, atf4, and emt
eset <- esetClean

erStress

genesets <- list(lmCLEup=as.character(cleLMup$GeneSet),
                 lmCLEdn=as.character(cleLMdn$GeneSet),
                 atf4dn=atf4Cluster1_dn,
                 atf4up=atf4Cluster2_up,
                 emt=as.character(emt_genes$EMT_Genes),
                 lamUP=lamUP,
                 lamDN=lamDN,
                 lm127dn=as.character(joshLM127dn$Genes.down.with.LM),
                 lm127up=as.character(joshLM127up$Genes.up.with.LM),
                 natMedDn=natMedDn,
                 natMedUp=natMedUp,
                 tgfBeta=tgfB,
                 lmCONup=as.character(conLMup$GeneSet),
                 lmCONdn=as.character(conLMdn$GeneSet),
                 erStress=as.character(getSYMBOL(erStress, data='org.Hs.eg')),
                 cilia=as.character(getSYMBOL(cilioGen, data='org.Hs.eg')))



temp <- getSYMBOL(return_entrez(featureNames(eset)), data='org.Hs.eg')
eset2 <- eset[!is.na(temp), ]
featureNames(eset2) <- getSYMBOL(return_entrez(featureNames(eset2)), data='org.Hs.eg')

gsvaResults <- gsva(eset2, genesets)
#gsvaResults <- gsva(eset2, genesets, no.bootstraps=10, method="zscore")

lm_up=1; lm_dn=2; a_up=4; a_dn=3; emt=5;
lam_UP=6; lam_DN=7; l127d=8; l127u=9; canDn=10; canUp=11; tbeta=12

noCopd.noCancer = gsvaResults$es.obs$indicator==1
noCopd.Cancer = gsvaResults$es.obs$indicator==2
copd.noCancer = gsvaResults$es.obs$indicator==3
copd.Cancer = gsvaResults$es.obs$indicator==4

noCancer = gsvaResults$es.obs$FinalCaDXc==0
cancer = gsvaResults$es.obs$FinalCaDXc==1

copd = gsvaResults$es.obs$COPD2_R7==1
noCopd = gsvaResults$es.obs$COPD2_R7==0


```

### GSVA Correlation Plots Across All Patients
```{r gsvaCorPlot1}


```
## Two Cancer Analysis  


## Contrast Analyses Group-by-Group (Chan)  


## ER Stress and Ciliogenesis Gene sets (Duclos)   


## EMT Genes Differential Expression (Byers et al.)  
