Using Lam Data to Test COPD+Cancer Signal
========================================================
`r Sys.Date()`
---------------------------------------------


```{r GlobalVariables, include=FALSE}
cacheOption = FALSE

```


```{r defineFunctions, include=FALSE}
foldChange <- function(eset, groups, genes){
  apply(exprs(eset)[genes, groups==0], 1, mean) - 
    apply(exprs(eset)[genes, groups==1], 1, mean)
}

medianFilter <- function(eset){
  md <- median(exprs(eset))
  passFilter <- logical(featureNumber(eset))
  for(i in 1:featureNumber(eset)){
    passFilter[i] <- sum(exprs(eset)[i, ] > md) > 0
  }
  
  eset <- eset[passFilter, ]
  return(eset)
  
}

saveGeneList <- function(analysis, filename=paste(Sys.Date(), "Temp", analysis$name, ".txt", sep="")){
  
  write.table(analysis$geneSymbols, file=filename, quote=FALSE,row.names=FALSE, col.names=FALSE)

}
```


```{r LoadData, echo=FALSE, cache=cacheOption, include=FALSE}
# set the working directory to the directory of this script
setwd("/restricted/projectnb/pulmarray/LinGA_protected/Allegro/COPD_Cancer/experiments/lam/")

# load the data using a script that removes patients with:
#  Cancer = NA
#  SMK    = 3
source("/protected/projects/pulmarray/Allegro/COPD_Cancer/scripts/AllegroSetup.R")

# load the Lam (n=238) data
lamData <- readRDS("/protected/projects/pulmarray/Allegro/COPD_Cancer/experiments/lam/LamCOPD_Validate_2014-12-01_ExpressionSet.rds")

# clean the data (i.e. remove DKs and NAs)
eset <- removeFactorLevel(eset, "FinalCaDXc", "DK")
eset <- removeFactorLevel(eset, "AllegroCOPDc", "DK")
eset <- cleanNAForAnalysis(eset, "AGEcalc")
eset <- cleanNAForAnalysis(eset, "RIN")
eset <- removeBioReps(eset)
eset <- calcIndicator(eset, "FinalCaDXc", "AllegroCOPDc")

# fix the one patient sample who has a FEV1% predicted of 89.2
eset$FEV1Pc[eset$FEV1Pc==89.2] <- 0.892

eset$BATCH <- factor(eset$BATCH)
holdEset <- eset

library(znorm)
library(mclust)

```

## Thoughts for the day

I need a more directed approach than the semi-amorphous approach I have been taking
Specifically identify the goal of the analysis. This has not really changed. The goal
of the analyis is to find genes related to COPD pathogenesis that are playing a role in 
carcinogenesis of those patients with COPD. The biggest problem I've encountered is the
unexplained heterogeneity of both patient groups. I think that may be OK. The other
problem I've had is that the analysis method I used to get to my Evan's Day results
was not particularly statistically sound - I grouped patients into two groups,
1) COPD+Cancer and 2) disease-free, cancer-only, and COPD-only. On the plus side I did find
genes that seemed to differ between cancer and COPD+Cancer that were still cancer related.
They did not drastically differ between COPD and COPD+Cancer, which again, I think may be OK.

One problem - the COPD-only group is very small
One solution - use bootstrapping to upsample the group

1.) One possible method: Find cancer related genes acting differently in COPD and/or COPD+Cancer
This is not quite right because we're looking for why COPD has increased incidence of cancer

2.) A better possible method: Find the COPD genes (i.e. use Katie's paper as a model)

3.) Another possible idea is to use molecular signals for cancer, COPD, smoking in lieu of clinical
(i.e. use bronchogen, Katie's 98 genes, 100 smoking genes)

4.) Another idea - also use miRNA that Anna is working with to examine the miRNA-mRNA network 
interactions associated with COPD+Cancer

5.) Does the COPD of youth look different from teh COPD of old age
i.e. do genes change with PFTs differently at young age compared with old age
the question here would obviously be related to the field. Does the COPD field look different
in youth compared with elder age. Are low PFTs related to different mechanisms when younger?

6.) Find genes related to the transcription factors from Evans Day Enrichr results and examine
for interesting genes

```{r SOX2relatedGenes}
# Based on paper: "SOX2 Gene Regulates the Transcriptional Netework of Oncogenes and Affects Tumorigenesis of Human Lung Cancer Cells"
# Chen et al., 2012

# Generate model that was used to create Evans Day plots
eset <- holdEset
eset <- cleanNAForAnalysis(eset, "COPD2_R7")
eset <- removeFactorLevel(eset, "COPD2_R7", "DK")
eset <- calcIndicator(eset, "FinalCaDXc", "COPD2_R7")
eset$AllOrNone <- eset$indicator
eset$AllOrNone[eset$indicator != 4] <- 1
eset <- removeFactorLevel(eset, "AllOrNone", "2")
eset <- removeFactorLevel(eset, "AllOrNone", "3")
eset <- removeFactorLevel(eset, "SMKc", "1")
eset <- cleanNAForAnalysis(eset, "PYc")
eset <- removeFactorLevel(eset, "GENDERc", "DK")
esetF <- medianFilter(eset)

analysisED <- lmFitWrapper(esetF, c("AllOrNone", "AGEcalc", "RIN", "PYc", "GENDERc"), 1,
                          p.value=0.15,
                          name="Analysis9: All-or-none, PFT-defined COPD, formers only, PYs + Gender")

analysisED2 <- lmFitWrapper(eset, c("AllOrNone", "AGEcalc", "RIN", "PYc", "GENDERc"), 1,
                          p.value=0.15,
                          name="Analysis9: All-or-none, PFT-defined COPD, formers only, PYs + Gender")

#analysisLam1 <- lmFitWrapper(lamData, c(""))

```

```{r interactionModelTesting}
eset <- holdEset
eset <- cleanNAForAnalysis(eset, "COPD2_R7")
eset <- removeFactorLevel(eset, "COPD2_R7", "DK")
eset <- calcIndicator(eset, "FinalCaDXc", "COPD2_R7")

analysisIntrx <- lmFitWrapper(eset,
                              c("FinalCaDXc", "COPD2_R7", "SMKc",
                                "AGEcalc", "FinalCaDXc*COPD2_R7"), 5,
                                adjust.method="none", p.value=0.005)

#write.table(analysisIntrx$geneSymbols, file="Temp.analysisIntrxGenes120414.txt", quote=FALSE,row.names=FALSE)

```

In these interaction models, regardless of the terms we include, we still see a huge confounding of SMK

```{r COPDmodelFilterToIntrx}
eset <- holdEset
eset <- cleanNAForAnalysis(eset, "COPD2_R7")
eset <- removeFactorLevel(eset, "COPD2_R7", "DK")
eset <- removeFactorLevel(eset, "FinalCaDXc", "1")

analysisCOPD <- lmFitWrapper(eset, c("COPD2_R7"), 1, name="COPDFilterModel")

esetCOPD <- holdEset[analysisCOPD$inds, ]
esetCOPD <- cleanNAForAnalysis(esetCOPD, "COPD2_R7")
esetCOPD <- removeFactorLevel(esetCOPD, "COPD2_R7", "DK")
analysisIntrx2 <- lmFitWrapper(esetCOPD, c("FinalCaDXc", "COPD2_R7", "COPD2_R7*FinalCaDXc"), 3,
                               p.value=0.02, name="COPDModelFilterToIntrx")

saveGeneList(analysisIntrx2)

generate_heatmap2(analysisIntrx2, "indicator")

```




```{r extraCodeFromPreviousAnalysis}
kegg_pathways_in_cancer <- read.table("geneSet_pathwaysInCancer(KEGG).txt", header=TRUE)
kegg_pathways_in_cancer <- as.character(kegg_pathways_in_cancer[, 1])

picKEGGind <- which(analysisED2$fit$genes$Symbol %in% kegg_pathways_in_cancer)
eset <- holdEset[picKEGGind, ]
eset <- cleanNAForAnalysis(eset, "COPD2_R7")
eset <- removeFactorLevel(eset, "COPD2_R7", "DK")
eset <- calcIndicator(eset, "FinalCaDXc", "COPD2_R7")
eset <- removeFactorLevel(eset, "GENDERc", "DK")

esetN <- eset
exprs(esetN) <- znorm(exprs(esetN), margin=1)
ancova1 <- GlobalAncova(xx=exprs(esetN), group=esetN$FinalCaDXc, covars=eset$COPD2_R7)

```

7.) Send Adam a gmt gene list signature to test using OpenSesame

8.) Use GSEA to test my gene list/ranked list against known signals

```{r LogitCancerModel}

# Purpose here - determine if in this set, COPD contributes to Cancer prediction using
# logistic regression

eset <- holdEset
eset <- cleanNAForAnalysis(eset, "COPD2_R7")
eset <- removeFactorLevel(eset, "COPD2_R7", "DK")

myLogit <- glm(eset$FinalCaDXc ~ eset$COPD2_R7 + eset$SMKc + eset$AGEcalc, family="binomial")

eset <- removeFactorLevel(eset, "SMKc", "1")
myLogit2 <- glm(eset$FinalCaDXc ~ eset$COPD2_R7 + eset$AGEcalc, family="binomial")

eset <- holdEset
myLogit3 <- glm(eset$FinalCaDXc ~ eset$AllegroCOPDc + eset$SMKc + eset$AGEcalc, family="binomial")

eset <- holdEset
eset <- cleanNAForAnalysis(eset, "COPD2_R7")
eset <- removeFactorLevel(eset, "COPD2_R7", "DK")
eset <- cleanNAForAnalysis(eset, "PYc")
eset <- removeFactorLevel(eset, "GENDERc", "DK")

myLogit4 <- glm(eset$FinalCaDXc ~ eset$COPD2_R7 + eset$SMKc + eset$AGEcalc + eset$PYc + eset$GENDERc,
                family="binomial")

eset <- removeFactorLevel(eset, "SMKc", "1")
myLogit5 <- glm(eset$FinalCaDXc ~ eset$COPD2_R7 + eset$AGEcalc + eset$PYc + eset$GENDERc, family="binomial")
```


```{r pickingCovariates}
eset <- holdEset
eset <- cleanNAForAnalysis(eset, "COPD2_R7")
eset <- removeFactorLevel(eset, "COPD2_R7", "DK")
eset <- removeFactorLevel(eset, "SMKc", "1")
eset <- removeFactorLevel(eset, "GENDERc", "DK")

analysisCovar1 <- lmFitWrapper(eset, c("COPD2_R7"), 1, name="pickCovar_COPD",
                               adjust.method="none", p.value=0.05)

analysisCovar2 <- lmFitWrapper(eset, c("FinalCaDXc"), 1, name="pickCovar_Cancer",
                               adjust.method="none", p.value=0.05)

analysisCovar3 <- lmFitWrapper(eset, c("BATCH"), 1, name="pickCovar_Batch",
                               adjust.method="none", p.value=0.05)

analysisCovar4 <- lmFitWrapper(eset, c("RIN"), 1, name="pickCovar_RIN",
                               adjust.method="none", p.value=0.05)

analysisCovar5 <- lmFitWrapper(eset, c("AGEcalc"), 1, name="pickCovar_Age",
                               adjust.method="none", p.value=0.05)

analysisCovar6 <- lmFitWrapper(eset, c("GENDERc"), 1, name="pickCovar_Gender",
                               adjust.method="none", p.value=0.05)

analysisCovar7 <- lmFitWrapper(eset, c("BATCH"), 1, name="pickCovar_Batch",
                               adjust.method="none", p.value=0.05)

analysisCovar8 <- lmFitWrapper(eset, c("FEV1Pc"), 1, name="pickCovar_FEV1P",
                               adjust.method="none", p.value=0.05)

tempEset <- eset
eset <- removeFactorLevel(eset, "AllegroCorticosteroidsc", "DK")

analysisCovar9 <- lmFitWrapper(eset, c("AllegroCorticosteroidsc"), 1, name="pickCovar_Corticosteroids",
                               adjust.method="none", p.value=0.05)

eset <- tempEset
eset <- cleanNAForAnalysis(eset, "RACE4PFTs")
eset <- removeFactorLevel(eset, "RACE4PFTs", "1")
eset <- removeFactorLevel(eset, "RACE4PFTs", "2")
eset <- removeFactorLevel(eset, "RACE4PFTs", "6")


analysisCovar10 <- lmFitWrapper(eset, c("RACE4PFTs"), 1, name="pickCovar_Race",
                               adjust.method="none", p.value=0.05)

eset <- tempEset
eset <- cleanNAForAnalysis(eset, "RATIOc")

analysisCovar11 <- lmFitWrapper(eset, c("RATIOc"), 1, name="pickCovar_Ratio",
                               adjust.method="none", p.value=0.05)

eset <- tempEset
analysisCovar12 <- lmFitWrapper(eset, c("FinalCaDXc", "COPD2_R7", "FinalCaDXc*COPD2_R7"), 3,
                                name="pickCovar_Race",
                                adjust.method="none", p.value=0.05)

eset <- cleanNAForAnalysis(eset, "PYc")
analysisCovar13 <- lmFitWrapper(eset, c("PYc"), 1, name="pickCovar_PackYears",
                               adjust.method="none", p.value=0.05)

# now generate the set that can utilize all of the significant terms
# includes COPD2_R7, FinalCaDXc, RIN, AGEcalc, GENDERc, FEV1Pc, AllegroCorticosteroidsc, RATIOc
eset <- tempEset
eset <- removeFactorLevel(eset, "AllegroCorticosteroidsc", "DK")
eset <- cleanNAForAnalysis(eset, "RATIOc")

analysisFullModel <- lmFitWrapper(eset, c("COPD2_R7", "FinalCaDXc", "RIN", "AGEcalc", "GENDERc", 
                                            "FEV1Pc", "AllegroCorticosteroidsc", "RATIOc"),
                                    name="FullCovariateModel", varOfInterest=1,
                                    adjust.method="none", p.value=0.05)

analysisInteraction <- lmFitWrapper(eset, c("COPD2_R7", "FinalCaDXc", "RIN", "AGEcalc", "GENDERc", 
                                            "FEV1Pc", "AllegroCorticosteroidsc",
                                            "RATIOc", "FinalCaDXc*COPD2_R7"),
                                    name="FullCovariateModel", varOfInterest=9,
                                    adjust.method="fdr", p.value=0.25)

analysisInteraction2 <- lmFitWrapper(eset, c("COPD2_R7", "FinalCaDXc", "AGEcalc", "GENDERc", 
                                            "FEV1Pc", "AllegroCorticosteroidsc",
                                            "RATIOc", "FinalCaDXc*COPD2_R7"),
                                    name="FullCovariateModel", varOfInterest=8,
                                    adjust.method="fdr", p.value=0.25)

analysisInteraction3 <- lmFitWrapper(eset, c("COPD2_R7", "FinalCaDXc", "RIN", "AGEcalc", "GENDERc", 
                                            "AllegroCorticosteroidsc",
                                            "FinalCaDXc*COPD2_R7"),
                                    name="FullCovariateModel", varOfInterest=7,
                                    adjust.method="fdr", p.value=0.25)


# Katie's analysis in the 2013 paper corrects for SMK, age, sex, PY
# She uses 3 measures of lung function/COPD status - COPD, FEV1%, FEV1/FVC (RATIO)
# I'm going to try the same models but including an interaction of the 3 terms with cancer
# First we'll try in formers only, then in all samples

tempEset2 <- eset
eset <- cleanNAForAnalysis(eset, "PYc")

analysis2013.1 <- lmFitWrapper(eset, c("COPD2_R7", "FinalCaDXc", "AGEcalc",
                                       "GENDERc", "PYc", "COPD2_R7*FinalCaDXc"),
                               name="2013 Analysis - COPD, Formers Only",
                               adjust.method="none", p.value=0.05,
                               varOfInterest=6)

analysis2013.2 <- lmFitWrapper(eset, c("FEV1Pc", "FinalCaDXc", "AGEcalc",
                                       "GENDERc", "PYc", "FEV1Pc*FinalCaDXc"),
                               name="2013 Analysis - COPD, Formers Only",
                               adjust.method="none", p.value=0.05,
                               varOfInterest=6)

analysis2013.3 <- lmFitWrapper(eset, c("RATIOc", "FinalCaDXc", "AGEcalc",
                                       "GENDERc", "PYc", "RATIOc*FinalCaDXc"),
                               name="2013 Analysis - COPD, Formers Only",
                               adjust.method="none", p.value=0.05,
                               varOfInterest=6)

inds1in2 <- analysis2013.1$inds[which(analysis2013.1$inds %in% analysis2013.2$inds)]
inds1in2in3 <- inds1in2[which(inds1in2 %in% analysis2013.3$inds)]

genes2013 <- analysis2013.1$fit$genes$Symbol[inds1in2in3]

write.table(genes2013, file="Genes2013_308overlapping_p05.txt", quote=FALSE,
            row.names=FALSE, col.names=FALSE)

# need to separate into patterns of up and down genes that are repeated in the same
# manner in each analysis

# try removing PYs
analysis2013.1a <- lmFitWrapper(eset, c("COPD2_R7", "FinalCaDXc", "AGEcalc",
                                       "GENDERc", "COPD2_R7*FinalCaDXc"),
                               name="2013 Analysis - COPD, Formers Only, no Pack Years",
                               adjust.method="none", p.value=0.05,
                               varOfInterest=5)

analysis2013.2a <- lmFitWrapper(eset, c("FEV1Pc", "FinalCaDXc", "AGEcalc",
                                       "GENDERc", "FEV1Pc*FinalCaDXc"),
                               name="2013 Analysis - COPD, Formers Only, no Pack Years",
                               adjust.method="none", p.value=0.05,
                               varOfInterest=5)

analysis2013.3a <- lmFitWrapper(eset, c("RATIOc", "FinalCaDXc", "AGEcalc",
                                       "GENDERc", "RATIOc*FinalCaDXc"),
                               name="2013 Analysis - COPD, Formers Only, no Pack Years",
                               adjust.method="none", p.value=0.05,
                               varOfInterest=5)

inds1in2a <- analysis2013.1a$inds[which(analysis2013.1a$inds %in% analysis2013.2a$inds)]
inds1in2in3a <- inds1in2a[which(inds1in2a %in% analysis2013.3a$inds)]

genes2013a <- analysis2013.1a$fit$genes$Symbol[inds1in2in3a]

write.table(genes2013a, file="Genes2013a_292overlapping_p05.txt", quote=FALSE,
            row.names=FALSE, col.names=FALSE)


# try all samples
# variables include PFT measure, FinalCaDXc, AGEcalc, GENDERc, SMKc, Interaction Term, PYc?

eset <- holdEset
eset <- cleanNAForAnalysis(eset, "COPD2_R7")
eset <- removeFactorLevel(eset, "COPD2_R7", "DK")
eset <- removeFactorLevel(eset, "GENDERc", "DK")
eset <- cleanNAForAnalysis(eset, "RATIOc")
eset <- cleanNAForAnalysis(eset, "PYc")

analysis2013.1.All <- lmFitWrapper(eset, c("COPD2_R7", "FinalCaDXc", "AGEcalc",
                                            "GENDERc", "SMKc", "PYc", "COPD2_R7*FinalCaDXc"),
                                    name="2013 Analysis - COPD, Formers Only",
                                    adjust.method="none", p.value=0.05,
                                    varOfInterest=7)

analysis2013.2.All <- lmFitWrapper(eset, c("FEV1Pc", "FinalCaDXc", "AGEcalc",
                                       "GENDERc", "SMKc", "PYc", "FEV1Pc*FinalCaDXc"),
                               name="2013 Analysis - COPD, Formers Only",
                               adjust.method="none", p.value=0.05,
                               varOfInterest=7)

analysis2013.3.All <- lmFitWrapper(eset, c("RATIOc", "FinalCaDXc", "AGEcalc",
                                       "GENDERc", "SMKc", "PYc", "RATIOc*FinalCaDXc"),
                               name="2013 Analysis - COPD, Formers Only",
                               adjust.method="none", p.value=0.05,
                               varOfInterest=7)

# check for overlap in the 2013 All Patients analyses
inds1in2.All <- analysis2013.1.All$inds[which(analysis2013.1.All$inds %in% analysis2013.2.All$inds)]
inds1in2in3.All <- inds1in2.All[which(inds1in2.All %in% analysis2013.3.All$inds)]

genes2013.All <- analysis2013.1.All$fit$genes$Symbol[inds1in2in3.All]

write.table(genes2013.All, file="Genes2013.All_170overlapping_p05.txt", quote=FALSE,
            row.names=FALSE, col.names=FALSE)

# of the 3 comparisons, 82 genes overlap at p value < 0.05
# save out the 82 genes
genes2013TripleOverlap_p05 <- genes2013a[genes2013a %in% genes2013.All]

write.table(genes2013TripleOverlap_p05, file="Genes2013TripleOverlap_p05.txt", quote=FALSE,
            row.names=FALSE, col.names=FALSE)

# print a figure of these genes based on indicator status (i.e. healthy, COPD, Cancer, Both)
g2Tinds <- match(genes2013TripleOverlap_p05,analysis2013.3.All$fit$genes$Symbol)
eset <- analysis2013.3.All$eset
eset <- calcIndicator(eset, "FinalCaDXc", "COPD2_R7")
hmapOverlap <- generate_heatmap(g2Tinds, eset, tp="indicator")

overlap_clusters <- return_cluster(g2Tinds, analysis2013.3.All$eset, n.clusters=3, mthd="ward.D")
summary(as.factor(overlap_clusters))

hmapOverlapc1 <- generate_heatmap(g2Tinds, eset[])

# look at a heatmap of formers only
tempEsetg2T <- eset
eset <- removeFactorLevel(tempEsetg2T, "SMKc", "1")
hmapOverlapFormers <- generate_heatmap(g2Tinds[overlap_clusters==1], eset, tp="indicator")
hmapOverlapFormers <- generate_heatmap(g2Tinds[overlap_clusters==2], eset, tp="indicator")
hmapOverlapFormers <- generate_heatmap(g2Tinds[overlap_clusters==3], eset, tp="indicator")

# look at a heatmap of currents only
eset <- removeFactorLevel(tempEsetg2T, "SMKc", "2")
hmapOverlapCurrents <- generate_heatmap(g2Tinds, eset, tp="indicator")

# generate PCA and look at t-tests
pca_overlapping <- prcomp(znorm(x=exprs(eset), margin=1), center=FALSE, scale.=FALSE)
boxplot(pca_overlapping$rotation[, 1] ~ eset$indicator)
boxplot(pca_overlapping$rotation[, 2] ~ eset$indicator)
boxplot(pca_overlapping$rotation[, 3] ~ eset$indicator)

# look at all samples but exclude the PY term (just to be consistent with the above analyses)
eset <- analysis2013.1.All$eset
analysis2013.1a.All <- lmFitWrapper(eset, c("COPD2_R7", "FinalCaDXc", "AGEcalc",
                                            "GENDERc", "SMKc", "COPD2_R7*FinalCaDXc"),
                                    name="2013 Analysis - COPD, Formers Only",
                                    adjust.method="none", p.value=0.05,
                                    varOfInterest=6)

analysis2013.2a.All <- lmFitWrapper(eset, c("FEV1Pc", "FinalCaDXc", "AGEcalc",
                                       "GENDERc", "SMKc", "FEV1Pc*FinalCaDXc"),
                               name="2013 Analysis - COPD, Formers Only",
                               adjust.method="none", p.value=0.05,
                               varOfInterest=6)

analysis2013.3a.All <- lmFitWrapper(eset, c("RATIOc", "FinalCaDXc", "AGEcalc",
                                       "GENDERc", "SMKc", "RATIOc*FinalCaDXc"),
                               name="2013 Analysis - COPD, Formers Only",
                               adjust.method="none", p.value=0.05,
                               varOfInterest=6)

# create the lists of overlapping genes
inds1in2.Alla <- analysis2013.1a.All$inds[which(analysis2013.1a.All$inds %in% analysis2013.2a.All$inds)]
inds1in2in3.Alla <- inds1in2.Alla[which(inds1in2.Alla %in% analysis2013.3a.All$inds)]

genes2013.Alla <- analysis2013.1a.All$fit$genes$Symbol[inds1in2in3.Alla]


#This is what I did lazily - forgot to check overlap between the last three analyses
#genesAllOverlap <- genes2013TripleOverlap_p05[genes2013TripleOverlap_p05 %in% analysis2013.3a.All$geneSymbols]
#write.table(genesAllOverlap, file="Genes2013AllOverlap_73_p05.txt", quote=FALSE,
#            row.names=FALSE, col.names=FALSE)

genesAllOverlap <- genes2013TripleOverlap_p05[genes2013TripleOverlap_p05 %in% genes2013.Alla]
write.table(genesAllOverlap, file="Genes2013AllOverlap_71_p05_010715.txt", quote=FALSE,
            row.names=FALSE, col.names=FALSE)


# why not? - try using the Allegro COPD term and check for overlap in gene lists

```


```{r printHeatmapPDFs}
eset <- analysis2013.1a.All$eset
geneInds <- match(genesAllOverlap, analysis2013.1a.All$fit$genes$Symbol)


for(i in 1:length(genesAllOverlap)){
  tle <- genesAllOverlap[i]
  pdf(paste("OverlapBoxplot_", tle, ".pdf", sep=""))
  boxplot(exprs(eset)[geneInds[i], ] ~ eset$indicator, color=c("green", "red", "yellow", "blue"), title=tle)
  dev.off()
}

pdf("allOverlappingGenesBoxplots.pdf")
for(i in 1:length(genesAllOverlap)){
  tle <- genesAllOverlap[i]
  boxplot(exprs(eset)[geneInds[i], ] ~ eset$indicator, col=c("green", "red", "yellow", "blue"), main=tle)
}
dev.off()

```

```{r clusterAnalysis}
eset <- analysis2013.1a.All$eset
geneInds <- match(genesAllOverlap, analysis2013.1a.All$fit$genes$Symbol)

clusters <- return_cluster(geneInds, eset, n.clusters=2)
write.table(genesAllOverlap[clusters==1],
            file="Genes2013AllOverlap_71_p05_cluster1_63.txt",
            quote=FALSE,row.names=FALSE, col.names=FALSE)

write.table(genesAllOverlap[clusters==2],
            file="Genes2013AllOverlap_71_p05_cluster2_10.txt",
            quote=FALSE,row.names=FALSE, col.names=FALSE)

# Cluster 1 appears to be the cluster strongly associated with the ER pathways
# Cluster 2 does not appear to have any biological enrichment

pdf("genes2013AllOverlap_73_p05_cluster1_63_heatmap.pdf")
generate_heatmap(geneInds[clusters==1], eset, tp="indicator",
                 mn="Genes 2013 All Patients; p < 0.05; \nCluster 1 - 63 genes")
dev.off()

pdf("genes2013AllOverlap_73_p05_cluster2_10_heatmap.pdf")
generate_heatmap(geneInds[clusters==2], eset, tp="indicator",
                 mn="Genes 2013 All Patients; p < 0.05; \nCluster 2 - 10 genes")
dev.off()

pdf("genes2013AllOverlap_71_p05_010715_heatmap.pdf")
generate_heatmap(geneInds, eset, tp="indicator",
                 mn="71 Genes, All Patients, p < 0.05 \n Overlapping 12 Models")
dev.off()

```


```{r compare2013GenesByGroup}
# Run t-tests between each of the 4 groups on the 73 genes overlapping across the 
# models run in the style of Katie's 2013 COPD paper
# i.e. run t-tests between COPD+Cancer v COPD, COPD+Cancer v Cancer, C+C v healthy
# also, between Cancer v COPD, Cancer v healthy
# and finally, COPD v healthy


```