
New List for Sunday night:
1. Steiling COPD signal enrichment and/or differential analysis - heatmap
2. Nature medicine cancer enrichment and/or differential analysis - heatmap
3. COPD+Cancer interaction in clean and super clean eset
4. Enrichr results from interaction - look for UPR and EMT


New List for Monday night:
1. <del>Choose eset - choices - clean eset, super clean eset, formers or all smokers</del> 7:25 pm
2. <del>Previous signature enrichment</del> 7:51 pm
3. <del>Genes by interaction overlap</del> <del>and by t-tests/contrasts</del>
4. <del>Figure of overlap and each set on its own</del> - None
5. ATF4 and/or emt targets
6. Splity by cluster for enrichment - GSEA, enrichr
7. row labeled heatmaps
8. atf4 leading edge(s) heatmaps in clinical and cell data
9. EMT enrichment by chi-squared
10. EMT leading edge heatmap in clinical (and cell?)
  
```{r defineFunctions, include=FALSE}
foldChange <- function(eset, groups, genes){
  apply(exprs(eset)[genes, groups==0], 1, mean) - 
    apply(exprs(eset)[genes, groups==1], 1, mean)
}

medianFilter <- function(eset){
  md <- median(exprs(eset))
  passFilter <- logical(featureNumber(eset))
  for(i in 1:featureNumber(eset)){
    passFilter[i] <- sum(exprs(eset)[i, ] > md) > 0
  }
  
  eset <- eset[passFilter, ]
  return(eset)
  
}

saveGeneList <- function(analysis, filename=paste(Sys.Date(), "Temp", analysis$name, ".txt", sep="")){
  
  write.table(analysis$geneSymbols, file=filename, quote=FALSE,row.names=FALSE, col.names=FALSE)

}

saveRankedList <- function(fit, varColInd, filename = "tempRankedList"){
  filename = paste(paste(Sys.Date(), colnames(fit$t)[varColInd], filename, sep="_"), "rnk", sep=".")
  print(filename)
  temp <- data.frame(geneSymbols = fit$genes$Symbol, ts = fit$t[, varColInd])
  write.table(temp, file=filename, row.names=FALSE, quote=FALSE, col.names=TRUE, sep="\t", eol="\n")
  
  
}

saveGeneSet <- function(fit, inds, filename = "tempGeneSet", geneSetName="GeneSet", description=""){
  filename = paste(paste(Sys.Date(), filename, sep="_"), "gmx", sep=".")
  print(filename)
  temp <- data.frame(geneSymbols=c(geneSetName, description, fit$genes$Symbol[inds]))
  write.table(temp, file=filename, row.names=FALSE, quote=FALSE, col.names=FALSE)
  
}


```

```{r setup, eval=TRUE, echo=TRUE, include=FALSE, results='hide'}

setwd("/protected/projects/pulmarray/Allegro/COPD_Cancer/experiments/2015-05-04_CBM_ATS_2015/")
source("/protected/projects/pulmarray/Allegro/COPD_Cancer/scripts/AllegroSetup.R")
source("../2015-05-03_CBM_ATS_2015/plotGSEA.R")
# fix the one patient with wonky data
# eventually this should just be saved in the RDS file
holdEset <- eset
holdEset$FEV1Pc[holdEset$FEV1Pc==89.2] <- 0.892

emt_genes <- read.csv("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/emt_gene_symbols_byers_2013.csv", header=FALSE)
temp <- as.factor(unique(as.character(emt_genes$V1)))
emt_genes <- data.frame("EMT_Genes" = temp)

GOLD_colors = c("0" = "gray100", "1" = "gray75", "2" = "gray50", "3" = "gray25", "4" = "black")

# load ATF4 BEASB2 gene perturbation data for analysis
atf4_data <- readRDS("/protected/projects/pulmarray/Allegro/ATF4/ATF4data_2015MAR17_ExpressionSet.rds")

lam_sign <- read.delim("/protected/projects/pulmarray/Allegro/COPD_Cancer/geneSubsets/katie98genesEntrez.txt")
lam_sign[, 3] <- paste(lam_sign[, 2], "_at", sep="")

#atf4_targets <- read.delim("ATF4_Targets_Enrichr_TRANSFACJASPAR_270Genes.txt", sep=",", as.is=TRUE, header=FALSE)
#atf4_targets <- as.character(atf4_targets)

#a4LEup <- read.delim("ATF4_targets_cluster1_270Genes_LeadingEdge_Up.csv", header=FALSE)
#a4LEup <- as.character(a4LEup$V1)

#a4LEdn <- read.delim("ATF4_targets_cluster2_270Genes_LeadingEdge_Down.csv", header=FALSE)
#a4LEdn <- as.character(a4LEdn$V1)
library(znorm)
library(diptest)



```

```{r cleanup}

# Remaining question - should we keep or remove the COPD GOLD Status 1 patients? i.e. mild disease
# What about patients using or their charts indicating they have used corticosteroids?
esetClean <- holdEset
esetClean <- cleanNAForAnalysis(esetClean, "COPD2_R7")
esetClean <- cleanNAForAnalysis(esetClean, "RATIOc")
esetClean <- cleanNAForAnalysis(esetClean, "PYc")
esetClean <- cleanNAForAnalysis(esetClean, "RIN")
esetClean <- removeFactorLevel(esetClean, "COPD2_R7", "DK")
esetClean <- removeFactorLevel(esetClean, "FinalCaDXc", "DK")
esetClean <- removeFactorLevel(esetClean, "COPD2_R7", "DK")
esetClean <- removeFactorLevel(esetClean, "GENDERc", "DK")

esetClean <- calcIndicator(esetClean, "FinalCaDXc", "COPD2_R7")
esetClean$smkindic <- as.numeric(as.character(esetClean$indicator))
esetClean$smkindic[esetClean$SMKc==2] <- esetClean$smkindic[esetClean$SMKc==2] + 4
esetClean$smkindic <- as.factor(esetClean$smkindic)

# Define the GOLD Stage appropriate for all patients
esetClean$GOLD <- factor(x=rep(0, sampleNumber(esetClean)), levels=c("0","1","2","3","4"))
# GOLD Stage 1 Ratio < 0.7, FEV1 >= 8
esetClean$GOLD[esetClean$RATIOc < 0.7 & esetClean$FEV1Pc >= 0.8] <- 1
# GOLD Stage 2 Ratio < 0.7, 50 <= FEV1 < 80
esetClean$GOLD[esetClean$RATIOc < 0.7 & esetClean$FEV1Pc < 0.8] <- 2
# GOLD Stage 3 Ratio < 0.7, 30 <= FEV1 <= 50
esetClean$GOLD[esetClean$RATIOc < 0.7 & esetClean$FEV1Pc < 0.5] <- 3
# GOLD Stage 4 Ratio < 0.7, FEV1 < 30%
esetClean$GOLD[esetClean$RATIOc < 0.7 & esetClean$FEV1Pc < 0.3] <- 4

esetCleanFormers <- removeFactorLevel(esetClean, "SMKc", "1")
# Generate a super clean dataset devoid of all patients with other pulmonary diseases
esetSuperClean <- esetClean
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroPNAc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroTBc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroSarcoidc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroIPFc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroIPFc", "DK")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroAtelectasisc", "1")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroAtelectasisc", "DK")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroAsthmac", "DK")
esetSuperClean <- removeFactorLevel(esetSuperClean, "AllegroAsthmac", "1")

# clean up ATF4 data
varLabels(atf4_data) <- "SampleID"
atf4_data$SampleID[grep("ATF4", atf4_data$SampleID)] <- "ATF4"
atf4_data$SampleID[grep("Ctrl", atf4_data$SampleID)] <- "Control"
atf4_data$SampleID <- as.factor(atf4_data$SampleID)

atf4_colors <- c("1" = "red", "2" = "white")


```

```{r previousSignatures}
eset <- esetCleanFormers

copdSignal <- lmFitWrapper(eset, c("COPD2_R7", "FinalCaDXc", "AGEcalc", "GENDERc", "PYc", "RIN"),varOfInterest=1, adjust.method="none", p.value=0.05, name="COPD_Signal")

cancerSignal <- lmFitWrapper(eset, c("COPD2_R7", "FinalCaDXc", "AGEcalc", "GENDERc", "PYc", "RIN"),varOfInterest=2, adjust.method="none", p.value=0.05, name="Cancer_Signal")

if(TRUE){
  saveRankedList(fit=copdSignal$fit, 2, filename="copdSignalForLamValidation")
  saveRankedList(fit=cancerSignal$fit, 3, filename="cancerSignalForNatMedValidation")
}

# create figures from GSEA results
# need to get these working for the poster
plot.gsea("dataForPreviousValidations/GSEA_results/esetCleanFormers_NatMed_Cancer_Validation.GseaPreranked.1430781996357/edb/2015-05-04_FinalCaDXc1_cancerSignalForNatMedValidation.rnk",c( "dataForPreviousValidations/GSEA_results/NEW_80_GENE_BIOMARKER_DOWN.txt", "dataForPreviousValidations/GSEA_results/NEW_80_GENE_BIOMARKER_UP.txt"))

plot.gsea("dataForPreviousValidations/GSEA_results/esetCleanFormers_Lam_COPD_Validation.GseaPreranked.1430782069373/edb/2015-05-04_COPD2_R71_copdSignalForLamValidation.rnk", c("dataForPreviousValidations/GSEA_results/LAM_44GENES_DOWN_IN_COPD_AIRWAY.txt", "dataForPreviousValidations/GSEA_results/LAM_54GENES_UP_IN_COPD_AIRWAY.txt"))

if (FALSE){
  pdf("intrx308ClustersGSEA.pdf")
  plot.gsea("intrxGenes_and_Enrichr/GSEA_Results/clusters308.GseaPreranked/edb/ATF4RankedList.rnk", "intrxGenes_and_Enrichr/GSEA_Results/clusters308.GseaPreranked/INTRX308_CLUSTER2.txt")

  plot.gsea("intrxGenes_and_Enrichr/GSEA_Results/clusters308.GseaPreranked/edb/ATF4RankedList.rnk", "intrxGenes_and_Enrichr/GSEA_Results/clusters308.GseaPreranked/INTRX308_CLUSTER1.txt")
  dev.off()
  }


```



```{r interactionModels, eval=TRUE, echo=TRUE, results='hide'}

# Katie uses age, sex, smoking, and pack years
# She also attempts to account for corticosteroids by removing them, repeating
# the analysis and checking for enrichment in the overlapping genes

eset <- medianFilter(esetCleanFormers)
fdrVals <- c(0.05, 0.1, 0.2)
intrxA <- list()
intrxB <- list()
intrxC <- list()
intrxGenes <- list()

for(i in 1:length(fdrVals)){
  
  intrxA[[i]] <- lmFitWrapper(eset, c("COPD2_R7", "FinalCaDXc", "AGEcalc",
                                     "GENDERc", "PYc", "RIN",
                                    "COPD2_R7*FinalCaDXc"),
                             name="COPD*Cancer Signal",
                             adjust.method="fdr", p.value=fdrVals[i],
                             varOfInterest=7)
                             
	intrxB[[i]] <- lmFitWrapper(eset, c("FEV1Pc", "FinalCaDXc", "AGEcalc",
                                     "GENDERc", "PYc", "RIN",
                                     "FEV1Pc*FinalCaDXc"),
                             name="FEV1%*Cancer Signal",
                             adjust.method="fdr", p.value=fdrVals[i],
                             varOfInterest=7)

	intrxC[[i]] <- lmFitWrapper(eset, c("RATIOc", "FinalCaDXc", "AGEcalc",
                                     "GENDERc", "PYc", "RIN",
                                     "RATIOc*FinalCaDXc"),
                             name="RATIO*Cancer Signal",
                             adjust.method="fdr", p.value=fdrVals[i],
                             varOfInterest=7)

	overlappingGenes <- intrxA[[i]]$geneSymbols[intrxA[[i]]$geneSymbols %in% intrxB[[i]]$geneSymbols]
	intrxGenes[[i]] <- overlappingGenes[overlappingGenes %in% intrxC[[i]]$geneSymbols]

	length(intrxA[[i]]$geneSymbols)
	length(intrxB[[i]]$geneSymbols)
	length(intrxC[[i]]$geneSymbols)
	length(intrxGenes[[i]])
	
}


# generate ATF4 model(s)
atf4Model <- lmFitWrapper(atf4_data, c("SampleID"), name="ATF4 Perturbation Data", adjust.method="fdr",
                          p.value=0.05, varOfInterest=1)


if(FALSE){
  # save out the 308 genes
  write.table(intrxGenes[[1]], row.names=FALSE, quote=FALSE, col.names=FALSE, file="Intrxfdr05_308Genes.txt")
  
}

# load atf4 targets and generate heatmap of 308 genes labeled with tick marks

# 36 of 308 genes are ATF4 targets
atf4_targets <- read.csv("intrxGenes_and_Enrichr/Intrxfdr05_308Genes_TRANSFAC_and_JASPAR_PWMs_ATF4_targets.txt", header=FALSE)
atf4_targets <- as.character(unlist(atf4_targets))

rlabels <- numeric(length(intrxGenes[[1]]))
rlabels[match(atf4_targets,intrxGenes[[1]])] <- 1
rlabels[rlabels==0] <- 2
rlabels <- atf4_colors[rlabels]

inds308 <- match(intrxGenes[[1]], intrxA[[1]]$fit$genes$Symbol)
eset$indicator <- relevel(relevel(eset$indicator, ref="3"), "1")
clabels <- cbind(copdca_colors[eset$indicator])

if(FALSE){
pdf("intrxfdr05_308Genes_heatmap_with_atf4_targets.pdf")
heatmap3(exprs(eset)[inds308, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = eset$indicator, ColSideColors = clabels,
         main = "308 Genes Significant for Overlapping Interaction Terms\nFDR < 0.05\nFormer Smokers Only",
         labCol="", RowSideColors = rlabels)

dev.off()
}

intrx308_clusters <- return_cluster(inds308, eset, n.clusters=2)
# test the clusters are correct
rlabels <- atf4_colors[intrx308_clusters]
heatmap3(exprs(eset)[inds308, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = eset$indicator, ColSideColors = clabels,
         main = "308 Genes Significant for Overlapping Interaction Terms\nFDR < 0.05\nFormer Smokers Only",
         labCol="", RowSideColors = rlabels)


if(FALSE){
  saveRankedList(atf4Model$fit, 2, filename="atf4_ranked_list")
  write.table(intrxGenes[[1]][intrx308_clusters==1], file="intrxfdr05_308Genes_cluster1", quote=FALSE, row.names=FALSE, col.names=FALSE)
  
  write.table(intrxGenes[[1]][intrx308_clusters==2], file="intrxfdr05_308Genes_cluster2", quote=FALSE, row.names=FALSE, col.names=FALSE)  
}



indsATF4 <- match(atf4_targets, intrxA[[1]]$fit$genes$Symbol)
heatmap3(exprs(eset)[indsATF4, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = eset$indicator, ColSideColors = clabels,
         main = "36/308 ATF4 Target Genes Significant for Overlapping Interaction Terms\nFDR < 0.05\nFormer Smokers Only",
         labCol="")

c1LE <- read.table("atf4_GSEA/atf4_intrx_GSEA/cluster1_leadingEdge.txt")
c1LE <- as.character(c1LE$V1)
indsc1 <- match(c1LE, intrxA[[1]]$fit$genes$Symbol)
indsc1A <- match(c1LE, atf4Model$fit$genes$Symbol)
clabels <- copdca_colors[eset$indicator]
heatmap3(exprs(eset)[indsc1, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = eset$indicator, ColSideColors = clabels,
         main = "Cluster 1 Leading Edge of ATF4 Enrichment\nFDR < 0.05\nFormer Smokers Only",
         labCol="")

clabelsA <- cancer_colors[atf4_data$SampleID]
heatmap3(exprs(atf4_data)[indsc1A, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = atf4_data$SampleID, ColSideColors = clabelsA,
         main = "Cluster 1 Leading Edge of ATF4 Enrichment\nFDR < 0.05\nCell Data",
         labCol="")

c2LE <- read.table("atf4_GSEA/atf4_intrx_GSEA/cluster2_leadingEdge.txt")
c2LE <- as.character(c2LE$V1)
indsc2 <- match(c2LE, intrxA[[1]]$fit$genes$Symbol)
indsc2 <- indsc2[!is.na(indsc2)]
heatmap3(exprs(eset)[indsc2, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = eset$indicator, ColSideColors = clabels,
         main = "Cluster 2 Leading Edge of ATF4 Enrichment\nFDR < 0.05\nFormer Smokers Only",
         labCol="")
indsc2A <- match(c2LE, atf4Model$fit$genes$Symbol)
indsc2A <- indsc2A[!is.na(indsc2A)]
heatmap3(exprs(atf4_data)[indsc2A, ], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = atf4_data$SampleID, ColSideColors = clabelsA,
         main = "Cluster 2 Leading Edge of ATF4 Enrichment\nFDR < 0.05\nCell Data",
         labCol="")

if(TRUE){
  pdf("cluster1_308intrx_clinical_atf4LeadingEdge.pdf")
  heatmap3(exprs(eset)[indsc1, ], col = bluered,
           hclustfun=function(d) hclust(d, method="ward.D"),
           col.clustering = eset$indicator, ColSideColors = clabels,
           main = "Cluster 1 Leading Edge of ATF4 Enrichment\nFDR < 0.05\nFormer Smokers Only",
          labCol="")
  dev.off()
  
  pdf("cluster1_308intrx_cell_atf4LeadingEdge.pdf")
  heatmap3(exprs(atf4_data)[indsc1A, ], col = bluered,
           hclustfun=function(d) hclust(d, method="ward.D"),
           col.clustering = atf4_data$SampleID, ColSideColors = clabelsA,
           main = "Cluster 1 Leading Edge of ATF4 Enrichment\nFDR < 0.05\nCell Data",
           labCol="")
  dev.off()
  
  pdf("cluster2_308intrx_clinical_atf4LeadingEdge.pdf")
  heatmap3(exprs(eset)[indsc2, ], col = bluered,
           hclustfun=function(d) hclust(d, method="ward.D"),
           col.clustering = eset$indicator, ColSideColors = clabels,
           main = "Cluster 2 Leading Edge of ATF4 Enrichment\nFDR < 0.05\nFormer Smokers Only",
           labCol="")
  dev.off()
  
  pdf("cluster2_308intrx_cell_atf4LeadingEdge.pdf")
  heatmap3(exprs(atf4_data)[indsc2A, ], col = bluered,
           hclustfun=function(d) hclust(d, method="ward.D"),
           col.clustering = atf4_data$SampleID, ColSideColors = clabelsA,
           main = "Cluster 2 Leading Edge of ATF4 Enrichment\nFDR < 0.05\nCell Data",
           labCol="")
  dev.off()

  
}
```



```{r contrastModels}
# setup the variables for contrast building
eset <- medianFilter(esetCleanFormers)
eset$Cancer <- as.character(eset$FinalCaDXc)
eset$Cancer[eset$FinalCaDXc==1] <- "Cancer"
eset$Cancer[eset$FinalCaDXc==0] <- "noCancer"
eset$COPD <- as.character(eset$COPD2_R7)
eset$COPD[eset$COPD2_R7==1] <- "COPD"
eset$COPD[eset$COPD2_R7==0] <- "noCOPD"
eset$CC <- as.factor(paste(eset$Cancer, eset$COPD, sep="."))


design <- model.matrix(~0 + eset$CC)
colnames(design) <- levels(eset$CC)

fit <- lmFit(eset, design)

# of interest: (COPD+Cancer - COPD) - (Cancer - COPD) = COPD+Cancer - Cancer
# realistically we want a few groups of the genes significant for these terms
# 1. the genes differing between COPD.Cancer and Cancer
# 2. the (1) genes and those also changing between COPD.Cancer and COPD
cont.matrix <- makeContrasts(
  #Difference=(Cancer.COPD-noCancer.COPD)-(Cancer.noCOPD-noCancer.COPD),
  #COPDvCancer=Cancer.noCOPD-noCancer.COPD,
  COPDCavCOPD=Cancer.COPD-noCancer.COPD,
  COPDCavCancer=Cancer.COPD-Cancer.noCOPD,
  COPDCavNormal=Cancer.COPD-noCancer.noCOPD,
  COPDvNormal=noCancer.COPD-noCancer.noCOPD,
  CancervNormal=Cancer.noCOPD-noCancer.noCOPD,
  CancervCOPD=Cancer.noCOPD-noCancer.COPD,
  levels=design)


fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)

fdr <- 0.2
results <- decideTests(fit2, adjust.method="fdr", p.value=fdr)

copdcaGenes <- which(results[, 1] != 0 & results[, 2] != 0 & results[, 3] != 0)
contrGenes <- which(results[, 1] != 0 & results[, 2] != 0 & results[, 3] != 0 & results[, 4] != 0 & results[, 5] != 0)

COPDCaVNormOnlyGenes <- which(results[, 1] == 0 & results[, 2] == 0 & results[, 3] != 0 & results[, 4] == 0 & results[, 5] == 0)

eset$indicator <- relevel(relevel(eset$indicator, ref="3"), "1")
clabels <- cbind(copdca_colors[eset$indicator], smoking_colors[eset$SMKc])
heatmap3(exprs(eset)[copdcaGenes,], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = eset$indicator, ColSideColors = clabels,
         main = "Clean Eset fdr < 0.20",
         labCol="")

ccGenes <- fit2$genes$Symbol[copdcaGenes]
if(FALSE){
  write.table(ccGenes, file="copdcaGenesbyContrast_fdr20_supercleanFormers", quote=FALSE,
              row.names=FALSE, col.names=FALSE)
  
  pdf("contrModels_copdca_fdr20_80Genes_cleanFormers.pdf")
  heatmap3(exprs(eset)[copdcaGenes,], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = eset$indicator, ColSideColors = clabels,
         main = "Clean Eset fdr < 0.20\nFormers Only",
         labCol="")
  dev.off()
  
  write.table(fit2$genes$Symbol[COPDCaVNormOnlyGenes], file="COPDCavNormOnly_fdr20_cleanFormersMedianFilter.txt", quote=FALSE, 
              row.names=FALSE, col.names=FALSE)
  
}

emtInds <- match(emt_genes$EMT_Genes, fit2$genes$Symbol)
emtInds <- emtInds[!is.na(emtInds)]
heatmap3(exprs(eset)[emtInds,], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = eset$indicator, ColSideColors = clabels,
         main = "EMT Gene Set Byers 2013",
         labCol="")

if(FALSE){
  saveRankedList(fit2, 1, "fit2_COPDCavCOPD")
  saveRankedList(fit2, 2, "fit2_COPDCavCancer")
  saveRankedList(fit2, 3, "fit2_COPDCavNormal")
  saveRankedList(fit2, 4, "fit2_COPDvNormal")
  saveRankedList(fit2, 5, "fit2_CancervNormal")
  saveRankedList(fit2, 6, "fit2_CancervCOPD")
}

emtLE <- read.table("emt_ranked_lists/EMT_COPDCavNormal_LeadingEdge.txt", header=FALSE)
emtLE <- as.character(unlist(emtLE))

emtLEinds <- match(emtLE, fit2$genes$Symbol)
heatmap3(exprs(eset)[emtLEinds,], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = eset$indicator, ColSideColors = clabels,
         main = "EMT Leading Edge from COPD+Cancer v Normal",
         labCol="")

ela <- match(emtLE, atf4Model$fit$genes$Symbol)
heatmap3(exprs(atf4Model$eset)[ela,], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = atf4Model$eset$SampleID, ColSideColors = clabelsA,
         main = "EMT Leading Edge from COPD+Cancer v Normal in cell",
         labCol="")

if(FALSE){
  pdf("EMT_leadingEDGE_COPDCavNormal.pdf")
  heatmap3(exprs(eset)[emtLEinds,], col = bluered,
         hclustfun=function(d) hclust(d, method="ward.D"),
         col.clustering = eset$indicator, ColSideColors = clabels,
         main = "EMT Leading Edge from COPD+Cancer v Normal",
         labCol="")
  dev.off()
  
}

```



```{r emt}
eset <- esetCleanFormers

forGenes <- lmFitWrapper(eset, "FinalCaDXc", 1, "fdr", 0.05)
inEMT <- as.factor(forGenes$fit$genes$Symbol %in% emt_genes$EMT_Genes)
inCCC <- as.factor(forGenes$fit$genes$Symbol %in% ccGenes)
chisq.test(inEMT, inCCC)

generate_heatmap(copdcaGenes, eset, mthd="ward.D", tp="indicator")

```

```{r table1_demographics}

eset <- esetClean
eset$indicator <- relevel(relevel(eset$indicator, ref="3"), "1")

byPhen <- eset$indicator
tapply(eset$SMKc, byPhen, summary)
chisq.test(eset$SMKc, byPhen)

# cancer status by indicator
tapply(eset$FinalCaDXc, byPhen, summary)
chisq.test(eset$FinalCaDXc, byPhen)

# gender by indicator
tapply(eset$GENDERc, byPhen, summary)
chisq.test(eset$GENDERc, byPhen)

# age by indicator
tapply(eset$AGEcalc, byPhen,
       function(x){paste("mean:",round(mean(x), 2), "sd:", round(sd(x),2))})
summary(aov(eset$AGEcalc ~ byPhen))

# PY by indicator
tapply(eset$PYc, byPhen,
       function(x){paste("mean:",round(mean(x, na.rm=TRUE), 2), "sd:", round(sd(x, na.rm=TRUE),2))})
summary(aov(eset$PYc ~ byPhen))

# FEV1Pc by indicator
tapply(eset$FEV1Pc, byPhen,
       function(x){paste("mean:",round(mean(x, na.rm=TRUE), 2), "sd:", round(sd(x, na.rm=TRUE),2))})
summary(aov(eset$FEV1Pc ~ byPhen))
t.test(eset$FEV1Pc[byPhen==1], eset$FEV1Pc[byPhen==2])
        
# RATIOc by indicator
tapply(eset$RATIOc, byPhen,
       function(x){paste("mean:",round(mean(x, na.rm=TRUE), 2), "sd:", round(sd(x, na.rm=TRUE),2))})
summary(aov(eset$RATIOc ~ byPhen))

# RIN by indicator
tapply(eset$RIN, byPhen,
       function(x){paste("mean:",round(mean(x, na.rm=TRUE), 2), "sd:", round(sd(x, na.rm=TRUE),2))})
summary(aov(eset$RIN ~ byPhen))

```


```{r intrxVizContd}

eset <- intrxA[[1]]$eset
eset$indicator <- relevel(relevel(eset$indicator, ref="3"), "1")
inds <- match(intrxGenes[[1]], intrxA[[1]]$fit$genes$Symbol)
intrxGenes_pca <- prcomp(znorm(x=exprs(eset)[inds, ], margin=1), center=FALSE, scale.=FALSE)

boxplot(intrxGenes_pca$rotation[, 1] ~ eset$indicator, names=c("Neither", "COPD", "Cancer", "COPD.Cancer"),
            col=copdca_colors,
            main="PC1 from 308 interaction genes at fdr < 0.05",
            xlab="Group", ylab="First Principal Component")

intrxC1 <- intrxGenes[[1]][intrx308_clusters==1]
intrxC2 <- intrxGenes[[1]][intrx308_clusters==2]

intrxC1i <- match(intrxC1, intrxA[[1]]$fit$genes$Symbol)
intrxC2i <- match(intrxC2, intrxA[[1]]$fit$genes$Symbol)

# mean of genes by grou
c1ind1mean <- apply(exprs(eset)[intrxC1i, eset$indicator==1], 2, mean)
c1ind2mean <- apply(exprs(eset)[intrxC1i, eset$indicator==2], 2, mean)
c1ind3mean <- apply(exprs(eset)[intrxC1i, eset$indicator==3], 2, mean)
c1ind4mean <- apply(exprs(eset)[intrxC1i, eset$indicator==4], 2, mean)
meanData <- matrix(nrow=sampleNumber(eset), ncol=2)
meanData[, 1] <- c(c1ind1mean, c1ind2mean, c1ind3mean, c1ind4mean)
meanData[, 2] <- c(rep(1,length(c1ind1mean)), rep(2,length(c1ind2mean)), rep(3, length(c1ind3mean)), rep(4, length(c1ind4mean)))


boxplot(meanData[, 1] ~meanData[, 2])


c2ind1mean <- apply(exprs(eset)[intrxC2i, eset$indicator==1], 2, mean)
c2ind2mean <- apply(exprs(eset)[intrxC2i, eset$indicator==2], 2, mean)
c2ind3mean <- apply(exprs(eset)[intrxC2i, eset$indicator==3], 2, mean)
c2ind4mean <- apply(exprs(eset)[intrxC2i, eset$indicator==4], 2, mean)
meanData2 <- matrix(nrow=sampleNumber(eset), ncol=2)
meanData2[, 1] <- c(c2ind1mean, c2ind2mean, c2ind3mean, c2ind4mean)
meanData2[, 2] <- c(rep(1,length(c2ind1mean)), rep(2,length(c2ind2mean)), rep(3, length(c2ind3mean)), rep(4, length(c2ind4mean)))


boxplot(meanData2[, 1] ~meanData2[, 2])
```