SILCC/Allegro - COPD+Cancer Analysis
========================================================

Finding COPD-specific genes/pathways involved in tumorigenesis
-----------------------------------------------------------------

**Goals for this experiment:**
  1. Generate a table 1 including age, smoking, PY, cell type, FEV1%, FEV1/FVC
    * check for confounding in this table 1
  2. Find genes differentially expressed between COPD+cancer and everyone else
  3. Subsample the PFT subjects and generate a sample subset balanced for at least smoking
  4. Do pairwise comparisons between Disease Free, Cancer, COPD, COPD+Cancer groups
    * esp. check for differences between COPD and COPD+Cancer group (re Duncan's Q)
  
```{r GlobalVariables, echo=FALSE}
 cacheOption = FALSE

```


### GOAL 1: Generate a table 1 with age, smoking, PY, cell type, FEV1%, FEV1/FVC

First we need to load the data and create a subset with all samples with available PFTs
```{r Setup, cache=cacheOption, echo=FALSE, warning=FALSE, results="hide", message=FALSE}
# set the working directory to the directory of this script

# CREATE SIMULATED DATA USING MARC'S CODE
make.random.data <- function(real.matrix) {
  d <- matrix(rnorm(nrow(real.matrix)*ncol(real.matrix)), nrow=nrow(real.matrix))
  colnames(d) <- colnames(real.matrix)
  rownames(d) <- rownames(real.matrix)
  return(d)
}




setwd("/restricted/projectnb/pulmarray/LinGA_protected/Allegro/COPD_Cancer/experiments/2014-07-10")

# load the data using a script that removes patients with:
#  Cancer = NA
#  SMK    = 3
source("/protected/projects/pulmarray/Allegro/COPD_Cancer/scripts/AllegroSetup.R")
library(znorm)

# REPLACE REAL EXPRS DATA WITH SIMULATED DATA
eset(exprs) <- make.random.data(exprs(eset))

# remove the patients who have a DK for Cancer status as this is the a
# phenotype of interest (along with COPD)
eset <- removeFactorLevel(eset, "FinalCaDXc", "DK")

# remove the biological replicates
eset <- removeBioReps(eset)

# create a subset of all those subjects with PFTs and remove any NA/DK values in necessary fields
esetPFT <- cleanNAForAnalysis(eset, "COPD2_R7")
esetPFT <- removeFactorLevel(esetPFT, "COPD2_R7", "DK")
esetPFT <- calcIndicator(esetPFT, "FinalCaDXc", "COPD2_R7")

# generate the values to put in the table
ns <- summary(esetPFT$indicator)
# Age mean and standard deviation
ageM <- round(tapply(esetPFT$AGEcalc, esetPFT$indicator, mean))
ageS <- round(tapply(esetPFT$AGEcalc, esetPFT$indicator, sd))
ageP <- paste(ageM, ageS, sep="+/-")

# Pack years mean and standard deviation
pyM <- round(tapply(esetPFT$PYc, esetPFT$indicator, mean, na.rm=TRUE))
pyS <- round(tapply(esetPFT$PYc, esetPFT$indicator, sd, na.rm=TRUE))
pyP <- paste(pyM, pyS, sep="+/-")

# Smoking status breakdown
smkM <- tapply(esetPFT$SMKc, esetPFT$indicator, summary)

# Cancer subtype breakdown (i.e. NSCLC vs SCLC)
casubM <- tapply(esetPFT$CA_CELL_TYPE, esetPFT$indicator, summary)

# FEV1% mean and standard deviation
fev1pM <- round(tapply(esetPFT$FEV1Pc, esetPFT$indicator, mean),2)
fev1pS <- round(tapply(esetPFT$FEV1Pc, esetPFT$indicator, sd),2)
fev1pP <- paste(fev1pM, fev1pS, sep="+/-")

# FEV1/FVC mean and standard deviation
ratioM <- round(tapply(esetPFT$RATIOc, esetPFT$indicator, mean, na.rm=TRUE), 2)
ratioS <- round(tapply(esetPFT$RATIOc, esetPFT$indicator, sd, na.rm=TRUE), 2)
ratioP <- paste(ratioM, ratioS, sep="+/-")

```

                |  Disease Free  |   Cancer        |    COPD         |  Cancer+COPD
--------------- |:--------------:|:---------------:|:---------------:|:------------:
   **n**        |  `r ns[1]`     |   `r ns[2]`     |   `r ns[3]`     |  `r ns[4]`
   **Age**      |  `r ageP[1]`   |    `r ageP[2]`  |   `r ageP[3]`   |  `r ageP[4]`
   **Pack Years** | `r pyP[1]`    |   `r pyP[2]`    |   `r pyP[3]`    |  `r pyP[4]`
   **Smoking(C,F)** | `r smkM[[1]]` | `r smkM[[2]]`    | `r smkM[[3]]`  | `r smkM[[4]]`
   **Subtype(NSCLC, SCLC)**  | NA | `r casubM[[2]][2:3]` | NA  | `r casubM[[4]][2:3]`
   **FEV1%**   |  `r fev1pP[1]` |  `r fev1pP[2]`  |  `r fev1pP[3]`   |  `r fev1pP[4]`
   **FEV1/FVC**   |  `r ratioP[1]` |  `r ratioP[2]`  |  `r ratioP[3]`   |  `r ratioP[4]`
Table 1: demographics for samples with available PFT data. COPD is defined here as a ratio of less than 0.7.

### GOAL 2: Find genes differentially expressed between COPD+cancer and everyone else
```{r goal2Mash, cache=cacheOption, fig.width=8, fig.height=7, fig.align='center'}
# generate a binary factor that indicates COPD+Cancer +/-
cc <- numeric(sampleNumber(esetPFT))
cc[esetPFT$indicator==4] <- 1
cc <- factor(cc)
esetPFT$cc <- cc

designAllOrNone <- model.matrix(~ 1 + cc, data=esetPFT)
colnames(designAllOrNone) <- c("Intercept", "COPD+Cancer")

fitAllOrNone <- lmFit(esetPFT, designAllOrNone)
fitAllOrNone <- eBayes(fitAllOrNone)

resultsAllOrNone <- decideTests(fitAllOrNone, adjust.method="fdr", p.value=0.05)
summary(resultsAllOrNone)

# Try the same model but filter for signal/noise first using the following criteria:
#   filter out those genes for which there is not at least 1 sample with expression > total median

md <- median(exprs(esetPFT))
passFilter <- logical(featureNumber(esetPFT))
for(i in 1:featureNumber(esetPFT)) {
  passFilter[i] <- sum(exprs(esetPFT)[i, ] > md) > 0
}

# remove those genes not passing the median filter
esetPFTFilter <- esetPFT[passFilter, ]

designAllOrNoneF <- model.matrix(~ 1 + cc, data=esetPFTFilter)

fitAllOrNoneF <- lmFit(esetPFTFilter, designAllOrNoneF)
fitAllOrNoneF <- eBayes(fitAllOrNoneF)

resultsAllOrNoneF <- decideTests(fitAllOrNoneF, adjust.method="fdr", p.value=0.05)
summary(resultsAllOrNoneF)

generate_heatmap(which(resultsAllOrNoneF[, 2] != 0), esetPFTFilter, tp="indicator")

# try a model including age and/or smoking status
designAoNFSMK <- model.matrix(~ 1 + SMKc + cc, data=esetPFTFilter)

fitAoNFSMK <- lmFit(esetPFTFilter, designAoNFSMK)
fitAoNFSMK <- eBayes(fitAoNFSMK)

resultsAoNFSMK <- decideTests(fitAoNFSMK, adjust.method="fdr", p.value=0.05)
summary(resultsAoNFSMK)

temp <- esetPFTFilter
exprs(temp) <- znorm(exprs(temp), 1)
generate_heatmap(which(resultsAoNFSMK[, 3] != 0), temp, tp="indicator")

# generate model with smoking only for heatmap of residuals
designFSMK <- model.matrix(~ 1 + SMKc, data=esetPFTFilter)

fitFSMK <- lmFit(esetPFTFilter, designFSMK)
fitFSMK <- eBayes(fitFSMK)

residFSMK <- residuals(fitFSMK, esetPFTFilter)
temp <- esetPFTFilter
exprs(temp) <- znorm(residFSMK, margin=1)
generate_heatmap(which(resultsAoNFSMK[, 3] != 0), temp, tp="indicator")

resultsAoNFSMK <- decideTests(fitAoNFSMK, adjust.method="fdr", p.value=0.25)
summary(resultsAoNFSMK)
generate_heatmap(which(resultsAoNFSMK[, 3] != 0), temp, tp="indicator")

```

```{r boxPlots1, include=FALSE}
#Generate some boxplots and output as a pdf but do not include in the output HTML
pdf("AllOrNoneSMKresid.pdf")
for(i in 1:length(which(resultsAoNFSMK[, 3] != 0))){ 
  # generate the gene name
  ind <- which(resultsAoNFSMK[,3] != 0)[i]
 
  gn <- paste(fitFSMK$genes[ind,1], ":", fitFSMK$genes[ind,2],  sep="")
 
  boxplot(exprs(temp[which(resultsAoNFSMK[,3] != 0)[i], ])[1, ] ~ temp$indicator,
          names=c("Disease Free", "Cancer", "COPD", "Cancer+COPD"),
          col=c("green", "red", "blue", "yellow"),
          xlab="Group", ylab="Expression",
          main=gn)
  
 # legend("topleft", c("Disease Free", "Cancer", "COPD", "Cancer+COPD"), 
 #        fill=c("Green", "Red", "Blue", "Yellow"), inset=0.05)
  }

dev.off()
```


### GOAL 3: Subsample the PFT subjects and balance for [at least] smoking
```{r goal3SubSample, cache=cacheOption, fig.width=8, fig.height=7, fig.align='center', echo=FALSE}
tapply(esetPFT$SMKc, esetPFT$indicator, summary)

esetPFTFilterSub <- esetPFT

disFreeCurrent <- which(esetPFTFilterSub$SMKc==1 & esetPFTFilterSub$indicator==1)
disFreeFormer <- which(esetPFTFilterSub$SMKc==2 & esetPFTFilterSub$indicator==1)
disFreeFormer <- sample(disFreeFormer, size=length(disFreeCurrent), replace=FALSE)

cancerCurrent <- which(esetPFTFilterSub$SMKc==1 & esetPFTFilterSub$indicator==2)
cancerFormer <- which(esetPFTFilterSub$SMKc==2 & esetPFTFilterSub$indicator==2)
cancerFormer <- sample(cancerFormer, size=length(cancerCurrent), replace=FALSE)

copdCurrent <- which(esetPFTFilterSub$SMKc==1 & esetPFTFilterSub$indicator==3)
copdFormer <- which(esetPFTFilterSub$SMKc==2 & esetPFTFilterSub$indicator==3)
copdCurrent <- sample(copdCurrent, size=length(copdFormer), replace=FALSE)

bothCurrent <- which(esetPFTFilterSub$SMKc==1 & esetPFTFilterSub$indicator==4)
bothFormer <- which(esetPFTFilterSub$SMKc==2 & esetPFTFilterSub$indicator==4)
bothFormer <- sample(bothFormer, size=length(bothCurrent), replace=FALSE)

subSample <- c(disFreeCurrent, disFreeFormer, cancerCurrent, cancerFormer,
               copdCurrent, copdFormer, bothCurrent, bothFormer)

esetPFTFilterSub <- esetPFTFilterSub[, subSample]
tapply(esetPFTFilterSub$SMKc, esetPFTFilterSub$indicator, summary)

# perform signal/noise filtering (as above) based on median
md <- median(exprs(esetPFTFilterSub))
passFilter <- logical(featureNumber(esetPFTFilterSub))
for(i in 1:featureNumber(esetPFTFilterSub)) {
  passFilter[i] <- sum(exprs(esetPFTFilterSub)[i, ] > md) > 0
}

# remove those genes not passing the median filter
esetPFTSub <- esetPFTFilterSub
esetPFTFilterSub <- esetPFTFilterSub[passFilter, ]

# Generate a table 1 (as above) but for the now subsetted sample
# generate the values to put in the table
ns <- summary(esetPFTFilterSub$indicator)
# Age mean and standard deviation
ageM <- round(tapply(esetPFTFilterSub$AGEcalc, esetPFTFilterSub$indicator, mean))
ageS <- round(tapply(esetPFTFilterSub$AGEcalc, esetPFTFilterSub$indicator, sd))
ageP <- paste(ageM, ageS, sep="+/-")

# Pack years mean and standard deviation
pyM <- round(tapply(esetPFTFilterSub$PYc, esetPFTFilterSub$indicator, mean, na.rm=TRUE))
pyS <- round(tapply(esetPFTFilterSub$PYc, esetPFTFilterSub$indicator, sd, na.rm=TRUE))
pyP <- paste(pyM, pyS, sep="+/-")

# Smoking status breakdown
smkM <- tapply(esetPFTFilterSub$SMKc, esetPFTFilterSub$indicator, summary)

# Cancer subtype breakdown (i.e. NSCLC vs SCLC)
casubM <- tapply(esetPFTFilterSub$CA_CELL_TYPE, esetPFTFilterSub$indicator, summary)

# FEV1% mean and standard deviation
fev1pM <- round(tapply(esetPFTFilterSub$FEV1Pc, esetPFTFilterSub$indicator, mean),2)
fev1pS <- round(tapply(esetPFTFilterSub$FEV1Pc, esetPFTFilterSub$indicator, sd),2)
fev1pP <- paste(fev1pM, fev1pS, sep="+/-")

# FEV1/FVC mean and standard deviation
ratioM <- round(tapply(esetPFTFilterSub$RATIOc,
                       esetPFTFilterSub$indicator, mean, na.rm=TRUE), 2)
ratioS <- round(tapply(esetPFTFilterSub$RATIOc,
                       esetPFTFilterSub$indicator, sd, na.rm=TRUE), 2)
ratioP <- paste(ratioM, ratioS, sep="+/-")

```


                |  Disease Free  |   Cancer        |    COPD         |  Cancer+COPD
--------------- |:--------------:|:---------------:|:---------------:|:------------:
   **n**        |  `r ns[1]`     |   `r ns[2]`     |   `r ns[3]`     |  `r ns[4]`
   **Age**      |  `r ageP[1]`   |    `r ageP[2]`  |   `r ageP[3]`   |  `r ageP[4]`
   **Pack Years** | `r pyP[1]`    |   `r pyP[2]`    |   `r pyP[3]`    |  `r pyP[4]`
   **Smoking(C,F)** | `r smkM[[1]]` | `r smkM[[2]]`    | `r smkM[[3]]`  | `r smkM[[4]]`
   **Subtype(NSCLC, SCLC)**  | NA | `r casubM[[2]][2:3]` | NA  | `r casubM[[4]][2:3]`
   **FEV1%**   |  `r fev1pP[1]` |  `r fev1pP[2]`  |  `r fev1pP[3]`   |  `r fev1pP[4]`
   **FEV1/FVC**   |  `r ratioP[1]` |  `r ratioP[2]`  |  `r ratioP[3]`   |  `r ratioP[4]`
Table 1: demographics for subset samples with available PFT data. COPD is defined here as a ratio of less than 0.7. subsampling was done to more-or-less match the ratio of former to current smokers from the COPD only group above (was 18/17, now 17/17)

#### GOAL 3A: Run models with subsampled and balanced groups looking for interaction term
```{r goal3aInteractionModel}
designIntrx <- model.matrix(~ 1 + AGEcalc + SMKc + COPD2_R7*FinalCaDXc, data=esetPFTFilterSub)

fitIntrx <- lmFit(esetPFTFilterSub, designIntrx)
fitIntrx <- eBayes(fitIntrx)

resultsIntrx <- decideTests(fitIntrx, adjust.method="fdr", p.value=0.05)
summary(resultsIntrx)

temp <- esetPFTFilterSub
# generate smoking residuals
designFSubSMK <- model.matrix(~ 1 + SMKc, data=temp)
fitFSubSMK <- lmFit(temp, designFSubSMK)
fitFSubSMK <- eBayes(fitFSubSMK)
residFSubSMK <- residuals(fitFSubSMK, temp)
exprs(temp) <- znorm(residFSubSMK)

generate_heatmap(which(resultsIntrx[, 6] != 0), temp, tp="indicator")

```

#### GOAL 3B: Run models with subsampled and balanced group using models from GOAL 2
```{r goal3bMashModel, fig.width=8, fig.height=7, fig.align='center', cache=cacheOption}
designFSubAoN1 <- model.matrix( ~ 1 + AGEcalc + cc, data=esetPFTFilterSub)
designFSubAoN2 <- model.matrix( ~ 1 + SMKc + cc, data=esetPFTFilterSub)
designFSubAoN3 <- model.matrix( ~ 1 + AGEcalc + SMKc + cc, data=esetPFTFilterSub)

fitFSubAoN1 <- lmFit(esetPFTFilterSub, designFSubAoN1)
fitFSubAoN2 <- lmFit(esetPFTFilterSub, designFSubAoN2)
fitFSubAoN3 <- lmFit(esetPFTFilterSub, designFSubAoN3)

fitFSubAoN1 <- eBayes(fitFSubAoN1)
fitFSubAoN2 <- eBayes(fitFSubAoN2)
fitFSubAoN3 <- eBayes(fitFSubAoN3)

resultsFSubAoN1 <- decideTests(fitFSubAoN1, adjust.method="fdr", p.value=0.05)
resultsFSubAoN2 <- decideTests(fitFSubAoN2, adjust.method="fdr", p.value=0.05)
resultsFSubAoN3 <- decideTests(fitFSubAoN3, adjust.method="fdr", p.value=0.05)

summary(resultsFSubAoN1)
summary(resultsFSubAoN2)
summary(resultsFSubAoN3)




```


### GOAL 4: Do pairwise comparisons between 4 groups (esp. COPD and COPD+Cancer)
```{r goal4Pairwise}